\AtBeginDocument{%
  \providecommand\BibTeX{{Bib\TeX}}
}

\usepackage{mathtools}
\usepackage{mathrsfs}
\usepackage{listings}
\usepackage{multirow}
\usepackage{graphicx}

\AtBeginDocument{%
  \theoremstyle{acmdefinition}%
  \newtheorem{remark}[theorem]{Remark}%
}

\lstdefinestyle{scala}{
  language=Scala,
  basicstyle=\ttfamily\small,
  breaklines=true,
  showstringspaces=false,
  columns=flexible,
}

\lstdefinestyle{haskell}{
  language=Haskell,
  basicstyle=\ttfamily\small,
  breaklines=true,
  showstringspaces=false,
  columns=flexible,
}

\lstdefinestyle{nix}{
  language=,
  basicstyle=\ttfamily\small,
  breaklines=true,
  showstringspaces=false,
  columns=flexible,
  commentstyle=\color[gray]{0.5},
  morecomment=[l]{\#},
}

%% Conditional citation: hide self-citations under anonymous review.
%% Use \@firstoftwo / \@secondoftwo so the unchosen branch is never
%% expanded---preventing \cite from writing to the .aux file.
\makeatletter
\if@ACM@anonymous
\newcommand{\selfcite}[2]{#2}%
\else
\newcommand{\selfcite}[2]{#1}%
\fi
\makeatother

\newcommand{\Path}{\mathtt{Path}}
\newcommand{\dom}{\mathtt{dom}}
\newcommand{\properties}{\mathtt{properties}}
\newcommand{\overrides}{\mathtt{overrides}}
\newcommand{\supers}{\mathtt{supers}}
\newcommand{\inherits}{\mathtt{inherits}}
\newcommand{\defines}{\mathtt{defines}}
\newcommand{\bases}{\mathtt{bases}}
\newcommand{\resolve}{\mathtt{resolve}}
\newcommand{\this}{\mathtt{this}}
\newcommand{\snoc}{\mathbin{\triangleright}}
\newcommand{\init}{\mathtt{init}}
\newcommand{\last}{\mathtt{last}}
\newcommand{\qualifiedthis}[1]{#1.\mathbf{this}}
\newcommand{\dbi}[1]{\uparrow^{\!#1}}

%% Semantic functions for the direct λ-calculus fixpoint semantics (Appendix E)
%% -- Scope-level functions --
\newcommand{\result}{\mathtt{result}}
%% \newcommand{\visiblebindings}{\mathtt{visiblebindings}}
%% -- Application-level functions --
\newcommand{\head}{\mathtt{head}}
\newcommand{\argument}{\mathtt{argument}}
%% -- Calling context --
%% callerscope/callerargument inlined as projections from \contexts
\newcommand{\letcontexts}{\mathtt{letcontexts}}
\newcommand{\tailcontexts}{\mathtt{tailcontexts}}
\newcommand{\contexts}{\mathtt{contexts}}
%% -- Shared resolution --
\newcommand{\scoperesolution}{\mathtt{scoperesolve}}
\newcommand{\resolveref}{\mathtt{resolveref}}
%% -- Auxiliary --
\newcommand{\enclosingscope}{\mathtt{enclosingscope}}

\begin{document}

\title{A Calculus of Inheritance}

\author{Bo Yang}
\affiliation{%
  \institution{Figure AI Inc.}
  \city{San Jose}
  \state{California}
  \country{USA}
}
\email{yang-bo@yang-bo.com}
\thanks{This work was conducted independently prior to the author's employment at Figure AI.}

\begin{abstract}
  Just as the $\lambda$-calculus uses three primitives (abstraction, application, variable) as the foundation of functional programming, inheritance-calculus uses three primitives (record, definition, inheritance) as the foundation of declarative programming. By unifying modules, classes, objects, methods, fields, and locals under a single record abstraction, the calculus models inheritance simply as set union. Consequently, composition is inherently commutative, idempotent, and associative, structurally eliminating the multiple-inheritance linearization problem. Its semantics is first-order~\cite{vanemden1976-predicate-logic-semantics, reynolds1972-definitional-interpreters, aczel1977-inductive-definitions}, denotational, and computable by tabling~\cite{tamaki1986-tabled-resolution}, even for cyclic inheritance hierarchies. These three properties extend to the $\lambda$-calculus, since B\"ohm tree equivalence~\cite{barendregt1984-lambda-calculus} is fully abstract for the first-iteration approximation of a sublanguage of inheritance-calculus. As a corollary, this establishes a convergence hierarchy $\text{eager} \subsetneq \text{lazy} \subsetneq \text{fixpoint}$ among $\lambda$-calculi sharing the same $\lambda$-syntax.

  Inheritance-calculus is distilled from MIXINv2, a practical implementation in which the same code acts as different function colors~\cite{nystrom2015-function-color}; ordinary arithmetic yields the relational semantics of logic programming~\cite{vanemden1976-predicate-logic-semantics}; $\mathtt{this}$ resolves to multiple targets; and programs are immune to nonextensibility in the sense of the Expression Problem~\cite{wadler1998-expression-problem}. This makes inheritance-calculus strictly more expressive than the $\lambda$-calculus in both common sense and Felleisen's sense~\cite{felleisen1991-expressive-power}.
\end{abstract}

\begin{CCSXML}
  <ccs2012>
  <concept>
  <concept_id>10003752.10010124.10010131</concept_id>
  <concept_desc>Theory of computation~Program semantics</concept_desc>
  <concept_significance>500</concept_significance>
  </concept>
  <concept>
  <concept_id>10003752.10010124.10010125.10010128</concept_id>
  <concept_desc>Theory of computation~Object oriented constructs</concept_desc>
  <concept_significance>500</concept_significance>
  </concept>
  <concept>
  <concept_id>10003752.10010124.10010125.10010127</concept_id>
  <concept_desc>Theory of computation~Functional constructs</concept_desc>
  <concept_significance>300</concept_significance>
  </concept>
  <concept>
  <concept_id>10003752.10010124.10010125.10010129</concept_id>
  <concept_desc>Theory of computation~Program schemes</concept_desc>
  <concept_significance>100</concept_significance>
  </concept>
  <concept>
  <concept_id>10011007.10011006.10011039</concept_id>
  <concept_desc>Software and its engineering~Formal language definitions</concept_desc>
  <concept_significance>500</concept_significance>
  </concept>
  <concept>
  <concept_id>10011007.10011006.10011008.10011009.10011019</concept_id>
  <concept_desc>Software and its engineering~Extensible languages</concept_desc>
  <concept_significance>500</concept_significance>
  </concept>
  <concept>
  <concept_id>10011007.10011006.10011008.10011009.10011011</concept_id>
  <concept_desc>Software and its engineering~Object oriented languages</concept_desc>
  <concept_significance>100</concept_significance>
  </concept>
  </ccs2012>
\end{CCSXML}

\ccsdesc[500]{Theory of computation~Program semantics}
\ccsdesc[500]{Theory of computation~Object oriented constructs}
\ccsdesc[300]{Theory of computation~Functional constructs}
\ccsdesc[100]{Theory of computation~Program schemes}
\ccsdesc[500]{Software and its engineering~Formal language definitions}
\ccsdesc[500]{Software and its engineering~Extensible languages}
\ccsdesc[100]{Software and its engineering~Object oriented languages}

\keywords{declarative programming, inheritance, fixpoint semantics,
  self-referential records, expression problem,
  first-order semantics,
  \texorpdfstring{$\lambda$-calculus}{lambda-calculus},
\texorpdfstring{B\"ohm trees}{Bohm trees}, configuration languages}

\maketitle

\section{Introduction}
\label{sec:introduction}

Declarative and configuration languages are ubiquitous in modern software
engineering. Systems such as NixOS modules~\cite{dolstra2010-nixos,nixosmodules},
Jsonnet~\cite{jsonnet}, Hydra~\cite{hydra2023}, CUE~\cite{cue2019}, Dhall~\cite{dhall2017},
Kustomize~\cite{kustomize}, and JSON Patch~\cite{rfc6902-json-patch} all provide mechanisms for composing
structured data through inheritance or merging. Among these, the NixOS
module system stands out: its recursive attribute set merging with
fixpoint semantics and deferred modules~\cite{nixosmodules} allows
independently authored modules to extend the system configuration
without modifying existing modules;
in practice, the ecosystem has not been troubled by the
multi-dimensional extension difficulty analogous to the
Expression Problem~\cite{wadler1998-expression-problem}.%
\footnote{
  As a rough proxy for ecosystem scale:
  according to Repology~\cite{repology2025},
  \texttt{nixpkgs} contains over 113{,}000 packages,
  nearly three times the count of Debian or Ubuntu,
  with a comparable number of contributors.
  Not every package exercises the module system's
  extensibility, but the same caveat applies to other distributions.
}
The mechanism has been adopted well beyond NixOS itself: Home
Manager~\cite{homemanager}, nix-darwin~\cite{nixdarwin},
disko~\cite{disko}, flake-parts~\cite{flakeparts},
dream2nix~\cite{dream2nix}, devenv~\cite{devenv},
KubeNix~\cite{kubenix}, and nixidy~\cite{nixidy} use it to manage
user environments, macOS configuration, disk partitioning, flake
structure, multi-language packaging, developer environments, and
Kubernetes clusters respectively. Yet no computational
theory captures the extensibility that this mechanism exhibits.

The $\lambda$-calculus serves as the foundational computational model for
functional programming. No analogous calculus exists for declarative
programming. This gap matters because the two paradigms differ in
fundamental ways. Configuration languages are inherently declarative:
their values are immutable and they have no first-class
functions.
Note that a function body is closed: its internal structure
cannot be extended without modifying the definition, whereas
the configuration languages above all compose by extending
existing structure from outside.
At first glance, Turing completeness appears incompatible
with these constraints. The three classical models of computation each
violate at least one of them:

\begin{table}[h]
  \centering
  \begin{tabular}{lccc}
    \textbf{Model} & \textbf{Turing complete} & \textbf{Immutable} & \textbf{No $\lambda$} \\
    \hline
    Turing Machine     & $\checkmark$ & $\times$     & $\checkmark$ \\
    $\lambda$-calculus & $\checkmark$ & $\checkmark$ & $\times$ \\
    RAM Machine        & $\checkmark$ & $\times$     & $\checkmark$ \\
  \end{tabular}
  \caption{Classical computational models and the declarative gap.}
  \label{tab:computational-models}
\end{table}

\noindent
The Turing Machine and RAM Machine require mutable state;
the $\lambda$-calculus requires first-class functions.
A computational model that is Turing complete, immutable, and
function-free would sit in the empty cell of Table~\ref{tab:computational-models}.
The NixOS module system already suggests that such a model exists:
its inheritance-based composition over recursive records, without
explicit functions, is expressive enough to configure entire operating
systems.

Most configuration languages operate on finite, well-founded
structures (initial algebras in the sense of universal algebra).
Nix is an exception: although designed specifically as a
configuration language for
NixOS~\cite{dolstra2006-purely-functional-deployment},
it is a complete programming language whose values are lazily
observable and possibly infinite.
The \texttt{nixpkgs} package collection (over 100{,}000
packages) behaves as a lazily observed structure, allowing any
single package to be evaluated without materializing the entire
set.
Guided by this observation, we set out to reduce the
NixOS module system to a minimal set of primitives.
The reduction produced MIXIN, an early implementation in Nix,
and subsequently
MIXINv2\selfcite{~\cite{mixin2025}}{~(included as supplementary
material)}, now implemented in Python: a declarative programming
language with only three constructs (record literals, definitions,
and inheritance) and no functions or let-bindings.%
\footnote{
  The executable MIXINv2
  implementation\selfcite{~\cite{mixin2025}}{~(supplementary
  material)} contains every example in this paper.
}

In developing real programs in MIXINv2, including boolean logic,
natural number arithmetic, and a web application, several
phenomena arose from off-label use of the adopted
design patterns.
Section~\ref{sec:mixin-trees} shows how the three constructs
give rise to a single abstraction that simultaneously serves as
module, class, object, method, field, and local binding.
The linearization problem of mixin-based
systems~\cite{bracha1990-mixin-inheritance,barrett1996-c3-linearization}
did not arise: inheritance is inherently commutative, idempotent,
and associative, and $\this$ naturally resolves to
\emph{multiple targets} rather than one, where prior systems
such as Scala and the NixOS module system reject the multi-target
situation.
Section~\ref{sec:case-study} traces how boolean logic and
arithmetic, implemented as separate files, extend in both
dimensions of the Expression
Problem~\cite{wadler1998-expression-problem} without any dedicated
mechanism, and how ordinary arithmetic yields the relational
semantics of logic
programming~\cite{vanemden1976-predicate-logic-semantics}.
Section~\ref{sec:emergent-phenomena} discusses these and further
emergent phenomena, including function color
blindness~\cite{nystrom2015-function-color}.

This paper distills these observations into \textbf{inheritance-calculus},
a minimal computational model for declarative programming.
Section~\ref{sec:mixin-trees} defines the semantics via six
mutually recursive first-order functions on paths and sets of paths,
with no function spaces, closures, or higher-order constructs;
the functions are computable by
tabling~\cite{tamaki1986-tabled-resolution}.
Section~\ref{sec:forward-translation} shows that
these functions extend to the $\lambda$-calculus, embedding it.
Remark~\ref{rem:convergence-lfp} derives a convergence hierarchy
among eager, lazy, and fixpoint
$\lambda$-calculi sharing the same syntax as a corollary.
Section~\ref{sec:asymmetry} establishes that the lazy
$\lambda$-calculus cannot macro-express the inheritance constructors,
so inheritance-calculus is strictly more expressive in
Felleisen's sense~\cite{felleisen1991-expressive-power}.

\section{Syntax}
\label{sec:syntax}

Let $e$ denote an expression,
$\ell$ a label serving as a property name,
and $k$ a nonnegative integer serving as the length of a downward projection.

\begin{align*}
  e \quad ::= \quad & \{c_1,\; \ldots,\; c_m\}
  && \text{(} m \ge 0 \text{)} \\
  \mid\quad & c
  && \text{(sugar for } \{c\} \text{)} \\[6pt]
  c \quad ::= \quad & \ell \mapsto e
  && \text{(definition)} \\
  \mid\quad & (\qualifiedthis{\ell_{\mathrm{up}}} \mid \dbi{n}).\;\ell_{\mathrm{down},1}.\ell_{\mathrm{down},2}\ldots\ell_{\mathrm{down},k}
  && \text{(inheritance: qualifier + downward path)} \\
  \mid\quad & \qualifiedthis{\ell_{\mathrm{up}}} \mid \dbi{n}
  && \text{(inheritance: qualifier only)} \\
  \mid\quad & \ell_{\mathrm{down},1}.\ell_{\mathrm{down},2}\ldots\ell_{\mathrm{down},k}
  && \text{(inheritance: downward path only)}
\end{align*}

An expression is a \emph{set} of elements enclosed in braces.
Each element is either a definition $\ell \mapsto e$ or
an inheritance source.
Because expressions are sets, element order is irrelevant and
duplicates have no effect; inheritance is inherently commutative,
idempotent, and associative.

The $\mapsto$ in $\ell \mapsto e$ defines a property named $\ell$ with value $e$.
MIXINv2 also provides scalar types through a
foreign-function interface (FFI); inheritance-calculus is what remains
after removing the FFI.
None of the examples in this paper uses the FFI; no scalar values are needed in the examples.

\paragraph{Inheritance}
An inheritance is written as a reference
that navigates the scope hierarchy upward and then projects
properties downward. The upward navigation admits two equivalent
notations:
\begin{description}
  \item[Named form]
    $\qualifiedthis{\ell_{\mathrm{up}}}.\ell_{\mathrm{down},1}\ldots\ell_{\mathrm{down},k}$,
    analogous to Java's \texttt{Outer.this.field}.
    Each $\{\ldots\}$ record has a label in its enclosing scope;
    $\ell_{\mathrm{up}}$ names the target enclosing scope.
  \item[Indexed form]
    $\dbi{n}.\ell_{\mathrm{down},1}\ldots\ell_{\mathrm{down},k}$,
    where $n \ge 0$ is a de~Bruijn index~\cite{debruijn1972-nameless-dummies}:
    $n = 0$ refers to the reference's enclosing scope,
    and each increment of $n$ moves one scope level further out.
\end{description}
Both forms retrieve the target scope, that is, the fully inherited record
after all inheritance has been applied, and then project properties
$\ell_{\mathrm{down},1}\ldots\ell_{\mathrm{down},k}$ from it.
The named form desugars to the indexed form during parsing
(Section~\ref{sec:mixin-trees}).

A new scope level is created at each $\{\ldots\}$ record literal.%
\footnote{
  We use \emph{record} for the syntactic construct, \emph{scope}
  for the name-resolution context it creates, and \emph{mixin} for its
  role as a composable unit; in inheritance-calculus these are three views
  of the same entity.
}
The scope contains all properties of that mixin, including those
inherited via inheritance sources.
For example, $\{a \mapsto \{\},\; b \mapsto \{\}\}$ is a single mixin whose scope
contains both $a$ and $b$; so is $\{r,\; a \mapsto \{\}\}$ where $r$
is a reference to a record containing $b$.
Inheritance sources do not create additional scope levels.

\begin{itemize}
  \item $\qualifiedthis{\mathrm{Foo}}$ (equivalently, $\dbi{n}$ where
    $\mathrm{Foo}$ is $n$ steps above the enclosing scope)
    retrieves the record named $\mathrm{Foo}$.
  \item $\qualifiedthis{\mathrm{Foo}}.\ell$, or equivalently, $\dbi{n}.\ell$,
    projects property $\ell$ from $\mathrm{Foo}$.
  \item When $k = 0$, the reference inherits the entire
    enclosing scope, which may produce an infinitely deep tree.
\end{itemize}

When the qualifier is unambiguous, it may be omitted:
the shorthand
$\ell_{\mathrm{down},1}.\ell_{\mathrm{down},2}\ldots\ell_{\mathrm{down},k}$ ($k \ge 1$)
resolves $\ell_{\mathrm{down},1}$ against the innermost enclosing
scope whose $\defines$ contains it,
and desugars to the corresponding indexed form.
Since this uses $\defines$, the shorthand can only access
properties defined in record literals, not those inherited through
inheritance; one of the two qualified forms is required for
inherited properties or to bypass variable shadowing.

\section{Mixin Trees}
\label{sec:mixin-trees}

In object-oriented programming, a
\emph{mixin}~\cite{bracha1990-mixin-inheritance} is a fragment of behavior
that can be composed onto a class via inheritance.
Inheritance-calculus has a single abstraction, the
\emph{deep-mergeable mixin},%
\footnote{
  Where no ambiguity arises, we write simply ``mixin.''
  The distinction from the Bracha--Cook
  mixin~\cite{bracha1990-mixin-inheritance} is detailed in
  Section~\ref{sec:related-work}.
}
which can simultaneously serve as module, class, object, method, field, and
local binding, because there are no opaque methods: every value is
a transparent record, and inheritance always performs a deep
merge of subtrees.
We call the resulting structure a \emph{mixin tree}.
Its semantics is purely \emph{observational}: given an
inheritance-calculus program, one queries whether a
label $\ell$ is a property at a given position $p$.

This section defines the semantics via six mutually recursive
first-order%
\footnote{
  First-order in three senses simultaneously:
  the semantic domain contains only paths (finite
  sequences of labels) and sets of paths, with no function
  spaces, closures, or strategies~\cite{vanemden1976-predicate-logic-semantics};
  the six equations are first-order functions whose
  inputs and outputs are paths and sets of paths,
  never functions~\cite{reynolds1972-definitional-interpreters};
  and the judgments are first-order inductive
  definitions~\cite{aczel1977-inductive-definitions} over paths
  and labels, where quantifiers range over paths and labels,
  never over functions or predicates.
}
functions on paths.
The six functions map object-oriented concepts to
set-theoretic operations: $\properties$ and $\supers$
generalize member lookup and inheritance chains,
$\overrides$ and $\bases$ capture method override and
base classes, and $\resolve$ and $\this$ handle name
resolution and qualified $\this$ resolution.

\subsection{Definitions}

\paragraph{Path}
A $\Path$ is a sequence of labels
$(\ell_1, \ell_2, \ldots, \ell_n)$ that identifies a position in
the tree.
We write $p$ for a path.
The \emph{root path} is the empty sequence $()$.
Given a path $p$ and a label $\ell$,
$p \snoc \ell$ is the sequence $p$ extended with $\ell$.
For a nonroot path, the parent path $\init(p)$ is $p$ with
its last element removed,
and the final label $\last(p)$ is the last element of $p$.
For convenience, we sometimes write $\ell$ for $\last(p)$
when the path is clear from context.
With paths in hand, we can state what the AST provides
at each path.

\paragraph{AST}
An expression $e$ (Section~\ref{sec:syntax}) is parsed into an AST.
The AST provides two primitive functions at each path $p$:
\begin{itemize}
  \item $\defines(p)$: the set of labels $\ell$ for which $p$
    contains a definition $\ell \mapsto e$.
  \item $\inherits(p)$: the set of reference pairs
    $(n,\; \ell_{\mathrm{down},*})$,
    where $n$ is the de~Bruijn index
    and $\ell_{\mathrm{down},*}$ is the list of downward projections.
\end{itemize}

During parsing, all three syntactic forms are resolved to
de~Bruijn index pairs $(n,\; \ell_{\mathrm{down},*})$.
The \emph{indexed form}
$\dbi{n}.\ell_{\mathrm{down},1}\ldots\ell_{\mathrm{down},k}$
already carries the de~Bruijn index $n$ directly;
no resolution is needed.
A \emph{named reference}
$\qualifiedthis{\ell_{\mathrm{up}}}.\ell_{\mathrm{down},1}\ldots\ell_{\mathrm{down},k}$
at path $p$ is resolved by finding the last occurrence of
$\ell_{\mathrm{up}}$ among the labels of $p$.
Let $p_{\mathrm{target}}$ be the prefix of $p$ up to and including
that occurrence.
The de~Bruijn index is $n = |p| - |p_{\mathrm{target}}| - 1$
and the projections are
$\ell_{\mathrm{down},*} = (\ell_{\mathrm{down},1}, \ldots, \ell_{\mathrm{down},k})$.
A \emph{lexical reference} $\ell_{\mathrm{down},1}.\ell_{\mathrm{down},2}\ldots\ell_{\mathrm{down},k}$ at path $p$
is resolved by finding the nearest prefix $p'$ of $p$
such that $\ell_{\mathrm{down},1} \in \defines(p')$.
The de~Bruijn index is $n = |p| - |p'| - 1$ and the projections are
$\ell_{\mathrm{down},*} = (\ell_{\mathrm{down},1}, \ldots, \ell_{\mathrm{down},k})$.
All three forms produce the same representation; the semantic functions
below operate only on de~Bruijn index pairs.

Parsing populates $\defines$ and $\inherits$ from each
element of a record.
A definition $\ell \mapsto e$
contributes $\ell$ to $\defines(p)$.
A reference contributes $(n,\; \ell_{\mathrm{down},*})$ to $\inherits(p)$.
Both $\defines$ and $\inherits$ are pure data; they are not functions
of runtime state.
Given these two primitives, we can now answer the question
posed at the beginning of this section.

\paragraph{Properties}
The properties of a path are its own properties together with
those inherited from all supers:
\begin{equation}\label{eq:properties}
  \properties(p) =
  \bigl\{\; \ell \;\big|\;
    (\_,\; p_{\mathrm{override}}) \in \supers(p),\;
    \ell \in \defines(p_{\mathrm{override}})
  \;\bigr\}
\end{equation}
The remainder of this section defines $\supers$,
the transitive inheritance closure,
and its dependencies.

\paragraph{Supers}
Intuitively, $\supers(p)$ collects every path that $p$ inherits from:
the $\overrides$ of $p$ itself,
plus the $\overrides$ of each
direct base $\bases$ of $p$, and so on transitively.
Each result is paired with the inheritance-site context
$\init(p_{\mathrm{base}})$ through which it is reached.
This provenance is needed by qualified $\this$ resolution, that is, the $\this$ function defined below,
to map a definition-site mixin back to
the inheritance-site paths that incorporate it:
\begin{equation}\label{eq:supers}
  \supers(p) =
  \bigl\{\; (\init(p_{\mathrm{base}}),\; p_{\mathrm{override}}) \;\big|\;
    p_{\mathrm{base}} \in \bases^*(p),\;
    p_{\mathrm{override}} \in \overrides(p_{\mathrm{base}})
  \;\bigr\}
\end{equation}
Here $\bases^*$ denotes the reflexive-transitive closure of $\bases$.
The $\supers$ formula depends on two functions:
$\overrides$ and the one-hop reference targets $\bases$.
We define $\overrides$ first.

\paragraph{Overrides}
Since a mixin can play the role of a method,
a child scope should be
able to find it through the parent's label even if the child
does not redefine that label; this is method override.
Concretely, inheritance can introduce multiple
definitions of the same label at the same scope level;
$\overrides(p)$ collects all such paths that share the
\emph{same identity} as $p$, so that their definitions
merge rather than shadow each other, enabling deep merge.
The overrides of $p$ include $p$ itself
and $p_{\mathrm{branch}} \snoc \last(p)$ for every
branch $p_{\mathrm{branch}}$ of $\init(p)$ that also defines $\last(p)$:
\begin{equation}\label{eq:overrides}
  \overrides(p) =
  \begin{cases}
    \{p\} & \text{if } p = ()  \\[6pt]
    \{p\}
    \;\cup\;
    \left\{\; p_{\mathrm{branch}} \snoc \last(p) \;\left|\;
      \begin{aligned}
        &(\_,\; p_{\mathrm{branch}}) \in \supers(\init(p)), \\
        &\text{s.t.}\; \last(p) \in \defines(p_{\mathrm{branch}})
      \end{aligned}
    \right.\right\}
    & \text{if } p \neq ()
  \end{cases}
\end{equation}
It remains to define $\bases$, the other dependency of $\supers$.

\paragraph{Bases}
Intuitively, $\bases(p)$ are the paths that $p$ directly
inherits from via references, analogous to the direct base
classes in object-oriented languages.
Concretely, $\bases$ resolves every reference in
$p$'s $\overrides$ one step:
\begin{equation}\label{eq:bases}
  \bases(p) =
  \left\{\; p_{\mathrm{target}} \;\left|\;
    \begin{aligned}
      &p_{\mathrm{override}} \in \overrides(p), \\
      &(n,\; \ell_{\mathrm{down},*}) \in \inherits(p_{\mathrm{override}}), \\
      &p_{\mathrm{target}} \in \resolve(\init(p),\; p_{\mathrm{override}},\; n,\; \ell_{\mathrm{down},*})
    \end{aligned}
  \right.\right\}
\end{equation}
The $\bases$ formula calls the reference resolution function $\resolve$, which we define next.

\paragraph{Reference resolution}
Intuitively, $\resolve$ turns a syntactic reference into the
set of paths it points to in the fully inherited tree.
It takes an inheritance-site path $p_{\mathrm{site}}$,
a definition-site path $p_{\mathrm{def}}$,
a de~Bruijn index $n$,
and downward projections $\ell_{\mathrm{down},*}$.
Resolution proceeds in two phases:
$\this$ performs $n$ upward steps
starting from the enclosing scope $\init(p_{\mathrm{def}})$,
mapping the definition-site path to inheritance-site paths;
then the downward projections $\ell_{\mathrm{down},*}$ are appended:
\begin{multline}\label{eq:resolve}
  \resolve(p_{\mathrm{site}},\; p_{\mathrm{def}},\; n,\; \ell_{\mathrm{down},*})
  = \\
  \bigl\{\;
    p_{\mathrm{current}} \snoc \ell_{\mathrm{down},1} \snoc \cdots \snoc \ell_{\mathrm{down},k}
    \;\big|\;
    p_{\mathrm{current}} \in
    \this(\{p_{\mathrm{site}}\},\; \init(p_{\mathrm{def}}),\; n)
  \;\bigr\}
\end{multline}
When multiple routes exist due to multi-target $\this$ resolution,
they may yield different target paths; all are collected.
We now define $\this$.

\paragraph{Qualified $\this$ resolution}
$\this$ answers the question:
``in the fully inherited tree, where does
the definition-site scope $p_{\mathrm{def}}$ actually live?''
Each step finds, among the supers of every path in the
frontier $S$, those whose override component matches
$p_{\mathrm{def}}$, and collects the corresponding
inheritance-site paths as the new frontier.
After $n$ steps the frontier contains the answer:
{\small
  \begin{equation}\label{eq:this}
    \this(S,\; p_{\mathrm{def}},\; n) =
    \begin{cases}
      S & \text{if } n = 0 \\[6pt]
      \this\!\left(
        \left\{\; p_{\mathrm{site}} \;\left|\;
          \begin{aligned}
            &p_{\mathrm{current}} \in S, \\
            &(p_{\mathrm{site}},\; p_{\mathrm{override}})
            \in \supers(p_{\mathrm{current}}), \\
            &\text{s.t.}\; p_{\mathrm{override}} = p_{\mathrm{def}}
          \end{aligned}
        \right.\right\},\;
        \init(p_{\mathrm{def}}),\;
        n - 1
      \right)
      & \text{if } n > 0
    \end{cases}
\end{equation}}%
At each step $\init$ shortens $p_{\mathrm{def}}$ by one label
and $n$ decreases by one.
Since $n$ is a nonnegative integer, the recursion terminates.%
\footnote{
  If no matching pair
  $(p_{\mathrm{site}},\; p_{\mathrm{override}})$ exists
  at some step, the frontier becomes the empty set, and
  $\resolve$ returns $\varnothing$: the reference has no
  target; no properties are inherited.
  This is the most self-consistent semantics for the
  untyped calculus.
  MIXINv2\selfcite{~\cite{mixin2025}}{~(supplementary
  material)} is stricter: its resolver rejects
  references whose frontier becomes empty, reporting a
  resolution error at compile time.
}

This completes the chain of definitions needed to compute $\properties(p)$.

\subsection{Recursive Evaluation}
The six equations~(\ref{eq:properties})--(\ref{eq:this})
define $\properties(p)$ by mutual recursion:
$\ell \in \properties(p)$ holds if and only if it can
be established by a finite chain of applications of
these functions.
The functions \emph{are} the interpreter: the observer
drives the computation by choosing which path to
inspect, and each equation unfolds on demand.
Because the functions are pure and order-independent,
the reflexive-transitive closure $\bases^*$
in~(\ref{eq:supers}) may be explored breadth-first or
depth-first without affecting the result;
paths may be interned so that structural equality
reduces to pointer equality;
and the six functions form a memoizable
dynamic-programming recurrence keyed by paths.

Operationally, evaluation uses \emph{tabled
evaluation}~\cite{tamaki1986-tabled-resolution}:
each query is memoised on first encounter, and
re-entrant calls, which correspond to cycles in the dependency graph,
return the current accumulator rather than diverging.
Denotationally, the six equations define a monotone
immediate consequence
operator~$T_P$~\cite{vanemden1976-predicate-logic-semantics}
whose right-hand sides use only $\cup$, $\in$, equality, and
positive set comprehension, with no negation, set
difference, or complement.
By the Knaster--Tarski
theorem~\cite{tarski1955-lattice-fixpoint-theorem}, $T_P$ has a unique
least fixed point $\mathrm{lfp}(T_P)$, which serves as
the denotational semantics.
This denotational characterization formalizes the observational
semantics introduced above: the observer-driven queries are
computed by the same fixed-point equations.
Whether evaluation terminates depends on the program.
Recall that $\supers$ (equation~\ref{eq:supers}) takes the
reflexive-transitive closure of $\bases$
(equation~\ref{eq:bases}); the resulting graph is the
\emph{inheritance hierarchy}.
Since each inheritance source occupies a distinct node in
the mixin tree, this hierarchy is directly readable from
the tree structure.
Programs whose inheritance hierarchy is acyclic terminate in
one pass.
When the hierarchy contains cycles,
evaluation terminates when each query has a finite
answer domain: the ascending chain condition on the finite
lattice guarantees
convergence~\cite{bancilhon1986-recursive-query-strategies},
as in Datalog (Appendix~\ref{app:datalog-encoding}).
When the hierarchy is infinite, evaluation may
diverge; this is inherent to Turing completeness.
Appendix~\ref{app:well-definedness} proves that tabled
evaluation never returns a wrong answer and formalizes
these termination conditions.

%% === 审稿人备忘：六个方程的渐近复杂度分析 ===
%%
%% 论文声称语义是一阶的，审稿人可能追问：一阶性是否带来额外的
%% 计算开销？下面按场景分析。
%%
%% 关键观察：set comprehension 自动去重（集合语义），加上路径
%% intern（结构相等退化为指针相等）和 memoization，相同的
%% (path, query) 对只计算一次。这不是可选的优化，而是集合语义
%% 的固有性质——集合不会包含重复元素。
%%
%% --- 场景 1：λ-calculus 翻译 ---
%% Single-Path Lemma (Lemma 3) 保证 this 的 frontier 集合 S
%% 始终只有一个元素。每个路径的 overrides 也只有一个元素。
%% bases 每个节点最多一个继承源。supers 的传递闭包是线性链。
%% 复杂度：O(n)，其中 n 是 ANF term 的大小。
%% 与传统环境机（environment machine）一致。
%%
%% --- 场景 2：Object algebra（k 个 constructor，t 个 operation 文件，m 个字段） ---
%% 多个 operation 文件各自继承同一个 data 文件，然后在最终组合时
%% merge。每个 constructor 在每个 operation 文件里各有一个定义，
%% 通过 overrides 合并，所以 overrides 大小为 O(t)。
%%
%% 关键点：this 的 frontier 是 1，不是"去重后 O(1)"，就是 1。
%% 原因：qualified this 解析的是外层 scope（如 NatFactory），
%% 而这个外层 scope 在继承链上只有一条路径到达——因为是从一个
%% 具体的 operation 定义点出发往上走的。multi-target this 的前提
%% 是同一个定义点被继承到多个不同的 site，这只在 trie union 时
%% 发生。Object algebra 的继承是把多个 operation merge 到同一个
%% factory 上，不是把多个 factory merge 成一个值。方向不同。
%%
%% 这也解释了与 Scala 的关系：Scala 拒绝 multi-target self-reference resolution
%%（Appendix B 的 "conflicting base types"），但在 object algebra
%% 场景下 this 本来就是 single-target 的，所以 Scala 的拒绝是
%% 过度保守——它拒绝了一类去重后（实际上根本不需要去重，因为
%% frontier 就是 1）完全合法的组合。
%%
%% 复杂度：O(k · m · t)，与 Scala trait linearization 一致。
%%
%% --- 场景 3：Nat/BinNat/Boolean 算术 ---
%% 数值的递归深度为 n（Nat 值的大小），每次递归触发一次 this
%% 解析（single-target），遍历 O(t) 个 override。
%% Nat 算术复杂度：O(n · t)，即 Peano 算术 + 常数因子。
%% BinNat 算术复杂度：O(log n · t)，即二进制算术 + 常数因子。
%%
%% --- 场景 4：Cartesian product / 逻辑编程 ---
%% 这是唯一出现真正 multi-target this 的场景。
%% trie union 的 bases 包含多条路径，每条路径的 addend 通过
%% this 解析到另一个 trie union，路径数相乘。
%% 对于 |R₁| × |R₂| 的 join，复杂度为 O(|R₁| · |R₂|)
%% （memoization 下）。
%%
%% 这与 B-Tree 引擎的 Datalog nested-loop join 复杂度一致。
%% trie 天然有索引——路径本身是前缀树，共享前缀的值自动共享
%% 计算（memoization 命中）。这与 Datalog 引擎对 B-Tree 做
%% prefix scan 在精神上一致。
%%
%% --- 总结 ---
%%
%% | 场景               | 复杂度              | 对比               |
%% |--------------------|---------------------|--------------------|
%% | λ-calculus         | O(n)                | = 环境机           |
%% | Object algebra     | O(k·m·t)            | = Scala线性化      |
%% | Nat 算术           | O(n·t)              | = Peano 算术       |
%% | BinNat 算术        | O(log n · t)        | = 二进制算术       |
%% | Cartesian product  | O(|R₁|·|R₂|)       | = nested-loop join |
%%
%% 每个场景都与对应的专用引擎复杂度一致。一阶性没有带来额外开销。
%%
%% Single-Path Lemma 的适用范围比论文明示的更大：不仅 λ-calculus
%% 翻译是 single-path，所有不涉及 trie union 的正常 OOP/object
%% algebra 代码都是 single-path。multi-target this 是 trie union
%% 的专属现象，而 trie union 对应的恰好是关系代数的 join，其
%% 复杂度是固有的，与引擎无关。
%% ================================================================

\section{Embedding the \texorpdfstring{$\lambda$}{λ}-Calculus}
\label{sec:translation}
\label{sec:forward-translation}

The following translation maps the $\lambda$-calculus in
A-normal form (ANF)~\cite{flanagan1993-essence-compiling-continuations} to inheritance-calculus.
In ANF, every intermediate result is bound to a name and each
expression contains at most one application.
We extend the standard ANF grammar with a \emph{value binding}
$\mathbf{let}\; x = V \;\mathbf{in}\; M$, which gives a name to a
value that is not an application result:%
\footnote{
  Standard ANF allows values inline as operands;
  the value-binding production names them so that every
  subexpression of the translated program is accessible
  via a lexical reference (Appendix~\ref{app:expressiveness}).
}
\[
  \begin{array}{r@{\;::=\;}l}
    M & \mathbf{let}\; x = V_1\; V_2 \;\mathbf{in}\; M
    \mid \mathbf{let}\; x = V \;\mathbf{in}\; M
    \mid V_1\; V_2
    \mid V \\
    V & x \mid \lambda x.\, M
  \end{array}
\]
Every $\lambda$-term can be mechanically converted to ANF by
naming intermediate results.\footnote{%
  This can optionally include naming values as well;
  since every ANF term is also a $\lambda$-term,
  ANF is neither a restriction nor an extension of the $\lambda$-calculus.
}
The ANF structure matches inheritance-calculus naturally: each
$\mathbf{let}$-binding becomes a named record property, and
each application is wrapped and projected via
$.\mathrm{result}$.

Let $\mathcal{T}$ denote the translation function.
Each $\lambda$-abstraction introduces one scope level
in the translated mixin tree.

\medskip
\begin{center}
  \small
  \begin{tabular}{l@{\quad$\longrightarrow$\quad}l}
    $x$ \text{\scriptsize($\lambda$-bound, index $n'$)} &
    $\dbi{n'}.\mathrm{argument}$ \\[4pt]
    $x$ \text{\scriptsize(let-bound)} &
    $x.\mathrm{result}$ \\[4pt]
    $\lambda x.\, M$ &
    $\{\mathrm{argument} \mapsto \{\},\;
    \mathrm{result} \mapsto \mathcal{T}(M)\}$ \\[4pt]
    $\mathbf{let}\; x = V_1\; V_2
    \;\mathbf{in}\; M$ &
    $\{x \mapsto \{\mathcal{T}(V_1),\;
      \mathrm{argument} \mapsto \mathcal{T}(V_2)\},$ \\
    & $\;\mathrm{result} \mapsto \mathcal{T}(M)\}$ \\[4pt]
    $\mathbf{let}\; x = V
    \;\mathbf{in}\; M$ &
    $\{x \mapsto \{\mathrm{result} \mapsto \mathcal{T}(V)\},\;
    \mathrm{result} \mapsto \mathcal{T}(M)\}$ \\[4pt]
    $V_1\; V_2$ \text{\scriptsize(tail call)} &
    $\{\mathrm{tailCall} \mapsto \{\mathcal{T}(V_1),\;
      \mathrm{argument} \mapsto \mathcal{T}(V_2)\},$ \\
    & $\;\mathrm{result} \mapsto \mathrm{tailCall}.\mathrm{result}\}$
  \end{tabular}
\end{center}
\medskip

A $\lambda$-bound variable $x$ with de~Bruijn index $n$
(the number of enclosing $\lambda$s between
the reference and its binder) becomes
$\dbi{n'}.\mathrm{argument}$,
which reaches the abstraction that binds $x$ and
accesses its $\mathrm{argument}$ slot.%
\footnote{\label{fn:debruijn-adjust}In the $\lambda$-calculus, de~Bruijn
  indices count only $\lambda$-abstractions.
  In inheritance-calculus, de~Bruijn indices count all
  enclosing scope levels, including those introduced by
  $\mathbf{let}$-bindings and tail calls.
  The translation therefore adjusts the index:
  if a $\lambda$-bound variable crosses $k$ intervening
  $\mathbf{let}$-binding or tail-call scopes
  (that is, $k$ counts the $\mathbf{let}$- and tail-call
    constructs on the path from the variable occurrence
  to its binding $\lambda$ in the ANF syntax tree),
  its inheritance-calculus
  index is $n' = n + k$,
  not~$n$.
  Let-bound variables are unaffected, as they use
  lexical references (such as $x.\mathrm{result}$) rather than
  de~Bruijn indices.
  We assume that the synthetic labels introduced by the
  translation ($\mathrm{argument}$, $\mathrm{result}$,
  $\mathrm{tailCall}$) and the user-defined labels
  (let-bound names such as~$x$) are drawn from disjoint
  alphabets, so no collision can arise.
}
A let-bound variable $x$ becomes the lexical reference
$x.\mathrm{result}$, projecting the application result
from the sibling property $x$ in the enclosing record.

An abstraction $\lambda x.\, M$ translates to a record with own
properties $\{\mathrm{argument}, \mathrm{result}\}$; we call this
the \emph{abstraction shape}.
In the application $\mathbf{let}$-binding rule, the name $x$
binds to the inherited record
$\{\mathcal{T}(V_1),\;\mathrm{argument} \mapsto \mathcal{T}(V_2)\}$,
and $\mathcal{T}(M)$ may reference $x.\mathrm{result}$ to obtain
the value of the application.
In the value $\mathbf{let}$-binding rule, $x$ wraps $\mathcal{T}(V)$
under a $\mathrm{result}$ child so that the uniform accessor
$x.\mathrm{result}$ yields $\mathcal{T}(V)$, the translated value
itself.
This extra wrapper is semantically inert but proof-relevant:
it lets the same $\mathrm{result}$-following argument used for
applications extend uniformly to value
$\mathbf{let}$-bindings in the adequacy proof
(Appendix~\ref{app:bohm-tree-proofs}).
In the tail-call rule, the fresh label $\mathrm{tailCall}$ serves
the same purpose, and $\mathrm{result}$ projects the answer.
In the application and tail-call cases, the inheritance that
performs the application is
encapsulated behind a named property, so its internal
structure (the $\mathrm{argument}$ and $\mathrm{result}$ labels
of the callee) does not leak to the enclosing scope.

The six rules above are a complete translation from the
$\lambda$-calculus (via ANF) to inheritance-calculus: every ANF
term has an image, and the construction is compositional.
Since the $\lambda$-calculus is Turing complete, so is
inheritance-calculus.%
\footnote{
  MIXINv2 separates symbol resolution (which resolves
  lexical references and qualified $\this$ references) from runtime
  mixin evaluation, but symbol resolution is performed
  just-in-time per symbol rather than ahead of time, so it does
  not provide static type checking in the traditional sense.
  Nevertheless, MIXINv2's resolution phase is stricter than the
  untyped inheritance-calculus.
  All examples in this paper are ported from MIXINv2
  and conform to both semantics.
  Elements added to satisfy the resolver are harmless
  for the algorithms presented here and may aid readability.
}

\subsection{B\"ohm Tree Correspondence}
\label{sec:bohm-tree}

The translation $\mathcal{T}$ embeds the $\lambda$-calculus
into a sublanguage of inheritance-calculus.
Intuitively, a mixin tree augments the AST with
additional semantic structure; for translated
$\lambda$-terms, the projection of this additional
structure is isomorphic to the B\"ohm tree.
A mixin tree $T$ \emph{converges}, written $T{\Downarrow}$,
if following finitely many $\mathrm{result}$ projections from
the root reaches a node with both $\mathrm{argument}$ and
$\mathrm{result}$ among its properties, that is, the \emph{abstraction shape}
produced by $\mathcal{T}(\lambda x.\, M)$.
The translation preserves and reflects convergence
(adequacy).
Write $\mathcal{T}(M) \approx_\lambda \mathcal{T}(N)$
when $\mathcal{T}(C[M]){\Downarrow} \Leftrightarrow
\mathcal{T}(C[N]){\Downarrow}$ for every closing
$\lambda$-calculus context $C[\cdot]$;
then $\approx_\lambda$ coincides with B\"ohm tree
equivalence~\cite{barendregt1984-lambda-calculus}.
Formal statements and proofs are in
Appendix~\ref{app:bohm-tree-proofs}.%
\footnote{
  Appendix~\ref{app:lambda-fixpoint} applies the same first-order
  technique directly on ANF to give the $\lambda$-calculus
  a self-contained fixpoint semantics, without passing
  through inheritance-calculus.
}
Note that $\approx_\lambda$ quantifies over $\lambda$-calculus
contexts only; inheritance-calculus contexts are strictly more
discriminating (Section~\ref{sec:asymmetry}).
B\"ohm tree equivalence is fully abstract for the
first-iteration approximation $T_P\!\uparrow\!1$ of the
equations.
The full $\mathrm{lfp}(T_P)$ is strictly more defined.
For example,
$\mathbf{let}\; x = x \;\mathbf{in}\; x$
diverges under $\beta$-reduction, but its translation
$\{x \mapsto \{\mathrm{result} \mapsto x.\mathrm{result}\},\;
\mathrm{result} \mapsto x.\mathrm{result}\}$
stabilizes under $\mathrm{lfp}(T_P)$:%
\footnote{
  This simple cycle could also be resolved by a
  fixed-point combinator such as~$Y$.
  The $\mathrm{lfp}(T_P)$ semantics is more general:
  it computes fixed points at the \emph{container} level,
  e.g.\ finding a set~$S$ such that $S = F(S)$.
  Recursive Datalog on cyclic graphs
  (Appendix~\ref{app:datalog-encoding},
  Section~\ref{sec:recursive-rules}) requires exactly this,
  and neither $Y$ nor Haskell's \texttt{mfix}
  (\texttt{RecursiveDo}) can express it.
}
tabled evaluation detects the cycle and returns
$\properties(\mathrm{root}) = \{x, \mathrm{result}\}$
with no abstraction shape, reflecting an undefined value
rather than diverging
(Remark~\ref{rem:convergence-lfp}).%
\footnote{
  We call the sublanguage of inheritance-calculus corresponding
  to $\lambda$-terms, evaluated under $\mathrm{lfp}(T_P)$ rather than
  $T_P\!\uparrow\!1$, the \emph{fixpoint $\lambda$-calculus};
  Remark~\ref{rem:convergence-lfp} elaborates.
}

The relationship between $\approx_\lambda$ and full
inheritance-contextual equivalence is clarified by the expressive
asymmetry of the next section:
contexts in the language universe $\mathscr{U}$
(Definition~\ref{def:language-universe}) that mix
$\lambda$-calculus and inheritance-calculus constructors
can separate $\lambda$-terms that are
B\"ohm-tree equivalent, so
${\approx_\lambda} \subsetneq {(\approx_{\mathscr{U}}|_{L_0})}$
where $L_0$ is the lazy $\lambda$-calculus.
This gap is precisely the source of inheritance-calculus's
additional expressiveness
(Theorem~\ref{thm:cartesian-nonexpressibility}).

\subsection{Expressive Asymmetry}
\label{sec:asymmetry}
\label{sec:formal-separation}
Two $\lambda$-terms that are B\"ohm-tree equivalent,
and therefore indistinguishable in the $\lambda$-calculus,
can be
separated by a context that mixes $\lambda$-calculus and
inheritance-calculus constructors, inheriting new observable
projections onto an existing definition via open extension.
By Felleisen's expressiveness
criterion~\cite{felleisen1991-expressive-power}, inheritance-calculus is
strictly more expressive than the lazy $\lambda$-calculus
(Theorem~\ref{thm:expressive-asymmetry}):
the $\lambda$-calculus can be macro-expressed in
inheritance-calculus via ANF conversion
and~$\mathcal{T}$
(Lemma~\ref{lem:anf-equiv}, Theorem~\ref{thm:forward-macro}),
but the lazy $\lambda$-calculus cannot macro-express inheritance
(Theorem~\ref{thm:cartesian-nonexpressibility}).
The formal definitions, constructions, and proofs are given in
Appendix~\ref{app:expressiveness};
the separating construction is itself a miniature instance of
the Expression Problem, adding a new observable projection to
an existing definition without modifying it.

The underlying mechanism is open extension: mixin
definitions compose via open extension, so existing
definitions gain new observable properties without
modification.  In the $\lambda$-calculus, a function's result
is closed: there is no way to add new observable projections
to an already-computed value without modifying the function's
definition.

\section{Case Study: Nat Arithmetic}
\label{sec:case-study}
\label{sec:expression-problem}

We implemented boolean logic and natural number arithmetic, both unary
and binary, as MIXINv2
files\selfcite{~\cite{mixin2025}}{~(supplementary material)}.%
\footnote{
  The examples in this section are rendered in
  inheritance-calculus notation; the executable MIXINv2 sources are
  available as supplementary material.
}
This section traces the unary Nat case in detail.
The implementation follows a declarative object-oriented style:
each concern is one top-level mixin;
UpperCamelCase names serve as data constructors and operations;
lowerCamelCase names serve as fields and parameters;
and since there are no methods, all names are nouns or adjectives: $\mathrm{Plus}$
is a structure, not an action.
Operations expose their result through a projection field (e.g., $\mathrm{sum}$).

\paragraph{NatData}
The data representation (Zero/Successor) is a Peano
encoding, an initial algebra of the functor $F(X) = 1 + X$.
The visitor pattern (NatVisitor, below) adds Church-encoding behavior:
a fold (catamorphism) over this algebra.
These two concerns are defined in separate files and composed via
inheritance\selfcite{~\cite{mixin2025}}{~(supplementary material)},
which is itself an instance of the Expression Problem.
Following this convention, we define a factory containing the abstract product type and two
labels that serve as data constructors.
A $\mathrm{Zero}$ value inherits $\mathrm{Product}$;
a $\mathrm{Successor}$ value inherits $\mathrm{Product}$ and
exposes a $\mathrm{predecessor}$ field.
The name $\mathrm{NatData}$ is the top-level scope:
\begin{align*}
  \mathrm{NatData} \mapsto \{\;&
    \mathrm{NatFactory} \mapsto \{
      \mathrm{Product} \mapsto \{\},\\
      &\quad
      \mathrm{Zero} \mapsto \{\mathrm{Product}\},\\
      &\quad
      \mathrm{Successor} \mapsto \{
        \mathrm{Product},\;
    \mathrm{predecessor} \mapsto \{\mathrm{Product}\}\}\},\\
    &\mathrm{Nat} \mapsto \{\mathrm{NatFactory},\;
  \mathrm{NatFactory}.\mathrm{Product}\}\}
\end{align*}

\paragraph{NatPlus}
Addition is a separate mixin that inherits $\mathrm{NatData}$
and extends each data constructor with a $\mathrm{Plus}$ operation
exposing a $\mathrm{sum}$ field.
The base case is $0 + m = m$; the recursive case reduces
$\mathrm{S}(n_0) + m$ to $n_0 + \mathrm{S}(m)$
by delegating to $n_0$'s $\mathrm{Plus}$ with an incremented addend:
\begin{align*}
  \mathrm{NatPlus} \mapsto \{\;&
    \mathrm{NatData},\\
    &\mathrm{NatFactory} \mapsto \{\\
      &\quad \mathrm{Product} \mapsto \{\mathrm{Plus} \mapsto \{
      \mathrm{sum} \mapsto \{\mathrm{Product}\}\}\},\\[2pt]
      &\quad \mathrm{Zero} \mapsto \{\mathrm{Plus} \mapsto \{
          \mathrm{addend} \mapsto \{\mathrm{Product}\},\;
      \mathrm{sum} \mapsto \{\mathrm{addend}\}\}\},\\[2pt]
      &\quad \mathrm{Successor} \mapsto \{\mathrm{Plus} \mapsto \{
          \mathrm{addend} \mapsto \{\mathrm{Product}\},\\
          &\qquad
          \mathrm{\_increasedAddend} \mapsto \{
            \mathrm{Successor},\;
          \mathrm{predecessor} \mapsto \{\mathrm{addend}\}\},\\
          &\qquad
          \mathrm{\_recursiveAddition} \mapsto \{
            \qualifiedthis{\mathrm{Successor}}.\mathrm{predecessor}.\mathrm{Plus},\\
            &\qquad\quad
          \mathrm{addend} \mapsto \{\mathrm{\_increasedAddend}\}\},\\
          &\qquad
          \mathrm{sum} \mapsto \{
  \mathrm{\_recursiveAddition}.\mathrm{sum}\}\}\}\}\}
\end{align*}

\paragraph{NatVisitor}
Equality requires case analysis on the data constructor; we use the
visitor pattern.
$\mathrm{NatVisitor}$ adds an $\mathrm{Acceptance}$ operation to each data constructor,
containing a $\mathrm{VisitorMap}$ with the appropriate
$\mathrm{ZeroVisitor}$ or $\mathrm{SuccessorVisitor}$ branch,
and an $\mathrm{Accepted}$ field that delegates to the matching branch:
\begin{align*}
  \mathrm{NatVisitor} \mapsto \{\;&
    \mathrm{NatData},\\
    &\mathrm{NatFactory} \mapsto \{\\
      &\quad \mathrm{Product} \mapsto \{\mathrm{Acceptance} \mapsto \{
      \mathrm{Accepted} \mapsto \{\}\}\},\\[2pt]
      &\quad \mathrm{Zero} \mapsto \{\mathrm{Acceptance} \mapsto \{
          \mathrm{VisitorMap} \mapsto \{
          \mathrm{ZeroVisitor} \mapsto \{\}\},\\
          &\qquad
      \mathrm{Accepted} \mapsto \{\mathrm{VisitorMap}.\mathrm{ZeroVisitor}\}\}\},\\[2pt]
      &\quad \mathrm{Successor} \mapsto \{\mathrm{Acceptance} \mapsto \{
          \mathrm{VisitorMap} \mapsto \{
          \mathrm{SuccessorVisitor} \mapsto \{\}\},\\
          &\qquad
  \mathrm{Accepted} \mapsto \{\mathrm{VisitorMap}.\mathrm{SuccessorVisitor}\}\}\}\}\}
\end{align*}
$\mathrm{BooleanData}$ defines the output sort:
\begin{align*}
  \mathrm{BooleanData} \mapsto \{\;&
    \mathrm{BooleanFactory} \mapsto \{
      \mathrm{Product} \mapsto \{\},\;
      \mathrm{True} \mapsto \{\mathrm{Product}\},\;
    \mathrm{False} \mapsto \{\mathrm{Product}\}\},\\
    &\mathrm{Boolean} \mapsto \{\mathrm{BooleanFactory},\;
  \mathrm{BooleanFactory}.\mathrm{Product}\}\}
\end{align*}
\paragraph{NatEquality}
\label{sec:nat-equality}
To test that $2 + 3 = 5$, we needed equality on Nat values.
The Cartesian product test
(page~\pageref{sec:case-study-cartesian}) will apply this
same $\mathrm{Equal}$ to a trie union of multiple Nat
values without modification.
$\mathrm{NatEquality}$ inherits $\mathrm{NatVisitor}$ and $\mathrm{BooleanData}$,
and adds an $\mathrm{Equal}$ operation to each data constructor.
$\mathrm{Zero}$ is equal only to another $\mathrm{Zero}$;
$\mathrm{Successor}$ is equal only to another $\mathrm{Successor}$
with an equal predecessor.
Case analysis on $\mathrm{other}$ is done via its $\mathrm{Acceptance}$:
{\small
  \begin{align*}
    \mathrm{NatEquality} \mapsto \{\;&
      \mathrm{NatVisitor},\; \mathrm{BooleanData},\\
      &\mathrm{NatFactory} \mapsto \{\\
        &\quad \mathrm{Product} \mapsto \{\mathrm{Equal} \mapsto \{
            \mathrm{other} \mapsto \{\mathrm{Product}\},\\
            &\qquad
            \mathrm{equal} \mapsto \{
        \qualifiedthis{\mathrm{NatEquality}}.\mathrm{Boolean}\}\}\},\\[2pt]
        &\quad \mathrm{Zero} \mapsto \{\mathrm{Equal} \mapsto \{
            \mathrm{other} \mapsto \{\mathrm{Product}\},\\
            &\qquad
            \mathrm{OtherAcceptance} \mapsto \{
              \mathrm{other}.\mathrm{Acceptance},\\
              &\qquad\;
              \mathrm{VisitorMap} \mapsto \{
                \mathrm{ZeroVisitor} \mapsto \{
                  \mathrm{equal} \mapsto \{
                \qualifiedthis{\mathrm{NatEquality}}.\mathrm{True}\}\},\\
                &\qquad\;\quad
                \mathrm{SuccessorVisitor} \mapsto \{
                  \mathrm{equal} \mapsto \{
              \qualifiedthis{\mathrm{NatEquality}}.\mathrm{False}\}\}\},\\
              &\qquad\;
              \mathrm{Accepted} \mapsto \{
                \mathrm{equal} \mapsto \{
            \qualifiedthis{\mathrm{NatEquality}}.\mathrm{Boolean}\}\}\},\\
            &\qquad
            \mathrm{equal} \mapsto \{
        \mathrm{OtherAcceptance}.\mathrm{Accepted}.\mathrm{equal}\}\}\},\\[2pt]
        &\quad \mathrm{Successor} \mapsto \{\mathrm{Equal} \mapsto \{
            \mathrm{other} \mapsto \{
              \mathrm{Product},\;
            \mathrm{predecessor} \mapsto \{\mathrm{Product}\}\},\\
            &\qquad
            \mathrm{RecursiveEquality} \mapsto \{
              \qualifiedthis{\mathrm{Successor}}.\mathrm{predecessor}.\mathrm{Equal},\\
              &\qquad\;
              \mathrm{other} \mapsto \{
            \qualifiedthis{\mathrm{Equal}}.\mathrm{other}.\mathrm{predecessor}\}\},\\
            &\qquad
            \mathrm{OtherAcceptance} \mapsto \{
              \mathrm{other}.\mathrm{Acceptance},\\
              &\qquad\;
              \mathrm{VisitorMap} \mapsto \{
                \mathrm{ZeroVisitor} \mapsto \{
                  \mathrm{equal} \mapsto \{
                \qualifiedthis{\mathrm{NatEquality}}.\mathrm{False}\}\},\\
                &\qquad\;\quad
                \mathrm{SuccessorVisitor} \mapsto \{
                  \mathrm{equal} \mapsto \{
              \mathrm{RecursiveEquality}.\mathrm{equal}\}\}\},\\
              &\qquad\;
              \mathrm{Accepted} \mapsto \{
                \mathrm{equal} \mapsto \{
            \qualifiedthis{\mathrm{NatEquality}}.\mathrm{Boolean}\}\}\},\\
            &\qquad
            \mathrm{equal} \mapsto \{
    \mathrm{OtherAcceptance}.\mathrm{Accepted}.\mathrm{equal}\}\}\}\}\}
\end{align*}}
%% === 审稿人备忘：关于代码"冗长"的回应 ===
%%
%% NatEquality 的定义看起来比等价的 Haskell 或 ML 模式匹配更长，
%% 但这种冗长与 point-free / 组合子风格的"简洁"形成的对比是表面的。
%% 需要区分两种"长"：
%%
%% 1. 自描述式的长：每个中间步骤都有名字（如 _recursiveEquality、
%%    OtherAcceptance），步骤写得详细。名字本身是零成本的语义锚点。
%% 2. 组合子膨胀式的长：如 SKI bracket abstraction 的指数级膨胀，
%%    在玩具例子里看起来精炼，但在真实程序中复杂度是 λ-calculus 的
%%    多项式甚至指数倍，而且没有涌现出新的计算模式（论文 Related Work
%%    已指出这一点）。
%%
%% Inheritance-calculus 的"冗长"属于第一种，而且是结构上的必要代价：
%% 每个有名字的中间步骤既是代码，也是 mixin 树上可路径寻址、可扩展的
%% 节点。这正是 open extension 的前提——无法扩展匿名的子表达式。
%% 如果加语法糖把名字隐藏（如同 M-表达式之于 S-表达式），反而会
%% 破坏这一性质。Lisp 社区的历史经验表明，S-表达式的"冗长"最终
%% 被证明是优势（代码即数据，宏系统直接操作统一结构），M-表达式
%% 的语法糖从未被真正需要。
%%
%% 在 AI 辅助编程的语境下，这种显式命名的风格尤其有利：
%% _recursiveEquality 这样的名字对 LLM 而言是自带的 chain-of-thought，
%% 比起让 AI 猜测 point-free 管道中第 n 个组合子的意图，
%% 每个自描述的步骤都降低了生成和理解的认知负荷。
%% 考虑到未来越来越多的代码将由 AI 生成和维护，
%% 多写几个 token 的显式名称不是成本，而是投资。
%%
%% 当然，如果确实需要更紧凑的表面语法，为 inheritance-calculus 加一个
%% ANF 预处理器（自动为匿名子表达式生成名字）并不困难，
%% 正如 ANF 变换本身是机械化的。但这是表面语法的选择，
%% 不影响演算本身的语义。
%% ================================================================
Composing $\mathrm{NatPlus}$ with $\mathrm{NatEquality}$ by
inheriting both, with no modifications to either mixin, gives the
resulting Nat values both $\mathrm{Plus}$ and $\mathrm{Equal}$
automatically.
The concrete numerals live in a shared mixin that inherits only $\mathrm{NatData}$:
\begin{align*}
  \mathrm{NatConstants} \mapsto \{\;&
    \mathrm{NatData},\\
    &\mathrm{One} \mapsto \{
      \qualifiedthis{\mathrm{NatConstants}}.\mathrm{NatFactory}.\mathrm{Successor},\\
      &\quad\quad
      \mathrm{predecessor} \mapsto \{
    \qualifiedthis{\mathrm{NatConstants}}.\mathrm{NatFactory}.\mathrm{Zero}\}\},\\
    &\mathrm{Two} \mapsto \{
      \qualifiedthis{\mathrm{NatConstants}}.\mathrm{NatFactory}.\mathrm{Successor},\\
      &\quad\quad
    \mathrm{predecessor} \mapsto \{\mathrm{One}\}\},\\
    &\mathrm{Three} \mapsto \{
      \qualifiedthis{\mathrm{NatConstants}}.\mathrm{NatFactory}.\mathrm{Successor},\\
      &\quad\quad
    \mathrm{predecessor} \mapsto \{\mathrm{Two}\}\},\\
    &\mathrm{Four} \mapsto \{
      \qualifiedthis{\mathrm{NatConstants}}.\mathrm{NatFactory}.\mathrm{Successor},\\
      &\quad\quad
    \mathrm{predecessor} \mapsto \{\mathrm{Three}\}\},\\
    &\mathrm{Five} \mapsto \{
      \qualifiedthis{\mathrm{NatConstants}}.\mathrm{NatFactory}.\mathrm{Successor},\\
      &\quad\quad
  \mathrm{predecessor} \mapsto \{\mathrm{Four}\}\}\}
\end{align*}
The arithmetic test inherits $\mathrm{NatConstants}$ together with the operations it needs:
\begin{align*}
  \mathrm{Test} \mapsto \{\;&
    \mathrm{NatConstants},\; \mathrm{NatPlus},\; \mathrm{NatEquality},\\
    &\mathrm{Addition} \mapsto \{
      \qualifiedthis{\mathrm{Test}}.\mathrm{Two}.\mathrm{Plus},\;
    \mathrm{addend} \mapsto \{\qualifiedthis{\mathrm{Test}}.\mathrm{Three}\}\},\\
    &\mathrm{Test2plus3} \mapsto \{
      \qualifiedthis{\mathrm{Test}}.\mathrm{Five}.\mathrm{Equal},\;
  \mathrm{other} \mapsto \{\mathrm{Addition}.\mathrm{sum}\}\}\}
\end{align*}
The path $\mathrm{Test}.\mathrm{Test2plus3}.\mathrm{equal}$
inherits $\mathrm{True}$.
No existing mixin was modified.
$\mathrm{NatEquality}$, a new operation, and $\mathrm{Boolean}$,
a new data type, each required only a new mixin: extensibility in both
dimensions of the Expression Problem~\cite{wadler1998-expression-problem}.

\paragraph{Tree-level inheritance and return types}
\label{sec:discussion-expression-problem}
Each operation is an independent mixin that extends a factory's
subtree.
Inheritance recursively merges subtrees that share a label,
so every label in the factory acquires all operations from all
inherited mixins.
Crucially, this applies to return types: $\mathrm{NatPlus}$ returns
values from $\mathrm{NatFactory}$, and $\mathrm{NatEquality}$
independently adds $\mathrm{Equal}$ to the same factory.
Inheriting both causes the values returned by $\mathrm{Plus}$ to
carry $\mathrm{Equal}$ automatically, without modifying either mixin.
In object algebras~\cite{oliveira2012-object-algebras} and finally
tagless interpreters~\cite{carette2009-finally-tagless}, independently defined
operations do not automatically enrich each other's return types;
additional boilerplate or type-class machinery is required.

Multi-target qualified $\this$ (equation~\ref{eq:this}) is essential.
In the test mixin above, $\mathrm{NatPlus}.\mathrm{Successor}$ and
$\mathrm{NatEquality}.\mathrm{Successor}$ both inherit
$\mathrm{NatData}.\mathrm{Successor}$.
When the two operations are composed, the resulting
$\mathrm{Successor}$ inherits the $\mathrm{NatData}$ schema
through two independent routes.
The reference
$\qualifiedthis{\mathrm{Successor}}.\mathrm{predecessor}$
inside each operation resolves through both routes, and since
inheritance is idempotent, all routes contribute the same
properties.
In Scala, the same pattern, where two independently defined inner
classes extend the same outer class's trait, triggers a
``conflicting base types'' rejection
(Appendix~\ref{app:scala-multi-target}), because DOT's self variable
is single-valued and cannot resolve through multiple inheritance
routes simultaneously.
The Expression Problem can still be solved in Scala using
object algebras or similar patterns, but composing independently
defined operations inherently produces multi-target $\this$ resolution to
shared schemas, which DOT's single-valued self variable cannot
accommodate.

\paragraph{Church-encoded Nats are tries}
\label{sec:case-study-cartesian}
A Church-encoded Nat is a record whose structure mirrors the
data-constructor path used to build it: $\mathrm{Zero}$ is a flat record;
$\mathrm{S}(\mathrm{Zero})$ is a record with a $\mathrm{predecessor}$
field pointing to a $\mathrm{Zero}$ record; and so on.
This is exactly the shape of a trie.%
\footnote{
  The unary Nat trie can be viewed as a sparse linked list
  whose depth equals the number it represents:
  in a trie union, some $\mathrm{Successor}$ nodes along
  the list point to $\mathrm{Zero}$ (leaf nodes),
  so not every position in the list is occupied.
  The\selfcite{ MIXINv2 source~\cite{mixin2025}}{
  supplementary material} includes a binary natural number
  (BinNat) arithmetic whose trie depth is logarithmic.
}
A scope that simultaneously inherits two Nat values is a trie union.
The following test applies $\mathrm{Plus}$ to a trie union of $\{1,2\}$
with addend a trie union of $\{3,4\}$, then compares the result to itself
with $\mathrm{Equal}$:
\begin{align*}
  \mathrm{CartesianTest} \mapsto \{\;&
    \mathrm{NatConstants},\; \mathrm{NatPlus},\; \mathrm{NatEquality},\\
    &\mathrm{OneOrTwo} \mapsto \{
      \qualifiedthis{\mathrm{CartesianTest}}.\mathrm{One},\;
    \qualifiedthis{\mathrm{CartesianTest}}.\mathrm{Two}\},\\
    &\mathrm{ThreeOrFour} \mapsto \{
      \qualifiedthis{\mathrm{CartesianTest}}.\mathrm{Three},\;
    \qualifiedthis{\mathrm{CartesianTest}}.\mathrm{Four}\},\\
    &\mathrm{Result} \mapsto \{
      \mathrm{OneOrTwo}.\mathrm{Plus},\;
    \mathrm{addend} \mapsto \{\mathrm{ThreeOrFour}\}\},\\
    &\mathrm{Check} \mapsto \{
      \mathrm{Result}.\mathrm{sum}.\mathrm{Equal},\;
  \mathrm{other} \mapsto \{\mathrm{Result}.\mathrm{sum}\}\}\}
\end{align*}
$\mathrm{Result}.\mathrm{sum}$ evaluates to the trie $\{4,5,6\}$, the set of all pairwise sums.
$\mathrm{Equal}$ then applies to every pair drawn from $\{4,5,6\} \times \{4,5,6\}$;
the path $\mathrm{CartesianTest}.\mathrm{Check}.\mathrm{equal}$
inherits both $\mathrm{True}$ and $\mathrm{False}$.
The Cartesian product arises directly from the semantic equations.
$\mathrm{OneOrTwo}$ inherits both $\mathrm{One}$ and $\mathrm{Two}$,
so $\bases(\mathrm{OneOrTwo})$ contains two paths.
When $\mathrm{Result}$ inherits
$\mathrm{OneOrTwo}.\mathrm{Plus}$, the $\bases$ of
$\mathrm{Result}$ include the $\mathrm{Plus}$ of both
$\mathrm{One}$ and $\mathrm{Two}$.
Each $\mathrm{Plus}$ resolves its $\mathrm{addend}$ via
$\this$, reaching $\mathrm{ThreeOrFour}$, which is itself a
trie union of $\mathrm{Three}$ and $\mathrm{Four}$:
the $\supers$ of the $\mathrm{addend}$ path contain both values.
Since $\mathrm{Successor}.\mathrm{Plus}$ recursively delegates to
$\mathrm{predecessor}.\mathrm{Plus}$ with an incremented addend,
and $\mathrm{predecessor}$ also resolves through the trie union,
the recursion branches at every trie node.
The result $\mathrm{sum}$ therefore collects all paths reachable
through any combination of the two operands, which is the
Cartesian product.

The same $\mathrm{Plus}$ and $\mathrm{Equal}$ files that operate on
single values thus produce, without modification, the relational
semantics of logic programming:%
\footnote{
  Appendix~\ref{app:datalog-encoding} gives the systematic
  encoding of Datalog into inheritance-calculus.
}
$\overrides$ and $\supers$ collect all inheritance paths, and
every operation distributes over them.

\section{Discussion}

\subsection{Emergent Phenomena}
\label{sec:emergent-phenomena}

In developing real programs in MIXINv2, three phenomena arose from off-label use of the adopted design patterns.%
\footnote{
  MIXINv2 programs adopt two standard patterns:
  data types follow a factory pattern isomorphic to Scala's
  \texttt{sealed trait} with \texttt{case class}es;
  pattern matching uses Scott encoding.
  Both are named after Gang-of-Four
  terminology~\cite{gamma1994-design-patterns}
  (factory, visitor, acceptance), with verbs replaced by
  nouns since the calculus has no methods.
}
They can be seen in the
source code\selfcite{~\cite{mixin2025}}{~(supplementary material)}
and the Case Study (Section~\ref{sec:case-study}).
Below we explain why they arise from the three primitives.

\begin{description}
  \item[Relational semantics]
    Single-valued computation and set-valued computation are
    the same mechanism.
    The semantic equations of Section~\ref{sec:mixin-trees}
    resolve every property to a set of paths, so operations
    defined on single values automatically distribute over
    all combinations of set-valued inputs
    (Section~\ref{sec:case-study}).
    When a mixin tree is viewed as a trie, the correspondence
    with relational algebra becomes precise.
    The carrier is a trie rather than a flat relation, and
    observation is coinductive: the observer drives computation
    by querying progressively deeper paths.
    This contrasts with
    Datalog's bottom-up materialization, but the relational
    structure of the computation is the same.
    Appendix~\ref{app:datalog-encoding} makes this correspondence
    precise with a systematic encoding of Datalog into
    inheritance-calculus, demonstrating that variable joins,
    constant matches, multi-body rules, and recursive rules are
    all expressible using only the three primitives of
    Section~\ref{sec:syntax}.

  \item[Immunity to nonextensibility]
    It is structurally impossible to write a nonextensible program.
    Every construct of inheritance-calculus
    (Section~\ref{sec:syntax}) is extensible via deep merge
    (equation~\ref{eq:overrides}).
    The Case Study (Section~\ref{sec:case-study}) and the
    Expressive Asymmetry (Section~\ref{sec:asymmetry};
    Theorem~\ref{thm:cartesian-nonexpressibility}) demonstrate
    and formalize this: new operations and data types are added,
    and new observable projections are introduced.
    This is the extensibility that the
    Expression Problem~\cite{wadler1998-expression-problem} demands.%
    \footnote{
      As an untyped calculus, inheritance-calculus cannot violate
      the extensibility requirement of the Expression Problem but does
      not encounter its type-safety requirement.
      A typed language built on
      inheritance-calculus\selfcite{~\cite{mixin2025}}{~(supplementary
      material)} inherits the immunity to nonextensibility and need
      only ensure that its type system does not introduce
      nonextensible constructs.
    }
    Programming languages typically provide a mix of extensible and
    nonextensible constructs, where extensibility is opt-in:
    frameworks must encode nonextensible constructs into extensible
    constructs using sophisticated
    mechanisms~\cite{liang1995-monad-transformers, lammel2003-scrap-your-boilerplate, loh2006-open-data-types,
      swierstra2008-data-types-a-la-carte, carette2009-finally-tagless, oliveira2012-object-algebras,
      kiselyov2013-extensible-effects, wu2014-effect-handlers-in-scope, kiselyov2015-freer-monads,
    wang2016-expression-problem-trivially, leijen2017-row-typed-algebraic-effects, poulsen2023-hefty-algebras},
    and code that uses nonextensible constructs
    cannot be extended unless rewritten in the framework.
    Section~\ref{sec:semantic-variants} analyzes why deep merge
    is the only composition mechanism that achieves this.

  \item[Function color blindness]
    \label{sec:practical-function-color}
    Business logic in MIXINv2 declares abstract property slots;
    different assemblies fill them with synchronous or
    asynchronous implementations via inheritance.
    Because the slot's color varies at assembly time without
    affecting the code that uses it, the same MIXINv2 file runs
    under both synchronous and asynchronous
    runtimes\selfcite{~\cite{mixin2025}}{~(supplementary
    material)}, an instance of function color
    blindness~\cite{nystrom2015-function-color}.
    The mechanism is dependency injection via inheritance:
    business logic declares empty slots, and the assembly
    inherits an FFI module that fills them, so the same
    business logic accepts implementations of any color.%
    \footnote{
      Function color~\cite{nystrom2015-function-color} arises when
      effectful functions have signatures $a \to f\,b$ (Kleisli
      arrows), requiring monadic bind to compose and making the
      effect $f$ visible at every call site.
      Because inheritance-calculus composes program structure
      rather than executing effects, this type distinction does
      not arise at the calculus level.
      The color is determined at assembly time by the choice of
      inherited implementation, and effect execution is delegated
      to the host language.
      Unstructured control flow, such as capturing and resuming
      continuations as in \texttt{call/cc}, cannot be abstracted
      this way, as it requires a global CPS transformation that
      inheritance-calculus does not express.
    }
\end{description}

\subsection{Semantic Alternatives}
\label{sec:semantic-variants}

The preceding item established that inheritance-calculus is immune
to nonextensibility.
We adopt this as a design goal and ask whether an alternative
design could share this property
by surveying every candidate known to us.

\paragraph{Degrees of freedom}
The analysis below assumes that the AST is a \emph{named tree}:%
every edge from parent to child carries a label, so that every
node is uniquely identified by the path from the root.
For configuration languages this is naturally the case: every
key in a JSON, YAML, or Nix attribute set is a label, and nesting
produces paths.%
\footnote{
  Nix is a functional language with first-class functions;
  its attribute sets alone do not form a named-tree DSL.
  The NixOS module system~\cite{nixosmodules}, however,
  restricts the user-facing interface to nested attribute sets
  composed by recursive merging, and it is this DSL-level
  structure that naturally forms a named tree.
}
For mainstream programming languages the AST is not directly a
named tree, since anonymous expression positions and implicit
scopes lack labels.
After ANF
transformation~\cite{flanagan1993-essence-compiling-continuations}, however, every intermediate
result is bound to a name, and the remaining anonymous positions%
\footnote{
  For example, the body of a $\mathbf{let}$.
}
can be assigned synthetic
fresh labels such as $\mathrm{result}$ and $\mathrm{tailCall}$
in Section~\ref{sec:forward-translation}, yielding a named
tree.

Under this premise, every language whose programs are such
named trees with references can be characterized by two finite sets:
a set~$\mathcal{N}$ of \emph{node types}, each determining how a
subtree organizes its children, and a set~$\mathcal{R}$ of
\emph{reference types}, each determining how one position in the
tree refers to another.
The syntax is fixed by the choice of $\mathcal{N}$ and
$\mathcal{R}$; the semantics is determined by how each node type
structures composition and how each reference type resolves its
target.

{\small
  \begin{center}
    \begin{tabular}{lp{0.38\columnwidth}p{0.33\columnwidth}}
      \textbf{Language}
      & \textbf{Node types $\mathcal{N}$}
      & \textbf{Reference types $\mathcal{R}$} \\
      \hline
      Java (OOP)
      & class, interface, method,
      block, conditional, loop,
      \texttt{throw}, lambda, \ldots
      & variable, field access, method call,
      \texttt{extends}/\texttt{implements},
      \texttt{import}, \ldots \\[3pt]
      Haskell (FP)
      & $\lambda$, application, \texttt{let},
      \texttt{case}, \texttt{data},
      type class, instance, \ldots
      & variable, constructor,
      type class dispatch,
      \texttt{import}, \ldots \\[3pt]
      Scala\footnote{
        Scala is a multi-paradigm language.
      }
      & class, trait, object, method,
      block, conditional, \texttt{match},
      \texttt{while}, \texttt{throw}, \ldots
      & variable, member access,
      \texttt{extends}, \texttt{with},
      implicits, \texttt{import}, \ldots \\
      \hline
      $\lambda$-calculus
      & abstraction, application
      & variable \\
      Inheritance-calculus
      & record
      & inheritance \\
    \end{tabular}
  \end{center}
}

\paragraph{Immunity to nonextensibility}
A path~$p$ in a program~$P$ is \emph{extensible} if there exists
a program~$Q$ such that composing $P$ with~$Q$ through a reference
makes new properties observable at~$p$ without modifying~$P$.
A language is \emph{immune to nonextensibility} if every path in
every well-formed program is extensible.

We now classify the design space to explain why we chose the
semantics of Section~\ref{sec:mixin-trees}.
Two axes remain free:
the \emph{composition mechanism}, which determines how the single reference
type incorporates definitions;
and the \emph{reference resolution}, which determines how the single reference
type finds its target.

\subsubsection*{The composition mechanism}

Every composition mechanism determines how the definitions of one
abstraction are incorporated into another.
The following table classifies the known candidates by whether
they satisfy commutativity~(C), idempotence~(I), and
associativity~(A), and by their extensibility model.

\begin{center}
  \begin{tabular}{lcccll}
    \textbf{Candidate} & \textbf{C} & \textbf{I} & \textbf{A}
    & \textbf{Extensibility} & \textbf{Representative} \\
    \hline
    Deep merge
    & $\checkmark$ & $\checkmark$ & $\checkmark$
    & universal & this paper \\
    Shallow merge
    & $\times$ & $\checkmark$ & $\checkmark$
    & top-level only & JS spread \\
    Conflict rejection
    & $\checkmark$ & $\checkmark$ & $\checkmark$
    & none on overlap
    & Harper--Pierce~\cite{harper1991-symmetric-record-concatenation} \\
    Priority merge
    & $\times$ & $\times$ & $\checkmark$
    & universal & Jsonnet~\texttt{+}, Nix~\texttt{//} \\
    $\beta$-reduction
    & $\times$ & $\times$ & $\times$
    & opt-in & Java, Haskell, Scala, \ldots \\
    Conditional
    & \multicolumn{3}{c}{causes divergence}
    & --- & --- \\
  \end{tabular}
\end{center}

\begin{description}
  \item[Deep merge]
    Merging same-label definitions recurses into their subtrees
    ($\overrides$, equation~\ref{eq:overrides}).
    Every label is an extension point, whether or not the
    original author anticipated extension.

  \item[Shallow merge]
    Nested labels cannot be independently extended.

  \item[Conflict rejection]
    Same-label definitions are rejected as errors.
    All three identities hold, but vacuously:
    the identities are satisfied because the compositions that
    would test them are refused.
    Conflict rejection prevents extending existing
    labels~\cite{harper1991-symmetric-record-concatenation}, precisely the composability that the
    Expression Problem~\cite{wadler1998-expression-problem} requires.

  \item[Priority merge]
    One source takes
    precedence~\cite{jsonnet,dolstra2010-nixos},
    so $A + B \neq B + A$, introducing
    an ordering dependency.

  \item[$\beta$-reduction]
    Only the parameters placed by the author are extension
    points; the function body is closed.
    To make a new aspect extensible, the original function must
    be rewritten to accept an additional parameter.
    Any composition mechanism based on function application
    inherits this limitation, because $\beta$-reduction
    closes the function body.
    The Expression Problem~\cite{wadler1998-expression-problem} is hard
    in such languages for the same reason:
    extensibility is opt-in.
    The frameworks that simulate immunity to nonextensibility
    (visitor pattern,
      object algebras~\cite{oliveira2012-object-algebras},
    finally tagless interpreters~\cite{carette2009-finally-tagless})
    are the machinery needed to work around this limitation.

  \item[Conditional incorporation]
    Removing or conditionally including definitions
    based on the presence of other definitions
    creates a dependency cycle:
    whether a definition exists depends on
    whether another definition exists, which in turn
    depends on the first.
    The recursive evaluation
    (Appendix~\ref{app:well-definedness}) cannot
    reach a fixed point when addition and removal
    alternate, causing divergence.
    This differs from the divergence inherent to
    Turing completeness: Turing-complete programs
    diverge because computation is unbounded, whereas
    conditional incorporation diverges because the
    immediate consequence operator $T_P$ oscillates
    rather than ascending monotonically.
\end{description}

\noindent
Among the candidates above, deep merge is the only one we found
that is commutative, idempotent, associative, and
immune to nonextensibility.
The other candidates each introduce a vulnerability to
nonextensibility or cause divergence.

\subsubsection*{The reference resolution}

Once deep merge over records is fixed as the composition
mechanism, four of the six semantic equations are determined
by the syntax:
$\properties$~(\ref{eq:properties}) reads the labels of a
record, $\bases$~(\ref{eq:bases}) reads its inheritance
sources, $\supers$~(\ref{eq:supers}) takes the transitive
closure, and $\overrides$~(\ref{eq:overrides}) implements
the merge.
The only remaining degree of freedom is how references
resolve: the $\resolve$~(\ref{eq:resolve}) and
$\this$~(\ref{eq:this}) equations.
We examine the three known candidates.

\begin{description}
  \item[Early binding]
    With early binding, references resolve at definition time to a
    fixed path, independently of how the record is later inherited.
    CUE~\cite{cue2019} exemplifies this: field references are
    statically resolved, and a nested field cannot refer to
    properties of its enclosing context contributed by later
    composition.
    This prevents the mixin from serving as a call stack for the
    $\lambda$-calculus: $\beta$-reduction requires the callee to
    observe the argument supplied at the call site, which is a
    later composition.
    With early binding, references are frozen before this
    composition occurs.
    In the equations,
    $\resolve$~(\ref{eq:resolve}) calls
    $\this$~(\ref{eq:this}), which walks the inherited
    tree structure; late binding is what allows
    $\this$ to see properties contributed by later-composed
    mixins.
    Without a Turing-complete mixin, functions must be
    reintroduced as the composition mechanism, re-exposing
    the opt-in extensibility problem above.

  \item[Dynamic scope]
    With dynamic scope, references resolve at the point of use
    rather than the definition site.
    Jsonnet~\cite{jsonnet} follows this approach: \texttt{self}
    always refers to the final merged object, not the object at
    the definition site.
    Union file systems exhibit the same property at the
    file-system level: a symbolic link's relative path resolves
    against the mount context at dereference time, not against the
    layer that defined the link.
    Dynamic scope breaks the $\lambda$-calculus embedding
    (Section~\ref{sec:forward-translation}).
    The Substitution Lemma (Lemma~\ref{lem:substitution})
    relies on the base case $M = y$ ($y \neq x$):
    the de~Bruijn index $m$ of~$y$ walks up from the
    definition-site path $\init(p_{\mathrm{def}})$ to a scope
    level different from~$x$'s binder, so the substitution
    $\mathrm{argument} \mapsto \mathcal{T}(V)$ at~$x$'s scope
    has no effect.
    With dynamic scope this separation fails.%
    \footnote{
      Counterexample:
      $(\lambda x.\, \lambda y.\, x)\; V_1\; V_2$.
      The reference to~$x$ inside $\lambda y.\, x$ translates to
      $\dbi{1}.\mathrm{argument}$ (de~Bruijn index~$1$).
      With late binding, $\this$ starts from the definition-site
      scope (the inner $\lambda$) and walks up one level to the
      outer $\lambda$, reaching the slot
      $\mathrm{argument} \mapsto \mathcal{T}(V_1)$.
      With dynamic scope, the reference resolves from the
      use-site context, where the innermost enclosing scope is
      the application $\{C_1.\mathrm{result},\;
      \mathrm{argument} \mapsto \mathcal{T}(V_2)\}$;
      walking up one level reaches this scope's
      $\mathrm{argument}$ slot, returning $\mathcal{T}(V_2)$
      instead of $\mathcal{T}(V_1)$.
      This is the classical variable-capture failure of dynamic
      scoping.
    }
    As with early binding, the mixin cannot serve as a complete
    replacement for functions.

  \item[Single-target $\this$]
    Single-target $\this$ is a promising candidate:
    the Single-Path Lemma~(\ref{lem:single-path}) shows that
    for translated $\lambda$-terms, the frontier set~$S$ in
    $\this$~(\ref{eq:this}) always contains exactly one path,
    so the $\lambda$-calculus embedding does not require
    multi-target resolution.
    However, single-target $\this$ is vulnerable to
    nonextensibility for a different reason.
    When multiple inheritance routes reach the same scope,
    single-target resolution must reject the situation or select
    one route.
    The NixOS module system~\cite{nixosmodules} raises a
    duplicate-declaration error
    (Appendix~\ref{app:scala-multi-target}).
    This rejects nested extensions where two independently authored
    modules define the same option, a composition that arises
    when solving the Expression
    Problem~\cite{wadler1998-expression-problem}
    (Section~\ref{sec:expression-problem}).

    Our first three implementations of MIXINv2 used
    standard techniques for $\this$ resolution: closure-based
    fixed points following Cook~\cite{cook1989-denotational-inheritance}, and
    stack-based environment lookup using de~Bruijn
    indices~\cite{debruijn1972-nameless-dummies}.
    All three produced subtle bugs when inheritance introduced
    multiple routes to the same enclosing scope.
    The difficulty was structural.
    A closure captures a single environment, so a fixed-point
    combinator solves for a single $\this$ target.
    A stack-based environment is a linear chain where each scope
    level has one parent.
    Both data structures embed a single-valued assumption
    incompatible with the multi-target situation that multiple
    inheritance naturally produces
    (Appendix~\ref{app:scala-multi-target}).
    Equation~(\ref{eq:this}) emerged from abandoning this
    assumption: it tracks a \emph{frontier set}~$S$ rather than a
    single current scope.
\end{description}

\noindent
The following table summarizes the vulnerabilities of all
alternatives at both levels.

\begin{center}
  \begin{tabular}{llll}
    & \textbf{Candidate} & \textbf{Vulnerability}
    & \textbf{Representative} \\
    \hline
    \multirow{5}{*}{\rotatebox[origin=c]{90}{\scriptsize composition}}
    & Shallow merge
    & nested extensions not composable
    & JS spread \\
    & Conflict rejection
    & same-label extensions rejected
    & Harper--Pierce~\cite{harper1991-symmetric-record-concatenation} \\
    & Priority merge
    & ordering dependency
    & Jsonnet~\texttt{+}~\cite{jsonnet},
    Nix~\texttt{//}~\cite{dolstra2010-nixos} \\
    & $\beta$-reduction
    & opt-in extensibility
    & Java, Haskell, Scala, \ldots \\
    & Conditional
    & causes divergence
    & --- \\
    \hline
    \multirow{3}{*}{\rotatebox[origin=c]{90}{\scriptsize resolution}}
    & Early binding
    & mixin cannot replace function
    & CUE~\cite{cue2019} \\
    & Dynamic scope
    & mixin cannot replace function
    & Jsonnet~\cite{jsonnet}, union FS \\
    & Single-target $\this$
    & nested extensions rejected
    & NixOS modules~\cite{nixosmodules} \\
  \end{tabular}
\end{center}

\noindent
No alternative we examined at either level is immune to
nonextensibility.

We arrived at this semantics by trying each alternative in this section
and finding that it failed:
shallow merge could not compose nested extensions;
priority merge introduced ordering dependencies;
early binding froze references before composition could
supply them;
dynamic scope broke the substitution that the
$\lambda$-calculus embedding relies on;
single-target $\this$ rejected the multi-target compositions
that the Expression Problem demands.
Deep merge with multi-target late-binding $\this$ is what
remained.

\section{Related Work}
\label{sec:related-work}

\subsection{Computational Models and \texorpdfstring{$\lambda$}{λ}-Calculus Semantics}

\paragraph{Minimal Turing-complete models}
Sch\"onfinkel~\cite{schonfinkel1924-combinatory-logic} and
Curry~\cite{curry1958-combinatory-logic} showed that the
$\lambda$-calculus reduces to three combinators $S$, $K$, $I$
\footnote{
  The $I$ combinator is redundant since $I = SKK$.
}.
The bracket abstraction that compiles $\lambda$-terms to SKI
is mechanical but incurs exponential blow-up in term size;
no new computational patterns emerge beyond those already
present in the $\lambda$-calculus.
Semantically, the two share the same models:
Meyer~\cite{meyer1982-model-of-lambda-calculus} showed that combinatory algebras
satisfying five additional conditions are exactly lambda
algebras.
No independent semantic domain for SKI has ever been
exhibited from which $\lambda$-calculus semantics could be
derived. The direction has always been the reverse:
Scott's $D_\infty$ was built for the $\lambda$-calculus,
and SKI was interpreted inside it afterwards.
Dolan~\cite{dolan2013-mov-turing-complete} showed that the x86 \texttt{mov}
instruction alone is Turing complete by exploiting
addressing modes as implicit arithmetic on mutable memory;
the resulting programs are correct but provide no
abstraction mechanisms beyond those of the RAM Machine.
The Turing Machine itself, while universal, lacks random
access: reading a cell at distance~$d$ requires $d$
sequential tape movements.
Each of these models achieves Turing completeness
within the semantic framework of an existing model,
and no independent semantic domain has been exhibited
for any of them.

%% === 审稿人备忘：SF-calculus 与 inheritance-calculus 的关系 ===
%%
%% Barry Jay 的 SF-calculus（Intensional Computation with Higher-Order
%% Functions, TCS 2019, doi:10.1016/j.tcs.2019.02.016；前身为
%% Confusion in the Church-Turing Thesis, arXiv:1410.7103, 2014）
%% 和 inheritance-calculus 都试图解决 λ-calculus 的 extensionality 限制，
%% 但方向完全不同。
%%
%% 共同出发点：λ-calculus 是 extensional 的——只能观察函数的输入输出行为，
%% 无法检查程序的内部结构。两个内部结构不同但行为相同的 λ 项不可区分。
%%   - Jay 的诊断：λ-calculus 缺乏 intensional 操作（检查项的结构），
%%     无法定义结构等式判定。
%%   - 本文的诊断：λ-calculus 的 function body 是 closed 的，无法从外部
%%     添加新的可观察投影（open extension）。
%%
%% 核心机制对比：
%%   SF-calculus：保留 S、K（或 λ），添加 F（factorisation）算子，
%%     可以将闭合范式拆解为组成部分。程序仍然是函数，只是多了检查函数
%%     结构的能力。属于 λ-calculus 的 *扩展*。
%%   Inheritance-calculus：用三个不同的原语（record、definition、
%%     inheritance）替代 λ-calculus 的三个原语。β-reduction 是 deep merge
%%     的特例（Section 3 的翻译 T 将函数应用映射为继承）。
%%     属于 λ-calculus 的 *包含*（subsumption）。
%%
%% 关于 F 算子的编码：
%%   F 做两件事：(1) 判断结构形态（atom vs. compound），
%%   (2) 提取子部分。
%%   在 inheritance-calculus 中：
%%     - 提取子部分是平凡的：所有值都是透明 record，通过 qualified this
%%       投影任意 slot 即可。Theorem 6 的 separating context 就在读取
%%       argument slot（↑².argument）。
%%     - 判断结构形态并分支：需要 Acceptance pattern（Section 4）。
%%       NatVisitor 对 Nat 做的事情与 F 对 SF-calculus 项做的事情
%%       完全对应：VisitZero/SuccessorVisitor 对应 F 的 atom/compound
%%       分支。Visitor 可以通过 open extension 添加到已有 term 上，
%%       不修改原始定义。
%%   Acceptance 在精神上就是 F：两者都是对 closed normal form 做
%%   constructor dispatch。区别在于 F 是不可扩展的 built-in primitive
%%   （硬编码了 S 和 F 两个 atom），而 Acceptance 是可扩展的 encoding
%%   （新增 constructor 只需继承新的 Accepted* 分支，即 Expression Problem
%%   immunity）。
%%
%% 精确差异：
%%   SF-calculus 把 intensional observation 做成了 object-level 的
%%   primitive reduction rule（F 是一个 term）；
%%   inheritance-calculus 把它留在了 meta-level（观察者总能通过
%%   properties(p) 查询任何路径），而 term-level 的分支需要通过
%%   Acceptance pattern 显式编码。
%%   换言之：inheritance-calculus 中 F 的 *读取* 能力是免费的
%%   （所有值都是透明 record），但 F 的 *分支* 能力需要 Acceptance。
%%   SF-calculus 的 F 则把读取和分支捆绑在一个不可扩展的 primitive 里。
%%   Acceptance 是可扩展的（新增 constructor 只需继承新分支，
%%   即 Expression Problem immunity），而 F 硬编码了固定的 atom 集合
%%   （S 和 F）。
%%
%% 关于 Jay 所谓的 "trivial solution of the Halting Problem"：
%%   Jay 用非标准编码将每个程序（包括不停机的）映射到闭合范式（静态数据），
%%   然后结构等式判定变得平凡。这并不推翻图灵定理——它改变了问题的
%%   形式化方式（把程序当静态数据而非可执行代码），而非解决了原来的问题。
%%   Jay 的真正论点是：λ-definability 和 Turing computability 不是同一回事，
%%   Church-Turing Thesis 的等价性证明隐含了编码假设。
%%
%% 正文篇幅不够论述 embeds SF-calculus，暂不展开。
%% 如果审稿人提到 SF-calculus，可以基于上述分析撰写回复或补充段落。
%% ================================================================

\paragraph{Semantic descriptions of the $\lambda$-calculus}
The semantics of the $\lambda$-calculus can be defined in
several fundamentally different ways.
\emph{Operational semantics} takes $\beta$-reduction
$(\lambda x.\,M)\,N \to M[N/x]$ as the primitive notion of
computation~\cite{church1936-lambda-calculus,church1941-calculi-of-lambda-conversion,barendregt1984-lambda-calculus}.
This requires capture-avoiding substitution, a mechanism whose
subtleties motivated de~Bruijn
indices~\cite{debruijn1972-nameless-dummies} and explicit substitution
calculi~\cite{abadi1991-explicit-substitutions}.
It also requires a choice of evaluation
strategy: call-by-name versus
call-by-value~\cite{plotkin1975-call-by-name-call-by-value} or the lazy
strategy of Abramsky~\cite{abramsky1990-lazy-lambda-calculus}.
\emph{Denotational semantics}~\cite{scott1970-mathematical-theory-of-computation,scott1976-data-types-as-lattices}
interprets terms in a mathematical domain, but giving meaning
to $\beta$-reduction requires a reflexive domain
$D \cong [D \to D]$, which has no set-theoretic solution by
Cantor's theorem. Scott's $D_\infty$ resolves this by
constructing $D$ as an inverse limit of continuous
lattices~\cite{scott1972-continuous-lattices}.
The resulting semantics is adequate but not fully
abstract~\cite{plotkin1977-lcf-as-programming-language,milner1977-fully-abstract-models}:
it identifies terms that observational equivalence
distinguishes.
\emph{Game semantics}~\cite{abramsky2000-full-abstraction-pcf,hyland2000-full-abstraction-pcf,nickau1994-hereditarily-sequential-functionals}
achieves full abstraction for PCF and for the lazy
$\lambda$-calculus~\cite{abramsky1993-full-abstraction-lazy-lambda,abramsky1995-games-full-abstraction-lazy-lambda},
requiring arenas, strategies, and innocence conditions.
In each of these approaches, the function space $[D \to D]$
is the central obstacle: operational semantics needs it
implicitly via substitution. In contrast, denotational semantics
constructs it explicitly, and game semantics refines it via
strategies.
Each of these approaches involves higher-order structure%
\footnote{
  Function spaces
  in denotational semantics, strategies
  in game semantics, and substitution
  in operational semantics.
}
and retains the function space in its semantic domain.
The denotational and operational accounts remain separate
formalisms in each case.
All three inherit $\beta$-reduction's divergence on
recursive programs: a na\"ive transitive closure on a cyclic
graph, for instance, diverges in every framework.

\paragraph{Set-theoretic models of the $\lambda$-calculus}
Several families of $\lambda$-calculus models avoid the
inverse-limit construction of Scott's $D_\infty$ by working
in set-theoretic rather than domain-theoretic settings.
These models eliminate Scott topology but not the function
space itself: they remain higher-order, encoding $[D \to D]$
inside $D$ via injections, intersection types, or multiset
relations.
\emph{Graph models}~\cite{engeler1981-algebras-and-combinators,plotkin1993-set-theoretic-lambda-models}
encode functions via a web injection on powerset lattices;
Bucciarelli and Salibra~\cite{bucciarelli2008-graph-lambda-theories}
proved that the greatest sensible graph theory equals the
B\"ohm tree theory~$\mathcal{B}$.
\emph{Filter models}~\cite{barendregt1983-filter-lambda-model,coppo1978-intersection-type-assignment}
interpret a term as the set of intersection types derivable
for it;
Dezani-Ciancaglini et al.~\cite{dezani1998-intersection-types-bohm-trees}
showed that the BCD filter model's $\lambda$-theory
coincides exactly with B\"ohm tree equality.
\emph{Relational models}~\cite{ehrhard2005-finiteness-spaces,bucciarelli2001-phase-semantics-exponentials}
arise from the relational semantics of linear logic via
finite multisets;
Ehrhard~\cite{ehrhard2005-finiteness-spaces} showed that a standard
reflexive object for the pure $\lambda$-calculus cannot
exist in this category.
All three families are semantic models: they interpret
$\lambda$-terms in an external mathematical structure and
validate $\beta$-reduction as an equation in that structure,
rather than defining an independent calculus with its own
primitives.
Each encodes the function space rather than eliminating it,
and each requires higher-order machinery to obtain full
abstraction for the lazy $\lambda$-calculus.

\paragraph{Recursive equations and fixpoint semantics}
Van~Emden and Kowalski~\cite{vanemden1976-predicate-logic-semantics} showed that
the semantics of predicate logic as a programming language admits
three equivalent characterizations:
operational\footnote{
  Typically SLD resolution.
}, model-theoretic\footnote{
  Based on least Herbrand models.
}, and fixed-point\footnote{
  Namely, the least fixed point of the immediate consequence operator~$T_P$ on the powerset of ground atoms.
}.
Aczel~\cite{aczel1977-inductive-definitions} gave a systematic treatment of
inductive definitions via monotone operators on powerset lattices,
providing the logical foundations that Datalog-style semantics
rests on.
Leroy and Grall~\cite{leroy2006-coinductive-big-step} treated divergence
in big-step semantics via coinductive definitions,
requiring a separate coinductive judgment to account for
non-termination alongside the inductive evaluation judgment.
Van~Emden and Kowalski's $T_P$ computes
$\mathrm{lfp}$ for recursive Datalog programs over a
first-order domain.
Conventional $\lambda$-calculus semantics requires function
spaces in the semantic domain, and recursive $\lambda$-terms
ordinarily diverge under $\beta$-reduction;
using $T_P$ over a first-order domain to give the
$\lambda$-calculus a fixpoint semantics has not been explored.

\subsection{Inheritance, Objects, and Records}

\paragraph{Denotational semantics of inheritance}
Cook~\cite{cook1989-denotational-inheritance} gave the first denotational
semantics of inheritance, modeling objects as recursive records.
Inheritance is composition of \emph{generators}
\footnote{
  These are functions from self to complete object.
} and
\emph{wrappers}\footnote{
  Wrappers are functions that modify generators.
},
solved by a fixed-point construction.
Cook's key insight is that inheritance is a general mechanism
applicable to any form of recursive definition, not only
object-oriented methods. This insight is one of the starting points of the
present work.
The model has four primitives (records, functions, generators, and
wrappers) and requires a fixed-point combinator. Composition is
asymmetric: a wrapper modifies a generator but not vice versa.
The fixed-point construction relies on Scott continuity in a
domain of functions. The resulting denotations are not directly
executable, so a separate operational semantics is needed to
define method dispatch and object construction at runtime.

\paragraph{Mixins and traits}
Bracha and Cook~\cite{bracha1990-mixin-inheritance} formalized mixins as
abstract subclasses, that is, functions from a superclass parameter to a
subclass.
Because mixin application is function composition, it is
\emph{neither commutative nor idempotent}; the C3 linearization
algorithm~\cite{barrett1996-c3-linearization} was later developed to impose a
deterministic order.
Sch\"arli et al.~\cite{scharli2003-traits} introduced traits,
whose composition is commutative but rejects same-name
conflicts~\cite{ducasse2006-traits-fine-grained-reuse}.
Both mechanisms operate at the \emph{method level}: composing two
definitions of the same name results in an override or a conflict,
not a recursive merge of nested structure.
Both presuppose the $\lambda$-calculus: method bodies are functions,
and the composition mechanism itself is not Turing complete.
Neither mixins nor traits are deep-mergeable.

\paragraph{Object and record calculi}
Abadi and Cardelli~\cite{abadi1996-theory-of-objects} treat objects as the
primitive notion, but methods are functions that take self as a
parameter and object extension is asymmetric.
Boudol~\cite{boudol2004-recursive-record-semantics} showed that self-referential
records require an unsafe fixed-point operator whose
well-definedness depends on the evaluation strategy.
Harper and Pierce~\cite{harper1991-symmetric-record-concatenation} gave a record
calculus with symmetric concatenation that rejects same-label
composition;
Cardelli~\cite{cardelli1992-extensible-records} studied extensible records
with subtyping;
R\'emy~\cite{remy1989-row-polymorphism} introduced row polymorphism for
records and variants in ML.
All of these calculi operate on \emph{flat} records: there is
no recursive merging of nested structure, no lazy observation,
and $\this$ requires an explicit fixed-point combinator.

\paragraph{Family polymorphism and DOT}
Ernst~\cite{ernst2001-family-polymorphism} introduced family polymorphism;
Ernst, Ostermann, and Cook~\cite{ernst2006-virtual-class-calculus}
formalized it in the virtual class calculus, where classes are
accessed via path expressions \texttt{this.out.C}.
Amin et al.~\cite{amin2016-dependent-object-types} formalized path-dependent types
in the DOT calculus, the theoretical foundation of Scala.
In both frameworks, each path resolves to a
\emph{single} class in a \emph{single} enclosing object.
When multiple inheritance routes lead to the same scope,
single-valued resolution must choose one route and discard the
others; DOT's self variable $\nu(x : T)\, d$ binds $x$ to a
single object, and Scala rejects the resulting intersection
type at compile time
(Appendix~\ref{app:scala-multi-target}).
Neither framework provides a mechanism for set-valued
resolution.

\subsection{Configuration Languages and Module Systems}

\paragraph{Configuration languages}
CUE~\cite{cue2019} unifies types and values in a single lattice
where unification, denoted by \texttt{\&}, is commutative, associative, and
idempotent.
CUE's design was motivated by Google's Borg Configuration
Language (GCL), which used graph unification.
CUE's lattice includes scalar types with a conflict
semantics where unifying incompatible scalars yields $\bot$; whether
scalars are necessary for a configuration language, or whether
tree structure alone suffices, is an interesting design question.
Jsonnet~\cite{jsonnet} provides object inheritance via the
\texttt{+} operator with mixin semantics: composition is
\emph{not} commutative since the right-hand side wins on scalar
conflicts, and deep merging requires explicit opt-in via the \texttt{+:}
syntax on a per-field basis, so deep merge is
available but not the default.
Dhall~\cite{dhall2017} takes a functional approach: it is a typed
$\lambda$-calculus with records, where composition is a record merge
operator, not an inheritance mechanism.

\paragraph{Module systems and deep merge}
The NixOS module system~\cite{dolstra2010-nixos,nixosmodules}
composes modules by \emph{recursively merging nested attribute sets},
not by linearization or method-level override, which is the same deep-merge
semantics that distinguishes tree-level inheritance from method-level
composition in prior mixin and trait calculi.
When two modules define the same nested path, their subtrees are
merged rather than overridden.
The NixOS module system incorporates deep merge,
recursive $\this$, deferred evaluation of modules, and a
rich type-checking layer on top.
The same mechanism powers
NixOS system configuration, Home
Manager~\cite{homemanager}, nix-darwin~\cite{nixdarwin},
flake-parts~\cite{flakeparts}, and dozens of other ecosystems,
collectively managing configurations of arbitrary complexity.
The one aspect that appears not to have been fully addressed is the
semantics of $\this$ under multiple inheritance.
When two independent modules that inherit from the same
deferred module are composed, each import introduces its own
declaration of the shared options; the module system's
\texttt{merge\-Option\-Decls} rejects duplicate declarations
with a static error (Appendix~\ref{app:scala-multi-target}).
The module system cannot express that both routes contribute
equally to the same option, that is, the multi-target situation that
recursive merging naturally gives rise to, which motivated the
observational semantics of Section~\ref{sec:mixin-trees}.

\subsection{Our Companion Work}

\paragraph{Modular software architectures}
\label{sec:practical-modular}
Plugin systems, component frameworks, and dependency injection frameworks
enable software extensibility through declarative inheritance. The
symmetric, associative nature of inheritance (Section~\ref{sec:mixin-trees}) explains why such
systems work: components can be inherited in any order without changing
behavior.
MIXINv2\selfcite{~\cite{mixin2025}}{~(supplementary material)}
is itself a dependency injection framework:
each scope declares its dependencies as named slots, and
composition resolves them by name across scope boundaries,
making the inheritance-calculus patterns explicit rather than ad~hoc.

\paragraph{Union file systems}
\label{sec:practical-union-fs}
Union file systems, such as UnionFS, OverlayFS, and AUFS, layer multiple directory
hierarchies to present a unified view. Their semantics resembles
inheritance-calculus in that later layers override earlier layers and files from all
layers remain accessible, but their designs differ in specific ways:
inheritance is asymmetric, conflict resolution for scalar files
is noncommutative, and there is no late-binding or dynamic
dispatch across layers.
We implemented inheritance-calculus directly as a union file
system\selfcite{~\cite{ratarmount2025}}{~(included as supplementary material)},
obtaining late-binding semantics and dynamic dispatch for file lookups
with no additional mechanism beyond the three primitives.
In package managers, build systems, and OS distributions,
external toolchains transform configuration into file trees;
here, the file system serves as configuration to itself,
and the transformation is inheritance: a
\emph{file-system-as-compiler} in which the directory tree is
both the source program and the compilation target.

\section{Future Work}
\label{sec:future-work}

\selfcite{MIXINv2~\cite{mixin2025}}{The supplementary
implementation} is an executable implementation of inheritance-calculus
that already goes beyond the untyped calculus presented here:
it includes compile-time checking that all references resolve to
valid paths in the mixin tree, and a foreign-function interface (FFI)
that introduces scalar values from the host language.
The future work described below concerns the distance between this
implementation and a fully practical language, spanning both
theoretical foundations and library-level encodings.

\subsection{Type System}
MIXINv2's existing compile-time checks verify that every
reference path resolves to a property that exists in the inherited
result. Formalizing the soundness of these checks, that is, proving that
well-typed programs do not produce dangling references at
runtime, is one direction of future work.
Such a formalization would relate to the DOT
calculus~\cite{amin2016-dependent-object-types} in a manner analogous to how
inheritance-calculus relates to the $\lambda$-calculus: a Typed
inheritance-calculus can be viewed as DOT without $\lambda$, retaining
path-dependent types while replacing functions with
mixin inheritance.
A formal type system would be amenable to mechanization in proof
assistants, such as Coq and Agda, for verifying type safety, progress, and
preservation properties.

Beyond soundness of the existing checks, additional type system
features would strengthen the language.
A key example is \emph{totality checking}: verifying that an
inheritance provides implementations for all required slots, not
merely that references resolve.
In untyped inheritance-calculus, writing $\mathrm{magnitude} \mapsto \{\}$
is purely documentary: it defines a structural slot but imposes no
constraint on what must fill it.
Totality checking would enforce such declarations statically,
rejecting inheritances that leave required slots unfilled.

\subsection{Standard Library}
The remaining items concern the standard library: they require
no extensions to the calculus or type system, but involve nontrivial
encodings within the existing framework.

Inheritance-calculus is \emph{open by default}: inheritance
can freely merge any two mixins, and a value may simultaneously
inhabit multiple data constructors (e.g.,
$\{\mathrm{Zero},\; \mathrm{Odd},\; \mathrm{half} \mapsto \mathrm{Zero}\}$).
This is the natural trie semantics of the
calculus, and the basis for solving the Expression Problem
(Section~\ref{sec:expression-problem}).
However, many operations, such as equality testing, assume that values
are \emph{linear}, inhabiting exactly one data constructor.
The observer pattern (Section~\ref{sec:case-study}) does not enforce
this: applied to a multi-data-constructor value, all callbacks fire and
their results are inherited.

\emph{Schema validation} can be implemented at the library level
using existing primitives.
An observer can test whether a particular data constructor is present,
returning a Boolean; a second Boolean dispatch then branches on the
result.
By chaining such tests, a factory can validate that a value matches
exactly one data constructor and carries the required fields.
This does not restrict inheritance itself, which remains
unconstrained, but provides a way to detect invariant violations
before they propagate.

The same technique, an observer returning Boolean followed by Boolean
dispatch, yields \emph{closed pattern matching}.
The observer pattern of Section~\ref{sec:case-study} is inherently
\emph{open}: new data-constructor cases can be added through inheritance.
Closed matching, where exactly one branch is taken, can be encoded as
a \emph{visitor chain}: a linked list of if-then-else nodes, each
testing one data constructor and falling through on mismatch.
This is analogous to GHC.Generics' $\mathrm{(:+:)}$ sum
representation~\cite{magalhaes2010-generic-deriving}, where each link
corresponds to one summand.
An important consequence is that such chains do \emph{not}
commute, as they have a fixed order, and therefore cannot be extended
through open inheritance, which is the correct semantics for closed
dispatch.

\section{Conclusion}

Immunity to nonextensibility guided the design of
inheritance-calculus.
Section~\ref{sec:semantic-variants} eliminated every
alternative composition mechanism and reference resolution,
leaving deep merge with multi-target late-binding qualified
$\this$ as the unique survivor.
The mixin served as the single abstraction unifying module,
class, object, method, field, and local binding;
deep merge of mixin trees made inheritance commutative, idempotent, and associative,
so the linearization problem of multiple inheritance did not arise.
The six first-order equations of Section~\ref{sec:mixin-trees},
denotational and computable by
tabling~\cite{tamaki1986-tabled-resolution}, suffice to
define the semantics; the observer drives computation by
querying paths in a lazily constructed tree.
These equations are a minimal departure from standard
object-oriented semantics: the only change is the $\overrides$ equation~(\ref{eq:overrides}),
which replaces method-level shadowing with deep merge of subtrees.
Object-oriented programming is traditionally classified as
imperative; removing mutable state and function bodies yields
pure object-oriented programming,%
\footnote{
  ``Pure'' carries a double meaning:
  not impure, since values are immutable and composition has no
  side effects; and not mixed, since composition relies entirely on
  inheritance, without functional abstractions such
  as~$\lambda$ or~$\beta$-reduction.
}
a minimal computational model for declarative programming
that fills the empty cell of Table~\ref{tab:computational-models}.

Since $\overrides$ merges subtrees rather than method bodies,
functions are no longer primitive.
The $\lambda$-calculus macro-expressed into
inheritance-calculus in six translation rules in
Section~\ref{sec:translation}.
The B\"ohm tree correspondence of
Section~\ref{sec:bohm-tree} extended this first-order
semantics to the $\lambda$-calculus:
the prior higher-order semantic frameworks proved faithful
to the first-iteration approximation of a sublanguage of
inheritance-calculus's first-order equations;
Appendix~\ref{app:lambda-fixpoint} reformulated this
semantics directly on ANF\@.
A convergence hierarchy among the eager, lazy, and fixpoint
$\lambda$-calculi sharing the same $\lambda$-syntax followed
as a corollary, detailed in Remark~\ref{rem:convergence-lfp}.

Inheritance-calculus further gains three phenomena
that the $\lambda$-calculus cannot express;
indeed, inheritance-calculus is strictly more expressive
than the lazy $\lambda$-calculus in Felleisen's
sense~\cite{felleisen1991-expressive-power}
(Section~\ref{sec:asymmetry}).
Section~\ref{sec:expression-problem} established immunity
to nonextensibility in the sense of the
Expression Problem~\cite{wadler1998-expression-problem}.
The case study of Section~\ref{sec:case-study} showed that
ordinary arithmetic yields the relational semantics of logic
programming: cyclic inheritance hierarchies converge by
the same ascending chain condition that governs recursive
Datalog, and Appendix~\ref{app:datalog-encoding} gave a
systematic encoding.
Section~\ref{sec:practical-function-color} exhibited
function color
blindness~\cite{nystrom2015-function-color}.

Inheritance is not a mechanism for organizing functions.
Functions are syntactic sugar for inheritance.

\bibliographystyle{ACM-Reference-Format}
\bibliography{references}

\appendix

\section{Well-Definedness of the Semantic Functions}
\label{app:well-definedness}

The six semantic
equations~(\ref{eq:properties})--(\ref{eq:this}) are
first-order: their inputs and outputs are paths and sets of
paths, with no function spaces, closures, or higher-order
constructs.
This section establishes well-definedness by showing that
the six equations define a monotone operator whose least
fixed point is computed soundly by tabled
evaluation~\cite{tamaki1986-tabled-resolution}.

\begin{definition}[Dependency graph]\label{def:dependency-graph}
  Fix an AST with its primitive functions $\defines$ and
  $\inherits$.
  The \emph{dependency graph}~$G$ has as vertices all
  semantic-function applications
  ($\properties(p)$, $\supers(p)$, $\overrides(p)$,
    $\bases(p)$, $\resolve(\ldots)$,
  $\this(\ldots)$) that arise from the AST\@.
  There is a directed edge from vertex~$u$ to vertex~$v$
  whenever computing~$u$ via
  equations~(\ref{eq:properties})--(\ref{eq:this})
  requires the value of~$v$.
\end{definition}

\begin{definition}[Immediate consequence operator $T_P$]%
  \label{def:tp-operator}
  Let $\mathcal{Q}$ be the set of all
  semantic-function queries
  (vertices of~$G$).
  For each query $q \in \mathcal{Q}$, let
  $\mathcal{V}_q$ be the set of answer values of the
  appropriate shape for~$q$
  (paths, sets of paths, or sets of pairs of paths,
  depending on the equation defining~$q$).
  A \emph{query--answer map} is an element of the
  pointwise product
  $\prod_{q \in \mathcal{Q}} \mathcal{P}(\mathcal{V}_q)$;
  equivalently, it assigns to each query~$q$ a set
  $M(q) \subseteq \mathcal{V}_q$ of candidate answers.
  The \emph{immediate consequence
  operator}~\cite{vanemden1976-predicate-logic-semantics}
  $T_P$ maps a query--answer map~$M$ to a new
  query--answer map~$T_P(M)$ by evaluating, for each
  query~$q$, the right-hand side of the equation that
  defines~$q$, interpreting every sub-query~$q'$ by
  the current approximation~$M(q')$.
\end{definition}

\begin{lemma}[Monotonicity]\label{lem:monotonicity}
  The operator $T_P$ is monotone on the complete lattice
  $\Bigl(\prod_{q \in \mathcal{Q}} \mathcal{P}(\mathcal{V}_q),\; \subseteq\Bigr)$.
\end{lemma}

\begin{proof}
  Each equation's right-hand side is built from
  $\cup$, $\in$, equality tests, and positive set
  comprehension $\{\, x \mid x \in S, \ldots \,\}$.
  None of the six equations uses negation, set
  difference, or complement.
  Therefore, if $M \subseteq M'$ (pointwise), then
  $T_P(M) \subseteq T_P(M')$.
\end{proof}

\begin{theorem}[Existence and uniqueness of
  $\mathrm{lfp}(T_P)$]\label{thm:lfp-exists}
  The operator $T_P$ has a unique least fixed point
  $\mathrm{lfp}(T_P)$.
\end{theorem}

\begin{proof}
  Each powerset lattice
  $(\mathcal{P}(\mathcal{V}_q), \subseteq)$
  is complete,
  so the pointwise product lattice
  $\Bigl(\prod_{q \in \mathcal{Q}} \mathcal{P}(\mathcal{V}_q), \subseteq\Bigr)$
  is also complete.
  By the Knaster--Tarski
  theorem~\cite{tarski1955-lattice-fixpoint-theorem},
  every monotone function on a complete lattice has a
  unique least fixed point.
\end{proof}

\begin{theorem}[Well-definedness]\label{thm:well-defined}
  The semantic functions
  $\properties$, $\supers$, $\overrides$, $\bases$,
  $\resolve$, $\this$ are well-defined:
  their denotational semantics is
  $\mathrm{lfp}(T_P)$, the unique least fixed point of
  the immediate consequence operator.
\end{theorem}

\begin{theorem}[Tabling soundness]%
  \label{thm:tabling-soundness}
  Tabled evaluation of
  equations~(\ref{eq:properties})--(\ref{eq:this})
  never returns a wrong answer, even when the
  dependency graph~$G$ is infinite.
  Specifically:
  \begin{enumerate}
    \item \textbf{Soundness (no false positives).}
      Every fact returned by tabling was produced by
      evaluating some equation's right-hand side on its
      inputs.
      By induction on the evaluation trace, every
      returned fact belongs to $\mathrm{lfp}(T_P)$.

    \item \textbf{Monotone accumulation.}
      The tabling accumulator only grows between
      iterations via set union; no fact is ever
      removed.

    \item \textbf{Correct fixpoint detection.}
      If no sub-query was re-entered during evaluation,
      all sub-queries were fully computed with no
      approximation, and the result is exact.
      If re-entry occurred but the accumulator is
      stable, meaning no new facts were added, the
      accumulator equals
      $\mathrm{lfp}(T_P)$ restricted to the reachable
      queries, because it is a pre-fixed point reached
      by ascending from~$\bot$.

    \item \textbf{Sound failure.}
      When tabling cannot determine the answer,
      whether because the iteration budget is exhausted
      or because of an infinite non-cyclic chain,
      it raises an error rather than
      returning a wrong answer.
  \end{enumerate}
\end{theorem}

\begin{proof}
  For~(1), each returned fact is the output of some
  equation's right-hand side applied to previously
  returned facts; by induction, all inputs are in
  $\mathrm{lfp}(T_P)$, so the output is too (by
  monotonicity and the fixed-point property).

  For~(2), the implementation accumulates via
  $A \leftarrow A \cup T_P(A)$, which is monotone in~$A$.

  For~(3), when no reentry occurs, the evaluation is
  ordinary memoised recursion on a finite acyclic
  reachable subgraph, so every sub-query is computed
  exactly and no approximation is involved.
  When reentry occurs, let $R \subseteq \mathcal{Q}$ be
  the set of queries reachable from the initial query,
  and let $T_P^R$ denote the restriction of $T_P$ to
  maps on~$R$.
  If the accumulator is stable, that is,
  $A = A \cup T_P^R(A)$, then
  $T_P^R(A) \subseteq A$, so $A$ is a pre-fixed point
  of~$T_P^R$.

  By~(1), every fact ever added to the accumulator was
  produced by some right-hand side evaluation and
  therefore belongs to $\mathrm{lfp}(T_P^R)$.
  Hence $A \subseteq \mathrm{lfp}(T_P^R)$.

  Conversely, by the Knaster--Tarski characterization
  of the least fixed point as the least pre-fixed
  point, $\mathrm{lfp}(T_P^R)$ is contained in every
  pre-fixed point of~$T_P^R$, in particular in~$A$.
  Thus $\mathrm{lfp}(T_P^R) \subseteq A$.

  Therefore $A = \mathrm{lfp}(T_P^R)$, that is, the stable
  accumulator equals the exact denotation on the
  reachable queries.

  For~(4), budget exhaustion and stack overflow both
  produce exceptions, never silent wrong answers.
\end{proof}

\paragraph{Termination}
Whether evaluation terminates depends on the
program:
\begin{itemize}
  \item \textbf{Acyclic, finite reachable subgraph.}
    One pass suffices; evaluation terminates.
  \item \textbf{Cyclic, finite per-query domain.}
    Multiple passes converge by the ascending chain
    condition on a finite
    lattice~\cite{bancilhon1986-recursive-query-strategies}.
    This is the Datalog case
    (Section~\ref{sec:recursive-rules}).
  \item \textbf{Infinite reachable subgraph.}
    Evaluation may diverge; this is inherent to
    Turing completeness and not a defect of the
    evaluation strategy.
\end{itemize}

\section{B\"ohm Tree Correspondence: Proofs}
\label{app:bohm-tree-proofs}

This appendix proves two results about the sublanguage of
inheritance-calculus corresponding to the $\lambda$-calculus
(Section~\ref{sec:bohm-tree}).
First, the first-iteration approximation
$T_P\!\uparrow\!1$ is fully abstract with respect to
B\"ohm tree equivalence
(Theorems~\ref{thm:adequacy} and~\ref{thm:first-order-semantics}),
establishing that the first iteration coincides with
the lazy $\lambda$-calculus.
Second, the full $\mathrm{lfp}(T_P)$ semantics,
the fixpoint $\lambda$-calculus, is strictly more defined:
$\lambda$-terms that diverge under $\beta$-reduction
may converge (Remark~\ref{rem:convergence-lfp}),
establishing a convergence hierarchy among the eager,
lazy, and fixpoint $\lambda$-calculi.

\subsection{B\"ohm Trees}

We briefly recall the definition of B\"ohm
trees~\cite{barendregt1984-lambda-calculus}.
A \emph{head reduction} $M \to_h M'$ contracts
the outermost $\beta$-redex only: if $M$ has the form
$(\lambda x.\, e)\; v\; M_1 \cdots M_k$, then
$M \to_h e[v/x]\; M_1 \cdots M_k$.
A term $M$ is in \emph{head normal form} (HNF) if it has no
head redex, that is, $M = \lambda x_1 \ldots x_m.\;
y\; M_1 \cdots M_k$ where $y$ is a variable.

The \emph{B\"ohm tree} $\mathrm{BT}(M)$ of a $\lambda$-term $M$
is an infinite labeled tree defined by:
\[
  \mathrm{BT}(M) =
  \begin{cases}
    \bot & \text{if } M \text{ has no HNF} \\[6pt]
    \lambda x_1 \ldots x_m.\;
    y\bigl(\mathrm{BT}(M_1),\; \ldots,\; \mathrm{BT}(M_k)\bigr)
    &
    \begin{aligned}[t]
      &\text{if } M \to^*_h \\[-2pt]
      &\lambda x_1 \ldots x_m.\; y\; M_1 \cdots M_k
    \end{aligned}
  \end{cases}
\]
B\"ohm tree equivalence $\mathrm{BT}(M) = \mathrm{BT}(N)$ is the
standard observational equivalence of the lazy
$\lambda$-calculus~\cite{abramsky1993-full-abstraction-lazy-lambda,abramsky1995-games-full-abstraction-lazy-lambda}.

\subsection{Path Encoding}

We define a correspondence between positions in the B\"ohm tree
and paths in the mixin tree.
When the translation $\mathcal{T}$ is applied to an ANF
$\lambda$-term, the resulting mixin tree has a specific shape:
\begin{itemize}
  \item An abstraction $\lambda x.\, M$ translates to a record
    with own properties $\{\mathrm{argument}, \mathrm{result}\}$,
    which is referred to as the \emph{abstraction shape}.
  \item A $\mathbf{let}$-binding
    $\mathbf{let}\; x = V_1\; V_2 \;\mathbf{in}\; M$ translates
    to a record with own properties $\{x, \mathrm{result}\}$,
    where $x$ holds the encapsulated application
    $\{\mathcal{T}(V_1),\; \mathrm{argument} \mapsto
    \mathcal{T}(V_2)\}$.
  \item A tail call $V_1\; V_2$ translates to a record with own
    properties $\{\mathrm{tailCall}, \mathrm{result}\}$, where
    $\mathrm{tailCall}$ holds the encapsulated application and
    $\mathrm{result}$ projects
    $\mathrm{tailCall}.\mathrm{result}$.
  \item A variable $x$ with de~Bruijn index $n$ translates to
    an indexed reference $\dbi{n}.\mathrm{argument}$.
\end{itemize}

\noindent
Each B\"ohm tree position is a sequence of navigation steps from
the root.
In the B\"ohm tree, the possible steps at a node
$\lambda x_1 \ldots x_m.\; y\; M_1 \cdots M_k$ are:
entering the body by peeling off one abstraction layer, and
entering the $i$-th argument $M_i$ of the head variable.
In the mixin tree, these correspond to following the
$\mathrm{result}$ label (for entering the body) and the
$\mathrm{argument}$ label after $\this$ resolution (for
entering an argument position).

Rather than comparing absolute paths across different mixin
trees, which would be sensitive to internal structure; hence, we
observe only the \emph{convergence} behavior accessible via
the $\mathrm{result}$ projection.
Recall the definition of inheritance-convergence
(Definition~\ref{def:convergence}).
Each $\mathrm{result}$-following step corresponds to one head
reduction step in the $\lambda$-calculus: a tail call's
$\mathrm{result}$ projects through the encapsulated
application, and a $\mathbf{let}$-binding's $\mathrm{result}$
enters the continuation.
The abstraction shape at the end signals that a head
normal form has been reached.

\subsection{Single-Path Lemma}

The $\this$ function (equation~\ref{eq:this}) is designed for the
general case of multi-target $\this$ resolution.
For translated $\lambda$-terms, we show it degenerates to
single-target resolution.

\begin{lemma}[Single-Path]\label{lem:single-path}
  For any closed ANF $\lambda$-term $M$, at every invocation of
  $\this(S,\; p_{\mathrm{def}},\; n)$ during evaluation of
  $\mathcal{T}(M)$, the frontier set $S$ contains exactly one
  path.
\end{lemma}

\begin{proof}
  By case analysis on the translation rules of~$\mathcal{T}$.
  Since $\mathcal{T}$ is compositional, the structural invariant
  established for each rule propagates to all inherited subtrees:
  every subtree appearing in the mixin tree is itself the image of
  some sub-term under~$\mathcal{T}$, so the invariant holds
  throughout the runtime composition.

  \paragraph{Case $\mathcal{T}(\lambda x.\, M)
    = \{\mathrm{argument} \mapsto \{\},\;
  \mathrm{result} \mapsto \mathcal{T}(M)\}$}
  This is a record literal with
  $\defines = \{\mathrm{argument}, \mathrm{result}\}$ and
  no inheritance sources, so $\inherits = \varnothing$.
  At the root path $p$, we have $\overrides(p) = \{p\}$ (a single
  path) since there are no inheritance sources to introduce
  additional branches.
  Inside $\mathcal{T}(M)$, any reference to $x$ is
  $\dbi{n}.\mathrm{argument}$ where $n$ is the de~Bruijn index
  pointing to this scope level.
  Since $\mathcal{T}(\lambda x.\, M)$ introduces exactly one
  scope level, there is exactly one
  mixin at that level, hence $\this$ finds exactly one matching
  pair $(p_{\mathrm{site}},\; p_{\mathrm{override}})$ in $\supers$.

  \paragraph{Case
    $\mathcal{T}(\mathbf{let}\; x = V_1\; V_2
    \;\mathbf{in}\; M)
    = \{x \mapsto \{\mathcal{T}(V_1),\;
      \mathrm{argument} \mapsto \mathcal{T}(V_2)\},\;
  \mathrm{result} \mapsto \mathcal{T}(M)\}$}
  The outer record has own properties $\{x, \mathrm{result}\}$
  and no inheritance sources at the outer level
  ($\inherits = \varnothing$), so $\overrides(\text{root})
  = \{\text{root}\}$.
  Inside the $x$ subtree, the inheritance
  $\{\mathcal{T}(V_1),\; \mathrm{argument} \mapsto
  \mathcal{T}(V_2)\}$ has exactly one inheritance source
  ($\mathcal{T}(V_1)$), creating exactly one $\bases$ entry.
  By induction on $V_1$, the single-path property holds inside
  $\mathcal{T}(V_1)$.
  Inside $\mathcal{T}(M)$, the induction hypothesis applies
  directly.

  \paragraph{Case
    $\mathcal{T}(\mathbf{let}\; x = V
    \;\mathbf{in}\; M)
    = \{x \mapsto \{\mathrm{result} \mapsto \mathcal{T}(V)\},\;
  \mathrm{result} \mapsto \mathcal{T}(M)\}$}
  The outer record has own properties $\{x, \mathrm{result}\}$
  and no inheritance sources at the outer level, so
  $\overrides(\text{root}) = \{\text{root}\}$.
  The $x$ subtree is itself a record literal with own
  property $\{\mathrm{result}\}$ and no inheritance
  sources.
  Thus neither entering the bound value nor entering the
  body introduces branching in $\supers$.
  By induction, the single-path property holds inside both
  $\mathcal{T}(V)$ and $\mathcal{T}(M)$.

  \paragraph{Case $\mathcal{T}(V_1\; V_2) =
    \{\mathrm{tailCall} \mapsto
      \{\mathcal{T}(V_1),\; \mathrm{argument} \mapsto
      \mathcal{T}(V_2)\},\;
      \mathrm{result} \mapsto
  \mathrm{tailCall}.\mathrm{result}\}$}
  Identical to the $\mathbf{let}$-binding case: the outer record
  has own properties $\{\mathrm{tailCall}, \mathrm{result}\}$
  with no inheritance sources at the outer level, and the
  $\mathrm{tailCall}$ subtree contains exactly one inheritance
  source.

  \paragraph{Case $\mathcal{T}(x)
  = \dbi{n}.\mathrm{argument}$}
  This is a reference, not a record.
  The $\this$ function walks up $n$ steps from the enclosing scope.
  By the compositionality argument above, every scope level
  encountered during the walk, including inherited subtrees
  reached through reference resolution, is the image of some
  sub-term under~$\mathcal{T}$, so each step of $\this$
  encounters exactly one matching pair in $\supers$.

  \medskip\noindent
  The key structural invariant is the following.
  For every subtree generated by~$\mathcal{T}$,
  its root has either no inheritance source or exactly one
  inheritance source.
  Moreover, the only roots with an inheritance source are the
  application wrappers introduced by the application
  $\mathbf{let}$-binding and tail-call rules, and in both cases
  that source is unique and encapsulated under a distinguished
  child ($x$ or $\mathrm{tailCall}$), never at the enclosing
  scope level.
  Therefore every scope level traversed by $\this$ contributes
  at most one matching pair in $\supers$, so the frontier set
  remains a singleton throughout evaluation.
\end{proof}

\subsection{Substitution Lemma}

The core mechanism of the translation is that inheritance with
$\mathrm{argument} \mapsto \mathcal{T}(V)$ plays the role of
substitution.  We make this precise.

\begin{lemma}[Substitution]\label{lem:substitution}
  Let $M$ be an ANF term in which $x$ may occur free,
  and let $V$ be a closed value.%
  \footnote{
    The proof uses the Single-Path Lemma
    (Lemma~\ref{lem:single-path}), which requires the inherited tree
    to occur within a closed term. This holds whenever
    $\lambda x.\, M$ applied to $V$ is a subterm of a closed ANF
    program.
  }
  Define the \emph{inherited tree}
  $C = \{\mathcal{T}(\lambda x.\, M),\;
  \mathrm{argument} \mapsto \mathcal{T}(V)\}$.
  Then for every path $p$ under $C \snoc \mathrm{result}$ and
  every label $\ell$:
  \[
    \ell \in \properties(p)
    \text{ in } C \snoc \mathrm{result}
    \quad\Longleftrightarrow\quad
    \ell \in \properties(p)
    \text{ in } \mathcal{T}(M[V/x])
  \]
  where $M[V/x]$ is the usual capture-avoiding substitution.
\end{lemma}

\begin{proof}
  By structural induction on $M$.

  \paragraph{Base case: $M = x$}
  Then $\mathcal{T}(x) =
  \dbi{n}.\mathrm{argument}$ (where $n$ is the de~Bruijn index
  of $x$), and
  $C \snoc \mathrm{result}$ contains this reference.
  In the inherited tree $C$, $\resolve$ and $\this$ resolve the reference by
  navigating from the reference's enclosing scope up $n$ steps
  to the binding $\lambda$, where the inheritance
  $\{\mathcal{T}(\lambda x.\, x),\;
  \mathrm{argument} \mapsto \mathcal{T}(V)\}$ inherits into the
  $\mathrm{argument}$ slot, providing $\mathcal{T}(V)$.
  Since $C$ is the $\mathrm{tailCall}$ subtree of
  $\mathcal{T}\bigl((\lambda x.\, M)\; V\bigr)$, which is a
  closed ANF term, Lemma~\ref{lem:single-path} applies and
  $\this$ finds exactly one path,
  so $\resolve$ returns $\mathcal{T}(V)$'s subtree.
  Since $M[V/x] = V$, we have
  $\mathcal{T}(M[V/x]) = \mathcal{T}(V)$, and the properties
  coincide.

  \paragraph{Base case: $M = y$ where $y \neq x$ and $y$ is
  $\lambda$-bound}
  Then $\mathcal{T}(y) = \dbi{m}.\mathrm{argument}$ where $m$
  is the de~Bruijn index of~$y$.
  This walks up to the scope that binds~$y$, which is different
  from the scope that binds~$x$, so the inheritance with
  $\mathrm{argument} \mapsto \mathcal{T}(V)$ at $x$'s binding
  scope has no effect.
  Since $M[V/x] = y$, the properties are identical.

  \paragraph{Base case: $M = y$ where $y \neq x$ and $y$ is
  let-bound}
  Then $\mathcal{T}(y) = y.\mathrm{result}$.
  This is a lexical reference to the sibling property~$y$ in the
  enclosing record, so the inherited tree~$C$ does not alter its
  resolution.
  Since $M[V/x] = y$, the properties are identical.

  \paragraph{Inductive case: $M = \lambda y.\, M'$}
  Then $\mathcal{T}(M) =
  \{\mathrm{argument} \mapsto \{\},\;
  \mathrm{result} \mapsto \mathcal{T}(M')\}$.
  The subtree at $C \snoc \mathrm{result}$ is another
  abstraction-shaped record.
  References to $x$ inside $\mathcal{T}(M')$ resolve through
  $\this$ in exactly the same way (with one additional scope
  level from the inner $\lambda$), and by induction on $M'$, the
  properties under
  $C \snoc \mathrm{result} \snoc \mathrm{result}$
  coincide with those of
  $\mathcal{T}(M'[V/x])$.
  Since
  $(\lambda y.\, M')[V/x] = \lambda y.\, (M'[V/x])$
  (assuming $y$ is fresh), the result follows.

  \paragraph{Inductive case:
    $M = \mathbf{let}\; z = V_1\; V_2$
  $\mathbf{in}\; M'$}
  Then
  \[
    \mathcal{T}(M) =
    \{z \mapsto \{\mathcal{T}(V_1),\;
      \mathrm{argument} \mapsto \mathcal{T}(V_2)\},\;
    \mathrm{result} \mapsto \mathcal{T}(M')\}.
  \]
  The substitution distributes:
  $M[V/x] =
  \mathbf{let}\; z = V_1[V/x]\; V_2[V/x]
  \;\mathbf{in}\; M'[V/x]$.
  By induction on $V_1$, $V_2$, and $M'$, the properties under
  each subtree ($z$ and $\mathrm{result}$) coincide,
  since references to $x$ inside each subtree resolve to
  $\mathcal{T}(V)$ via the same $\this$ mechanism.

  \paragraph{Inductive case:
    $M = \mathbf{let}\; z = V'
  \;\mathbf{in}\; M'$}
  Then
  \[
    \mathcal{T}(M) =
    \{z \mapsto \{\mathrm{result} \mapsto \mathcal{T}(V')\},\;
    \mathrm{result} \mapsto \mathcal{T}(M')\}.
  \]
  The substitution distributes:
  $M[V/x] =
  \mathbf{let}\; z = V'[V/x]
  \;\mathbf{in}\; M'[V/x]$.
  By induction on $V'$ and $M'$, the properties under both
  subtrees, specifically $z$ and $\mathrm{result}$, coincide.
  In particular, occurrences of~$z$ in $M'$ are translated as the
  lexical reference $z.\mathrm{result}$ on both sides, so the
  wrapper introduced by the value $\mathbf{let}$-binding is
  preserved by the substitution.

  \paragraph{Inductive case: $M = V_1\; V_2$ (tail call)}
  Analogous to the $\mathbf{let}$-binding case with
  $\mathrm{tailCall}$ in place of $z$.
\end{proof}

\subsection{Convergence Preservation}

The convergence criterion (Definition~\ref{def:convergence})
follows the $\mathrm{result}$ chain from the root.
Each step in this chain corresponds to one head reduction step
in the $\lambda$-calculus.
We make this precise.

\begin{lemma}[Result-step]\label{lem:result-step}
  Let $(\lambda x.\, M)\; V$ be a tail call, which is a $\beta$-redex.
  Then for every path $p$ and label $\ell$:
  \[
    \ell \in \properties\!\bigl(\,
    \text{root} \snoc \mathrm{result} \snoc p\,\bigr)
    \text{ in } \mathcal{T}\bigl((\lambda x.\, M)\; V\bigr)
    \;\Longleftrightarrow\;
    \ell \in \properties\!\bigl(\,
    \text{root} \snoc p\,\bigr)
    \text{ in } \mathcal{T}(M[V/x])
  \]
  That is, following one $\mathrm{result}$ projection in the
  tail call's mixin tree yields the same properties as the
  root of the reduct's mixin tree.
\end{lemma}

\begin{proof}
  The translation gives:
  \[
    \mathcal{T}\bigl((\lambda x.\, M)\; V\bigr) =
    \{\mathrm{tailCall} \mapsto
      \{\mathcal{T}(\lambda x.\, M),\;
      \mathrm{argument} \mapsto \mathcal{T}(V)\},\;
      \mathrm{result} \mapsto
    \mathrm{tailCall}.\mathrm{result}\}
  \]
  The $\mathrm{result}$ label at the root is defined as
  $\mathrm{tailCall}.\mathrm{result}$, so
  $\text{root} \snoc \mathrm{result}$ resolves to the
  $\mathrm{result}$ path inside the inheritance
  $\mathrm{tailCall} = \{\mathcal{T}(\lambda x.\, M),\;
  \mathrm{argument} \mapsto \mathcal{T}(V)\}$.
  This inheritance is exactly the inherited tree $C$ of
  Lemma~\ref{lem:substitution}, and
  $\mathrm{tailCall} \snoc \mathrm{result}$ is
  $C \snoc \mathrm{result}$.
  By Lemma~\ref{lem:substitution}, the properties under
  $C \snoc \mathrm{result}$ coincide with those of
  $\mathcal{T}(M[V/x])$.

  \medskip\noindent
  \emph{Let-binding variant.}
  An analogous result holds for
  $\mathbf{let}\; x = V_1\; V_2 \;\mathbf{in}\; M'$
  where $V_1 = \lambda y.\, M''$.
  The translation gives:
  \[
    \mathcal{T}\bigl(\mathbf{let}\; x = V_1\; V_2
    \;\mathbf{in}\; M'\bigr) =
    \{x \mapsto \{\mathcal{T}(V_1),\;
      \mathrm{argument} \mapsto \mathcal{T}(V_2)\},\;
    \mathrm{result} \mapsto \mathcal{T}(M')\}
  \]
  Since $\mathrm{result} \mapsto \mathcal{T}(M')$ is a property
  definition, the subtree at
  $\text{root} \snoc \mathrm{result}$ is $\mathcal{T}(M')$.
  Inside $\mathcal{T}(M')$, each reference to $x$ is
  $x.\mathrm{result}$, which resolves to the $\mathrm{result}$
  path inside the $x$ subtree
  $\{\mathcal{T}(\lambda y.\, M''),\;
  \mathrm{argument} \mapsto \mathcal{T}(V_2)\}$.
  This is the inherited tree $C$ of
  Lemma~\ref{lem:substitution}, so
  $x.\mathrm{result}$ has the same properties as
  $\mathcal{T}(M''[V_2/y])$.
  In the $\lambda$-calculus,
  $M = (\lambda x.\, M')(V_1\; V_2)$, and the head reduct is
  $N = M'[(V_1\; V_2)/x]$.
  In $\mathcal{T}(N)$, each occurrence of $x$ is replaced by
  $(V_1\; V_2)$, whose result, by the tail-call variant above,
  also yields $\mathcal{T}(M''[V_2/y])$.
  Therefore the properties at
  $\text{root} \snoc \mathrm{result} \snoc p$ in
  $\mathcal{T}(M)$ coincide with those at
  $\text{root} \snoc p$ in $\mathcal{T}(N)$ for all $p$.

  \medskip\noindent
  \emph{Value let-binding variant.}
  An analogous result holds for
  $\mathbf{let}\; x = V \;\mathbf{in}\; M'$.
  The translation gives:
  \[
    \mathcal{T}\bigl(\mathbf{let}\; x = V \;\mathbf{in}\; M'\bigr) =
    \{x \mapsto \{\mathrm{result} \mapsto \mathcal{T}(V)\},\;
    \mathrm{result} \mapsto \mathcal{T}(M')\}.
  \]
  Since $\mathrm{result} \mapsto \mathcal{T}(M')$ is a property
  definition, the subtree at
  $\text{root} \snoc \mathrm{result}$ is $\mathcal{T}(M')$.
  Inside $\mathcal{T}(M')$, each occurrence of $x$ is translated
  as the lexical reference $x.\mathrm{result}$, which resolves to
  the subtree $\mathcal{T}(V)$.
  This is exactly the translation of the reduct
  $M'[V/x]$.
  Therefore the properties at
  $\text{root} \snoc \mathrm{result} \snoc p$ in
  $\mathcal{T}(\mathbf{let}\; x = V \;\mathbf{in}\; M')$
  coincide with those at
  $\text{root} \snoc p$ in $\mathcal{T}(M'[V/x])$ for all $p$.
\end{proof}

\begin{theorem}[Convergence preservation]%
  \label{thm:convergence-preservation}
  If $M \to_h N$ (head reduction), then
  $\mathcal{T}(M){\Downarrow}$ if and only if
  $\mathcal{T}(N){\Downarrow}$.
\end{theorem}

\begin{proof}
  A head reduction step in ANF takes one of three forms:
  a tail call $(\lambda x.\, M')\; V \to_h M'[V/x]$,
  an application $\mathbf{let}$-binding
  $\mathbf{let}\; x = V_1\; V_2 \;\mathbf{in}\; M'
  \to_h M'[(V_1\; V_2)/x]$,
  or a value $\mathbf{let}$-binding
  $\mathbf{let}\; x = V \;\mathbf{in}\; M'
  \to_h M'[V/x]$.
  The second and third forms are simply the ANF
  presentation of one $\beta$-step in the underlying
  $\lambda$-term.
  In all three cases, Lemma~\ref{lem:result-step}
  (tail-call, application let-binding, and value
  let-binding variants) shows that
  the properties at $\text{root} \snoc \mathrm{result}^n$ in
  $\mathcal{T}(M)$ coincide with the properties at
  $\text{root} \snoc \mathrm{result}^{n-1}$ in
  $\mathcal{T}(N)$ for all $n \ge 1$, where $N$ is the
  head reduct.
  Hence the abstraction shape appears at depth $n$ in the
  pre-reduct if and only if it appears at depth $n - 1$ in
  the post-reduct.
\end{proof}

\begin{remark}[Convergence hierarchy]
  \label{rem:convergence-lfp}
  Section~\ref{sec:bohm-tree} showed that the B\"ohm tree
  correspondence holds at the first iteration
  $T_P\!\uparrow\!1$ and that $\mathrm{lfp}(T_P)$ is
  strictly more defined.
  Here we make the connection precise.
  Tabled evaluation~\cite{tamaki1986-tabled-resolution}
  computes $\mathrm{lfp}(T_P)$ by iterating:
  $T_P\!\uparrow\!0 = \bot$,
  $T_P\!\uparrow\!1 = T_P(\bot)$,
  $T_P\!\uparrow\!2 = T_P(T_P(\bot))$,
  \ldots\ until stabilisation
  (Theorem~\ref{thm:tabling-soundness}).

  \begin{description}
    \item[$T_P\!\uparrow\!1$ corresponds to
      $\beta$-reduction.]
      $T_P\!\uparrow\!1$ evaluates each equation once,
      with re-entrant calls returning~$\bot$
      (the value from $T_P\!\uparrow\!0$).
      The Result-Step Lemma
      (Lemma~\ref{lem:result-step}) establishes a
      structural isomorphism between the equation
      systems of $\mathcal{T}(M)$ and
      $\mathcal{T}(N)$, shifted by one
      $\mathrm{result}$ step.
      This isomorphism is purely structural and holds
      for any number of iterations.
      Theorems~\ref{thm:adequacy}
      and~\ref{thm:first-order-semantics}
      are proved at this level.

    \item[$\mathrm{lfp}(T_P)$ defines the fixpoint
      $\lambda$-calculus.]
      We call the sublanguage with $\mathrm{lfp}(T_P)$
      semantics the \emph{fixpoint $\lambda$-calculus};
      it shares the same syntax as the lazy $\lambda$-calculus
      but has strictly greater computational power:
      $M{\Downarrow}_\lambda$ implies
      $\mathcal{T}(M){\Downarrow}_{\mathrm{lfp}}$,
      but not conversely.
  \end{description}
\end{remark}

\subsection{Adequacy}

\begin{definition}[Inheritance-convergence]\label{def:convergence}
  A mixin tree $T$ \emph{converges}, written $T{\Downarrow}$,
  if there exists $n \ge 0$ such that:
  \[
    \{\mathrm{argument},\, \mathrm{result}\}
    \;\subseteq\;
    \properties\!\bigl(\,
      \underbrace{\text{root} \snoc \mathrm{result}
      \snoc \cdots \snoc \mathrm{result}}_{n}
    \,\bigr)
  \]
  in the semantics of
  Section~\ref{sec:mixin-trees}.
\end{definition}

\begin{theorem}[Adequacy]\label{thm:adequacy}
  A closed ANF $\lambda$-term $M$ has a head normal form if and
  only if $\mathcal{T}(M){\Downarrow}$.
\end{theorem}

A closed ANF term at the top level is either an abstraction
(a value) or a computation (an application
  $\mathbf{let}$-binding, a value $\mathbf{let}$-binding,
or a tail call).
An abstraction translates to a record whose root immediately
has the abstraction shape; a computation's root has the form
$\{x, \mathrm{result}\}$ or
$\{\mathrm{tailCall}, \mathrm{result}\}$, and one must follow
the $\mathrm{result}$ chain to find the eventual value.

\begin{proof}[Proof of Theorem~\ref{thm:adequacy}]
  ($\Rightarrow$)
  Suppose $M \to^*_h \lambda x.\, M'$ in $k$ head reduction
  steps.
  We show $\mathcal{T}(M)$ converges at depth $\le k$ by
  induction on $k$.

  \paragraph{Base case ($k = 0$)}
  $M$ is already an abstraction
  $\lambda x.\, M'$.
  Then $\mathcal{T}(M) =
  \{\mathrm{argument} \mapsto \{\},\;
  \mathrm{result} \mapsto \mathcal{T}(M')\}$,
  whose root has
  $\defines = \{\mathrm{argument}, \mathrm{result}\}$.
  These labels are directly in $\defines$ of the root, so
  the recursive evaluation returns them without further
  recursion, and $\mathcal{T}(M)$ converges at depth $n = 0$.

  \paragraph{Inductive step ($k \ge 1$)}
  $M$ is not an abstraction,
  so $M$ is either a tail call, an application
  $\mathbf{let}$-binding, or a value
  $\mathbf{let}$-binding.
  Each form corresponds to one head $\beta$-step in the
  underlying $\lambda$-term.
  Let $N$ be the head reduct, so
  $M \to_h N \to^{k-1}_h \lambda x.\, M'$.
  Note that $N$ is closed: head reduction substitutes a
  closed value into a term whose only free variable is the
  substitution target, preserving closedness.
  By Lemma~\ref{lem:result-step} (tail-call,
    application let-binding, or value let-binding
  variant), the properties at
  $\text{root} \snoc \mathrm{result} \snoc p$
  in $\mathcal{T}(M)$ coincide with those at
  $\text{root} \snoc p$ in $\mathcal{T}(N)$.
  By the induction hypothesis, $\mathcal{T}(N)$
  converges at depth $\le k - 1$, so $\mathcal{T}(M)$
  converges at depth $\le k$.

  \medskip\noindent
  ($\Leftarrow$)
  Suppose $\mathcal{T}(M){\Downarrow}$ at depth $n$.
  We show $M$ has a head normal form by induction on $n$.

  \paragraph{Base case ($n = 0$)}
  The root of $\mathcal{T}(M)$ has
  abstraction shape
  $\{\mathrm{argument}, \mathrm{result}\}$.
  Only the abstraction rule $\mathcal{T}(\lambda x.\, M')$
  produces a root with own property $\mathrm{argument}$.
  (For tail calls and $\mathbf{let}$-bindings, the root has
    $\defines = \{\mathrm{tailCall}, \mathrm{result}\}$
    or $\{x, \mathrm{result}\}$, and $\mathrm{argument}$ cannot
    appear via $\supers$ because
    $\inherits = \varnothing$ at the root, giving
  $\bases(\text{root}) = \varnothing$.)
  Therefore $M$ is an abstraction, hence already in head normal
  form.

  \paragraph{Inductive step ($n \ge 1$)}
  The abstraction shape
  appears at depth $n$ but not at depth $0$.
  By the base case argument, $M$ is not an abstraction, so at the
  head position of this closed top-level ANF term, $M$ is either
  a tail call, an application $\mathbf{let}$-binding, or a value
  $\mathbf{let}$-binding.
  In each case, the ANF grammar requires the function
  position to be a value~$V_1$
  ($V ::= x \mid \lambda x.\, M$).
  Since $M$ is closed, $V_1$ cannot be a free variable,
  so $V_1$ must be a $\lambda$-abstraction;
  the head form is therefore a $\beta$-redex.
  By Lemma~\ref{lem:result-step} (tail-call,
    application let-binding, or value let-binding
  variant), the properties at depth
  $n - 1$ in $\mathcal{T}(N)$ match those at depth $n$
  in $\mathcal{T}(M)$, where $N$ is the head reduct.
  So $\mathcal{T}(N)$ converges at depth $n - 1$.
  Since $M$ is closed and head reduction preserves closedness
  , which substitutes a closed value for the only free variable,
  $N$ is a closed ANF term.
  By the induction hypothesis, $N$ has a head normal
  form, and since $M \to_h N$, so does $M$.
\end{proof}

\subsection{First-Order Semantics of the Lazy \texorpdfstring{$\lambda$}{λ}-Calculus}

Adequacy relates a single term's head normal form to its
inheritance-convergence.
The following theorem lifts this to an equivalence between
terms, establishing that the translation $\mathcal{T}$
gives the lazy $\lambda$-calculus a first-order semantics
that preserves and reflects B\"ohm tree equivalence.

\begin{definition}[$\lambda$-contextual equivalence under $\mathcal{T}$]%
  \label{def:ctx-equiv}
  For closed $\lambda$-terms $M$ and $N$, write
  $\mathcal{T}(M) \approx_\lambda \mathcal{T}(N)$ if for every
  closing $\lambda$-calculus context $C[\cdot]$:
  \[
    \mathcal{T}(C[M]){\Downarrow}
    \;\Longleftrightarrow\;
    \mathcal{T}(C[N]){\Downarrow}
  \]
\end{definition}

\noindent
The subscript $\lambda$ emphasizes that the quantification ranges
over $\lambda$-calculus contexts, not over all inheritance-calculus contexts.
This definition observes only convergence\footnote{
  Convergence provides exactly a single bit of information.
}, but
quantifying over all $\lambda$-contexts makes it a fine-grained
equivalence: different contexts can supply different arguments
and project different results, probing every aspect of the
term's behavior within the $\lambda$-definable fragment.

\begin{theorem}[First-order semantics of the lazy $\lambda$-calculus]%
  \label{thm:first-order-semantics}
  For closed $\lambda$-terms $M$ and $N$:
  \[
    \mathcal{T}(M) \approx_\lambda \mathcal{T}(N)
    \quad\Longleftrightarrow\quad
    \mathrm{BT}(M) = \mathrm{BT}(N)
  \]
\end{theorem}

\begin{proof}[Proof of Theorem~\ref{thm:first-order-semantics}]
  The translation $\mathcal{T}$ is compositional: for every
  $\lambda$-calculus context $C[\cdot]$, the mixin tree
  $\mathcal{T}(C[M])$ is determined by the mixin tree
  $\mathcal{T}(M)$ and the translation of the context.
  This means $\mathcal{T}$ preserves the context structure
  needed for the argument.

  \medskip\noindent
  ($\Leftarrow$)
  Suppose $\mathrm{BT}(M) = \mathrm{BT}(N)$.
  B\"ohm tree equivalence is a congruence~\cite{barendregt1984-lambda-calculus},
  so for every context $C[\cdot]$,
  $\mathrm{BT}(C[M]) = \mathrm{BT}(C[N])$.
  In particular, $C[M]$ has a head normal form iff $C[N]$ does.
  By Theorem~\ref{thm:adequacy},
  $\mathcal{T}(C[M]){\Downarrow}$ iff
  $\mathcal{T}(C[N]){\Downarrow}$.
  Hence $\mathcal{T}(M) \approx_\lambda \mathcal{T}(N)$.

  \medskip\noindent
  ($\Rightarrow$)
  Suppose $\mathrm{BT}(M) \neq \mathrm{BT}(N)$.
  B\"ohm tree equivalence coincides with observational
  equivalence for the lazy
  $\lambda$-calculus~\cite{abramsky1993-full-abstraction-lazy-lambda,abramsky1995-games-full-abstraction-lazy-lambda}:
  there exists a context $C[\cdot]$ such that $C[M]$ converges
  and $C[N]$ diverges, or vice versa.
  By Theorem~\ref{thm:adequacy},
  $\mathcal{T}(C[M]){\Downarrow}$ and
  $\mathcal{T}(C[N]){\Uparrow}$.
  Hence $\mathcal{T}(M) \not\approx_\lambda \mathcal{T}(N)$.
\end{proof}

\noindent
The proof rests on two pillars: Adequacy
(Theorem~\ref{thm:adequacy}), which is our contribution, and
the classical result that B\"ohm tree equivalence equals
observational equivalence for the lazy
$\lambda$-calculus~\cite{abramsky1993-full-abstraction-lazy-lambda,abramsky1995-games-full-abstraction-lazy-lambda}.
Neither pillar requires defining or comparing
absolute paths inside mixin trees.

\section{Expressive Asymmetry: Proofs}
\label{app:expressiveness}

\begin{definition}[Macro-expressibility, after Felleisen~\cite{felleisen1991-expressive-power}]
  \label{def:macro-expressibility}
  A construct $C$ of language $\mathscr{L}_1$ is \emph{macro-expressible}
  in language $\mathscr{L}_0$ if there exists a translation $\mathcal{E}$
  such that for every program $P$ containing occurrences of $C$,
  the translation $\mathcal{E}(P)$ is obtained by replacing each
  occurrence of $C$ with an $\mathscr{L}_0$ expression that depends only
  on $C$'s subexpressions, leaving all other constructs of $P$
  unchanged.
\end{definition}

\begin{lemma}[Composition of macro-expressible translations]
  \label{lem:macro-composition}
  If $\mathcal{E}_1$ macro-expresses the constructs of
  $\mathscr{L}_1$ in $\mathscr{L}_0$ and
  $\mathcal{E}_2$ macro-expresses the constructs of
  $\mathscr{L}_2$ in $\mathscr{L}_1$, then the composition
  $\mathcal{E}_1 \circ \mathcal{E}_2$ macro-expresses the
  constructs of $\mathscr{L}_2$ in $\mathscr{L}_0$.
\end{lemma}

\begin{proof}
  Each construct~$F$ of~$\mathscr{L}_2$ has a fixed
  syntactic abstraction~$A_F^{(2)}$
  over~$\mathscr{L}_1$.
  Each $\mathscr{L}_1$-construct~$G$ appearing
  in~$A_F^{(2)}$ has a fixed
  syntactic abstraction~$A_G^{(1)}$
  over~$\mathscr{L}_0$.
  Substituting~$A_G^{(1)}$ for every~$G$
  in~$A_F^{(2)}$ yields a fixed syntactic
  abstraction over~$\mathscr{L}_0$ for~$F$,
  satisfying~E4.
\end{proof}

\begin{definition}[Language universe]
  \label{def:language-universe}
  Let $\mathscr{U}$ be the language whose constructors are
  the union of the constructors of inheritance-calculus,
  the constructors of the lazy $\lambda$-calculus
  , including $\lambda$-abstraction, application, and variable, and
  $\mathbf{let}$-binding.
  The semantics of $\mathscr{U}$ is given by translating
  every $\mathscr{U}$-program to inheritance-calculus in
  three steps:
  \begin{enumerate}
    \item Convert every maximal $\lambda$-calculus
      subexpression (that is, every subtree whose root
        and all internal nodes use only $\lambda$-calculus
      constructors) to A-normal
      form~\cite{flanagan1993-essence-compiling-continuations}.
    \item Apply the translation $\mathcal{T}$
      (Section~\ref{sec:forward-translation}) to every ANF
      $\lambda$-calculus subexpression.
    \item Evaluate the resulting inheritance-calculus
      program by the rules of
      Section~\ref{sec:mixin-trees}.
  \end{enumerate}
  When a $\mathscr{U}$-expression interleaves
  $\lambda$-calculus and inheritance-calculus constructors
  (e.g.\ a $\mathbf{let}$-binding whose bound value is an
  inheritance-calculus expression), the translation is
  applied compositionally from leaves to root: each
  $\lambda$-calculus construct is replaced by its
  inheritance-calculus equivalent via~$\mathcal{T}$,
  treating inheritance-calculus subexpressions as opaque
  values.
  Since each rule of~$\mathcal{T}$
  (Section~\ref{sec:forward-translation}) is a fixed
  syntactic abstraction depending only on the translated
  subexpressions, this bottom-up application is
  well-defined and yields a pure inheritance-calculus
  program for step~(3).

  A $\mathscr{U}$-program $P$ converges,
  $\mathrm{eval}_{\mathscr{U}}(P)$, iff its translation
  converges.
\end{definition}

\begin{remark}\label{rem:universe-properties}
  The lazy $\lambda$-calculus and inheritance-calculus are
  conservative restrictions of $\mathscr{U}$: each is
  obtained by removing the other's constructors.
  For a pure inheritance-calculus program, steps~(1)
  and~(2) are vacuous, so
  $\mathrm{eval}_{\mathscr{U}}$ agrees with the
  inheritance-calculus semantics.
  For a pure $\lambda$-calculus program, the composition
  of steps~(1)--(3) agrees with lazy evaluation by
  Adequacy (Theorem~\ref{thm:adequacy}).
\end{remark}

The three-step translation is effective (ANF conversion
and~$\mathcal{T}$ are syntactic transformations) and the
recursive evaluation of inheritance-calculus
(Section~\ref{sec:mixin-trees}) is a semi-decision
procedure, so $\mathrm{eval}_{\mathscr{U}}$ is
recursively enumerable, as required by
Felleisen's framework~\cite{felleisen1991-expressive-power}.

\begin{lemma}[$\mathbf{let}$-binding is macro-expressible
  in the lazy $\lambda$-calculus]
  \label{lem:let-macro}
  The translation
  $\mathbf{let}\; x = e_1 \;\mathbf{in}\; e_2
  \;\mapsto\; (\lambda x.\, e_2)\; e_1$
  is a macro-expressible elimination of $\mathbf{let}$
  in the lazy $\lambda$-calculus.
\end{lemma}

\begin{proof}
  The translation replaces each $\mathbf{let}$-binding
  with a $\lambda$-expression that depends only on the
  two subexpressions $e_1$ and~$e_2$.
  This is Proposition~3.7 of
  Felleisen~\cite{felleisen1991-expressive-power}.
\end{proof}

\begin{lemma}[Equal expressiveness of the lazy
  $\lambda$-calculus and its ANF fragment]
  \label{lem:anf-equiv}
  In $\mathscr{U}$, the lazy $\lambda$-calculus and the
  ANF lazy $\lambda$-calculus
  (Section~\ref{sec:forward-translation}) are equally
  macro-expressive.
\end{lemma}

\begin{proof}
  Every ANF term is a $\lambda$-term, so the embedding
  of the ANF fragment into the full $\lambda$-calculus is
  the identity, which is trivially macro-expressible.
  For the reverse direction, the ANF
  transformation~\cite{flanagan1993-essence-compiling-continuations} consists of
  two steps, each macro-expressible:
  \begin{enumerate}
    \item \emph{Application flattening.}
      Replace each nested application
      $(e_1\; e_2)\; e_3$ with
      $\mathbf{let}\; x = e_1\; e_2 \;\mathbf{in}\;
      x\; e_3$.
      The syntactic abstraction is
      $A(\cdot_1, \cdot_2, \cdot_3)
      = \mathbf{let}\; x = \cdot_1\; \cdot_2
      \;\mathbf{in}\; x\; \cdot_3$,
      depending only on the three subexpressions.
    \item \emph{Value lifting.}
      Replace each free value~$V$ used in an application
      position with
      $\mathbf{let}\; x = V \;\mathbf{in}\; x$.
      The syntactic abstraction is
      $A(\cdot_1) = \mathbf{let}\; x = \cdot_1
      \;\mathbf{in}\; x$.
  \end{enumerate}
  By Lemma~\ref{lem:let-macro}, each resulting
  $\mathbf{let}$-binding is itself
  macro-expressible in the $\lambda$-calculus.
  By Lemma~\ref{lem:macro-composition}, composing
  these macro-expressible translations
  yields a macro-expressible translation from the full
  $\lambda$-calculus to its ANF fragment.
\end{proof}

\begin{theorem}[Forward macro-expressibility]
  \label{thm:forward-macro}
  The translation $\mathcal{T}$
  (Section~\ref{sec:forward-translation}) is a macro-expressible
  embedding of the $\lambda$-calculus into inheritance-calculus.
\end{theorem}

\begin{proof}
  We verify condition~E4 of
  Felleisen~\cite[Definition~3.11]{felleisen1991-expressive-power}:
  for each $\lambda$-calculus construct~$F$ of arity~$a$,
  there must exist a fixed $a$-ary syntactic abstraction
  $A_F(\cdot_1,\ldots,\cdot_a)$ over
  inheritance-calculus such that
  $\mathcal{T}(F(e_1,\ldots,e_a)) =
  A_F(\mathcal{T}(e_1),\ldots,\mathcal{T}(e_a))$.
  We check each rule of $\mathcal{T}$
  (Section~\ref{sec:forward-translation}):
  \begin{enumerate}
    \item \emph{$\lambda$-bound variable}
      ($x$ with de~Bruijn index $n'$).
      This is a nullary construct and therefore has no subexpressions.
      The syntactic abstraction is the constant
      $A = \dbi{n'}.\mathrm{argument}$.
    \item \emph{Let-bound variable} ($x$).
      Nullary.
      $A = x.\mathrm{result}$.
    \item \emph{$\lambda$-abstraction}, denoted $\lambda x.\, M$.
      Unary in the body~$M$.
      $A(\cdot_1) = \{\mathrm{argument} \mapsto \{\},\;
      \mathrm{result} \mapsto \cdot_1\}$.
    \item \emph{Application let-binding}
      ($\mathbf{let}\; x = V_1\; V_2
      \;\mathbf{in}\; M$).
      Ternary in~$V_1$, $V_2$, and~$M$.
      $A(\cdot_1, \cdot_2, \cdot_3) =
      \{x \mapsto \{\cdot_1,\;
        \mathrm{argument} \mapsto \cdot_2\},\;
      \mathrm{result} \mapsto \cdot_3\}$.
    \item \emph{Value let-binding}
      ($\mathbf{let}\; x = V \;\mathbf{in}\; M$).
      Binary in~$V$ and~$M$.
      $A(\cdot_1, \cdot_2) =
      \{x \mapsto \{\mathrm{result} \mapsto \cdot_1\},\;
      \mathrm{result} \mapsto \cdot_2\}$.
    \item \emph{Tail call} ($V_1\; V_2$).
      Binary.\newline
      $A(\cdot_1, \cdot_2) =
      \{\mathrm{tailCall} \mapsto
        \{\cdot_1,\;
        \mathrm{argument} \mapsto \cdot_2\},\;
        \mathrm{result} \mapsto
      \mathrm{tailCall}.\mathrm{result}\}$.
  \end{enumerate}
  In each case the syntactic abstraction~$A_F$ is a fixed
  inheritance-calculus context whose holes are filled
  by the recursive translations of the subexpressions,
  satisfying~E4.
\end{proof}

\begin{theorem}[Nonexpressibility of inheritance]
  \label{thm:cartesian-nonexpressibility}
  The lazy $\lambda$-calculus cannot macro-express the
  inheritance constructors of $\mathscr{U}$
  (Definition~\ref{def:language-universe}).
\end{theorem}

\begin{proof}
  We apply Theorem~3.14(i) of
  Felleisen~\cite{felleisen1991-expressive-power} to the pair
  $L_0 = \text{lazy $\lambda$-calculus}$,
  $L_1 = \mathscr{U}$.
  By Definition~\ref{def:language-universe},
  $\mathscr{U} = L_0 + \{$mixin-tree constructors$\}$
  is a conservative extension of~$L_0$
  (Definition~3.2 of
  Felleisen~\cite{felleisen1991-expressive-power}).
  We verify the four conditions:
  \begin{enumerate}
    \item[(i)] \emph{$L_0 \subseteq L_1$:}
      The syntax of~$\mathscr{U}$ includes all
      $L_0$-phrases.
    \item[(ii)] \emph{No new $L_0$-phrases:}
      The mixin-tree constructors do not introduce new
      phrases in~$L_0$.
    \item[(iii)] \emph{Disjoint constructors:}
      The mixin-tree constructors are disjoint from the
      $L_0$-constructors.
    \item[(iv)] \emph{Semantics preserved:}
      For pure $\lambda$-terms, steps~(1) and~(2) of
      Definition~\ref{def:language-universe} are vacuous
      and step~(3) yields lazy evaluation by Adequacy
      (Theorem~\ref{thm:adequacy}), so
      $\mathrm{eval}_{\mathscr{U}}$ agrees with
      $\mathrm{eval}_{L_0}$ on $L_0$-phrases.
  \end{enumerate}
  The theorem requires us to exhibit two $L_0$-terms
  $M$ and $N$ with $M \cong_0 N$ but
  $M \not\cong_1 N$, where $\cong_0$ is operational
  equivalence in $L_0$ and $\cong_1$ is operational
  equivalence in $\mathscr{U}$.

  \paragraph{Operational equivalence in $L_0$}
  By B\"ohm tree
  equivalence~\cite{abramsky1993-full-abstraction-lazy-lambda,abramsky1995-games-full-abstraction-lazy-lambda}
  (Theorem~\ref{thm:first-order-semantics}), for closed
  $\lambda$-terms $M$ and $N$:
  $M \cong_0 N$ iff $\mathrm{BT}(M) = \mathrm{BT}(N)$.

  \paragraph{Separating pair}
  Define Church booleans and Church boolean equality:
  $\mathrm{true} = \lambda t.\,\lambda f.\, t$,
  $\mathrm{false} = \lambda t.\,\lambda f.\, f$, and
  $\mathrm{eq} = \lambda a.\,\lambda b.\,
  (a\;b)\;(b\;\mathrm{false}\;\mathrm{true})$.
  Let
  \[
    M = \mathrm{eq}\;\mathrm{false}\;\mathrm{false}
    \qquad
    N = \mathrm{eq}\;\mathrm{true}\;\mathrm{true}.
  \]
  Both reduce to Church~$\mathrm{true}$ in the
  lazy $\lambda$-calculus, so
  $\mathrm{BT}(M) = \mathrm{BT}(N)$ and hence
  $M \cong_0 N$.

  \paragraph{Distinguishing $\mathscr{U}$-context}
  We now construct a $\mathscr{U}$-context that
  separates $M$ and $N$.
  $M$ and $N$ are syntactically closed $\lambda$-terms;
  by Definition~\ref{def:language-universe}, their
  semantics in $\mathscr{U}$ is given by translating
  them to inheritance-calculus via ANF conversion followed
  by~$\mathcal{T}$.
  Under the extended ANF convention
  (Section~\ref{sec:forward-translation}), every free
  value used in an application is first bound to a name:
  \begin{align*}
    \mathrm{ANF}(M) ={} &
    \mathbf{let}\;\mathrm{eq} =
    \lambda a.\,\lambda b.\,
    \mathbf{let}\; x_1 = a\;b \;\mathbf{in}\; \\
    &\qquad
    \mathbf{let}\; f_1 = \mathrm{false}
    \;\mathbf{in}\;
    \mathbf{let}\; x_2 = b\;f_1 \;\mathbf{in}\; \\
    &\qquad
    \mathbf{let}\; t_1 = \mathrm{true}
    \;\mathbf{in}\;
    \mathbf{let}\; x_3 = x_2\;t_1
    \;\mathbf{in}\;
    x_1\;x_3 \\
    & \mathbf{in}\;\;
    \mathbf{let}\; f_1 = \mathrm{false}
    \;\mathbf{in}\;
    \mathbf{let}\;\mathrm{eqFalse} = \mathrm{eq}\;f_1 \\
    &\quad
    \;\mathbf{in}\;
    \mathbf{let}\; f_2 = \mathrm{false}
    \;\mathbf{in}\;
    \mathrm{eqFalse}\;f_2
  \end{align*}
  Applying~$\mathcal{T}$:
  \begin{align*}
    \mathcal{T}(\mathrm{ANF}(M)) \;=\; \{&
      \mathrm{eq} \mapsto \{\mathrm{result} \mapsto
      \mathcal{T}(\lambda a.\,\lambda b.\,\cdots)\}, \\
      & f_1 \mapsto \{\mathrm{result} \mapsto
      \mathcal{T}(\mathrm{false})\}, \\
      & \mathrm{eqFalse} \mapsto
      \{\mathrm{eq}.\mathrm{result},\;
        \mathrm{argument} \mapsto
      f_1.\mathrm{result}\}, \\
      & f_2 \mapsto \{\mathrm{result} \mapsto
      \mathcal{T}(\mathrm{false})\}, \\
      & \mathrm{tailCall} \mapsto
      \{\mathrm{eqFalse}.\mathrm{result},\;
        \mathrm{argument} \mapsto
      f_2.\mathrm{result}\}, \\
      & \mathrm{result} \mapsto
      \mathrm{tailCall}.\mathrm{result}
    \;\}
  \end{align*}
  and symmetrically for $N$ with $\mathrm{true}$ in place
  of $\mathrm{false}$.
  Note that $\mathrm{eq}$ is a named child
  of $\mathcal{T}(\mathrm{ANF}(M))$, bound via a value
  let-binding.
  Inside the translation of $\mathrm{eq}$, the inner
  $\lambda b$ scope contains the let-bindings
  for $x_1$, $f_1$, $x_2$, $t_1$, $x_3$, and a tail call.
  In particular, $\dbi{1}.\mathrm{argument}$ reaches
  the outer $\lambda a$ scope and
  $\dbi{0}.\mathrm{argument}$ reaches $\lambda b$
  itself.\footnote{
    These indices use the condensed
    convention, counting only $\lambda$-scopes.
    As noted in the translation footnote
    (Section~\ref{sec:forward-translation}),
    each let-binding and tail call introduces an
    additional scope level in inheritance-calculus;
    the formal index is correspondingly larger,
    but the target scope is the same.
  }

  The $\mathscr{U}$-context uses inheritance to add
  a $\mathrm{repr}$ record inside the $\lambda b$ scope
  of $\mathrm{eq}$ that projects the first operand.
  Define the context $C[\cdot]$ as:
  \begin{align*}
    C[\cdot] ={} & \mathbf{let}\;
    \mathrm{extended} = \{[\cdot],\;
      \mathrm{eq} \mapsto \{
        \mathrm{result} \mapsto \{ \\
          &\quad
          \mathrm{result} \mapsto \{
            \mathrm{repr} \mapsto \{
              \mathrm{firstOperand}
              \mapsto \dbi{2}.\mathrm{argument}
    \}\}\}\}\} \\
    & \mathbf{in}\;
    \mathrm{extended}.\mathrm{tailCall}.\mathrm{repr}.\mathrm{firstOperand}
  \end{align*}
  Here $[\cdot]$ is the hole.
  Note the two levels of $\mathrm{result}$ nesting in the
  mixin: the outer $\mathrm{result}$ enters the value
  let-binding wrapper for $\mathrm{eq}$, and the inner
  $\mathrm{result}$ enters the $\lambda a$ scope.
  The second $\mathrm{result}$ then contains the
  $\lambda b$ scope, so $\mathrm{repr}$ is placed at
  the $\lambda b$ scope level.
  The mixin $\{\mathrm{eq} \mapsto \{\mathrm{result}
      \mapsto \{\mathrm{result} \mapsto
  \{\mathrm{repr} \mapsto \cdots\}\}\}\}$ is an
  inheritance-calculus expression: it uses only the
  mixin-tree constructors of $\mathscr{U}$.
  The outer $\mathbf{let}$-binding and projection are
  $\lambda$-calculus constructs (Lemma~\ref{lem:let-macro}).
  Thus $C[\cdot]$ is a valid $\mathscr{U}$-context.

  The de~Bruijn index $\dbi{2}$ in
  $\mathrm{repr}$ counts scope levels
  (in the condensed convention):
  $\dbi{0}$ is $\mathrm{repr}$ itself,
  $\dbi{1}$ is the $\lambda b$ scope, and
  $\dbi{2}$ is the $\lambda a$ scope.
  So $\mathrm{firstOperand}$ projects
  $\lambda a$'s argument.

  Since $\mathrm{eqFalse}$ (respectively $\mathrm{eqTrue}$)
  inherits $\mathrm{eq}.\mathrm{result}$,
  and $\mathrm{tailCall}$ inherits
  $\mathrm{eqFalse}.\mathrm{result}$ (the $\lambda b$
  scope), the $\mathrm{repr}$ record placed by the mixin
  at the $\lambda b$ scope level propagates through
  the inheritance chain to $\mathrm{tailCall}$.
  In $C[M]$: the $\mathrm{eqFalse}$ binding supplies
  $\mathrm{argument} \mapsto f_1.\mathrm{result}$
  (that is, $\mathrm{false}$) to the $\lambda a$ scope, so
  $\mathrm{firstOperand}$ evaluates to
  $\mathrm{false} = \lambda t.\,\lambda f.\, f$.
  In $C[N]$: the $\mathrm{eqTrue}$ binding supplies
  $\mathrm{argument} \mapsto \mathrm{true}$, so
  $\mathrm{firstOperand}$ evaluates to
  $\mathrm{true} = \lambda t.\,\lambda f.\, t$.

  Define $D[\cdot] = (\lambda x.\, x\;\Omega\;I)\;
  C[\cdot]$.
  Then $\mathrm{eval}_{\mathscr{U}}$ holds for $D[M]$
  (since $\mathrm{false}\;\Omega\;I \to I$, which
  converges)
  but not for $D[N]$
  (since $\mathrm{true}\;\Omega\;I \to \Omega$, which
  diverges).
  Therefore $M \not\cong_1 N$.

  \paragraph{Conclusion}
  We have $M \cong_0 N$ but $M \not\cong_1 N$, so
  ${\cong_0} \neq {(\cong_1|_{L_0})}$.
  By Theorem~3.14(i) of
  Felleisen~\cite{felleisen1991-expressive-power},
  the lazy $\lambda$-calculus cannot macro-express the
  inheritance constructors of~$\mathscr{U}$.
\end{proof}

\begin{theorem}[Expressive asymmetry]
  \label{thm:expressive-asymmetry}
  In the language universe $\mathscr{U}$
  (Definition~\ref{def:language-universe}),
  inheritance-calculus is strictly more expressive than
  the lazy $\lambda$-calculus: it can macro-express every
  $\mathscr{U}$-construct that the $\lambda$-calculus
  can, but not conversely
  (Definition~3.17 of
  Felleisen~\cite{felleisen1991-expressive-power}).
\end{theorem}

\begin{proof}
  We must show:
  (a)~inheritance-calculus can macro-express every
  $\mathscr{U}$-construct that the lazy $\lambda$-calculus
  can macro-express, and
  (b)~the converse fails.

  \paragraph{Part (b): the lazy $\lambda$-calculus $\not\geq$
  inheritance-calculus}
  The $\lambda$-calculus contains its own constructors
  ($\lambda$, application, variable) and can macro-express
  $\mathbf{let}$-binding (Lemma~\ref{lem:let-macro}), but
  cannot macro-express the inheritance constructors of
  $\mathscr{U}$
  (Theorem~\ref{thm:cartesian-nonexpressibility}).
  Inheritance-calculus contains those constructors.
  Therefore the lazy $\lambda$-calculus is not at least
  as expressive as inheritance-calculus with respect
  to~$\mathscr{U}$.

  \paragraph{Part (a): inheritance-calculus $\geq$ the lazy
  $\lambda$-calculus}
  We show that inheritance-calculus can macro-express
  every $\mathscr{U}$-construct that the
  $\lambda$-calculus contains or can macro-express.
  The $\lambda$-calculus contains $\lambda$-abstraction,
  application, and variable; it can macro-express
  $\mathbf{let}$-binding
  (Lemma~\ref{lem:let-macro}).
  We must show that inheritance-calculus can
  macro-express all four of these constructs.
  The translation proceeds in two stages:
  \begin{enumerate}
    \item \emph{ANF conversion.}
      The A-normal form
      transformation~\cite{flanagan1993-essence-compiling-continuations}
      replaces nested applications with
      $\mathbf{let}$-bindings; it is macro-expressible
      in the $\lambda$-calculus by
      Lemma~\ref{lem:anf-equiv}.
      In the reverse direction,
      $\mathbf{let}$-bindings are macro-expressible in
      the $\lambda$-calculus by
      Proposition~3.7 of
      Felleisen~\cite{felleisen1991-expressive-power}
      (restated as Lemma~\ref{lem:let-macro}).
      Thus ANF conversion is a macro-expressible
      equivalence between the $\lambda$-calculus and its
      ANF fragment.
    \item \emph{The translation $\mathcal{T}$.}
      Each ANF $\lambda$-calculus construct maps to a
      fixed syntactic abstraction over
      inheritance-calculus
      (Theorem~\ref{thm:forward-macro}), satisfying
      condition~E4 of
      Felleisen~\cite[Definition~3.11]{felleisen1991-expressive-power}.
  \end{enumerate}
  By Lemma~\ref{lem:macro-composition}, composing the
  two macro-expressible translations gives a
  macro-expressible translation from
  the $\lambda$-calculus (with $\mathbf{let}$) into
  inheritance-calculus.
  Since inheritance-calculus trivially contains its own
  constructors, it can macro-express every
  $\mathscr{U}$-construct that the $\lambda$-calculus
  contains or can macro-express.
\end{proof}

\section{Multi-Target \texorpdfstring{$\this$}{this} Resolution in Other Systems}
\label{app:scala-multi-target}

This appendix demonstrates that two representative systems, Scala~3
and the NixOS module system, reject the multi-target $\this$ resolution
pattern that inheritance-calculus handles naturally.

\subsection*{Scala~3}

Consider two objects that independently extend an outer class,
each providing its own copy of an inner trait:

\begin{lstlisting}[style=scala]
class MyOuter:
  trait MyInner:
    def outer = MyOuter.this

object Object1 extends MyOuter
object Object2 extends MyOuter
object HasMultipleOuters extends Object1.MyInner
                     with Object2.MyInner
\end{lstlisting}

\noindent
Scala~3 rejects \texttt{HasMultipleOuters} with the error:

\begin{lstlisting}
trait MyInner is extended twice
object HasMultipleOuters cannot be instantiated since
  it has conflicting base types
  Object1.MyInner and Object2.MyInner
\end{lstlisting}

\noindent
The rejection is not a surface-level restriction but a consequence
of DOT's type-theoretic foundations~\cite{amin2016-dependent-object-types}.
In DOT, an object is constructed as $\nu(x : T)\, d$, where the
self variable~$x$ binds to a \emph{single} object.
\texttt{Object1.MyInner} and \texttt{Object2.MyInner} are distinct
path-dependent types, each constraining the outer $\this$
to a different object.
Their intersection requires \texttt{outer} to simultaneously
return \texttt{Object1} and \texttt{Object2}, but DOT's self
variable is single-valued, making this intersection unrealizable.

The equivalent inheritance-calculus definition is:
\begin{align*}
  \{ \quad \mathrm{MyOuter} &\mapsto \{
      \mathrm{MyInner} \mapsto \{
    \mathrm{outer} \mapsto \qualifiedthis{\mathrm{MyOuter}}\}\}, \\
    \mathrm{Object1} &\mapsto \{\mathrm{MyOuter}\}, \\
    \mathrm{Object2} &\mapsto \{\mathrm{MyOuter}\}, \\
    \mathrm{HasMultipleOuters} &\mapsto \{
      \mathrm{Object1}.\mathrm{MyInner},\;
    \mathrm{Object2}.\mathrm{MyInner}\}
  \quad \}
\end{align*}
This is well-defined in inheritance-calculus.
The reference
$\qualifiedthis{\mathrm{MyOuter}}$ inside $\mathrm{outer}$
has de~Bruijn index $n = 1$:
starting from $\mathrm{outer}$'s enclosing scope
$\mathrm{MyInner}$, one $\this$ step reaches
$\mathrm{MyOuter}$.
When $\mathrm{HasMultipleOuters}$ inherits from both
$\mathrm{Object1}.\mathrm{MyInner}$ and
$\mathrm{Object2}.\mathrm{MyInner}$,
the $\this$ function (equation~\ref{eq:this}) resolves
$\qualifiedthis{\mathrm{MyOuter}}$ by searching through
$\supers(\mathrm{HasMultipleOuters})$ for override paths matching
$\mathrm{MyInner}$'s definition site.
It finds two inheritance-site paths, one through
$\mathrm{Object1}$ and one through $\mathrm{Object2}$, and
returns both.
Both paths lead to records that inherit from the same
$\mathrm{MyOuter}$, so querying
$\mathrm{HasMultipleOuters}.\mathrm{outer}$ yields the
$\properties$ of $\mathrm{MyOuter}$ regardless of which
route is taken.
Because attribute resolution aggregates results via set union (which is an intrinsically idempotent operation), identical property definitions from the two routes merge without conflict. Thus, the calculus eliminates the need for any artificial route-priority or linearization rules.

\subsection*{NixOS Module System}

The NixOS module system rejects the same pattern with a static error.
The translation uses the module system's own abstractions:
\texttt{deferredModule} for traits, which are unevaluated module values imported
into other fixpoints, and \texttt{submoduleWith} for objects, which are evaluated
in their own fixpoints.

\begin{lstlisting}[style=nix]
let
  lib = (import <nixpkgs> { }).lib;
  result = lib.evalModules {
    modules = [
      (toplevel@{ config, ... }: {
        # class MyOuter { trait MyInner {
        #   def outer = MyOuter.this } }
        options.MyOuter = lib.mkOption {
          default = { };
          type = lib.types.deferredModuleWith {
            staticModules = [
              (MyOuter: {
                options.MyInner = lib.mkOption {
                  default = { };
                  type = lib.types.deferredModuleWith {
                    staticModules = [
                      (MyInner: {
                        options.outer = lib.mkOption {
                          default = { };
                          type =
                            lib.types.deferredModuleWith {
                              staticModules =
                                [ toplevel.config.MyOuter ];
                            };
                        };
                      })
                    ];
                  };
                };
              })
            ];
          };
        };
        # object Object1 extends MyOuter
        options.Object1 = lib.mkOption {
          default = { };
          type = lib.types.submoduleWith {
            modules = [ toplevel.config.MyOuter ];
          };
        };
        # object Object2 extends MyOuter
        options.Object2 = lib.mkOption {
          default = { };
          type = lib.types.submoduleWith {
            modules = [ toplevel.config.MyOuter ];
          };
        };
        # object HasMultipleOuters extends
        #   Object1.MyInner with Object2.MyInner
        options.HasMultipleOuters = lib.mkOption {
          default = { };
          type = lib.types.submoduleWith {
            modules = [
            toplevel.config.Object1.MyInner
            toplevel.config.Object2.MyInner
          ];
          };
        };
      })
    ];
  };
in
  builtins.attrNames result.config.HasMultipleOuters.outer
\end{lstlisting}

\noindent
The NixOS module system rejects this with:

\begin{lstlisting}
error: The option `HasMultipleOuters.outer'
  in `<unknown-file>'
  is already declared
  in `<unknown-file>'.
\end{lstlisting}

\noindent
The rejection occurs because
\texttt{Object1.\allowbreak{}My\-Inner} and
\texttt{Object2.\allowbreak{}My\-Inner} each carry the full
\texttt{static\-Modules} of the \texttt{My\-Inner} deferred
module, including the declaration
\texttt{options.\allowbreak{}outer}.
When \texttt{HasMultipleOuters} imports both, the module system's
\texttt{merge\-Option\-Decls} encounters two declarations of the
same option and raises a static error; it cannot express that
both declarations originate from the same trait and should be
unified.

The rejection is analogous to Scala's ``conflicting base types''
error above: both systems assume that each option or type member
has a single declaration site, and reject the multi-target situation
where two inheritance routes introduce the same declaration
independently.
In inheritance-calculus, the same pattern is well-defined because
$\overrides$ (equation~\ref{eq:overrides}) recognizes that both
routes lead to the same definition, and $\supers$
(equation~\ref{eq:supers}) collects both inheritance-site
contexts without duplication.

\section{First-Order Semantics of the ANF Fixpoint
\texorpdfstring{$\lambda$}{λ}-Calculus}
\label{app:lambda-fixpoint}

This appendix reformulates the semantics of
Section~\ref{sec:mixin-trees}, specialized to $\lambda$-terms via
the Single-Path Lemma (Lemma~\ref{lem:single-path}), directly on
ANF syntax. The first-iteration approximation gives the lazy
$\lambda$-calculus semantics.

The B\"ohm tree has exactly two path types---\emph{scope}
($\mathcal{S}$) and \emph{application}
($\mathcal{A}$)---with a unified $\mathtt{node}$ function
that maps paths to ADT values.
As in the main semantics (Section~\ref{sec:mixin-trees}),
all dynamic functions are set-valued, so that the
immediate consequence operator $T_P$
(Definition~\ref{def:tp-operator}) applies directly.
At $T_P\!\uparrow\!1$, the Single-Path Lemma
(Lemma~\ref{lem:single-path}) guarantees that each set
contains at most one element for translated
$\lambda$-terms; at $\mathrm{lfp}(T_P)$, re-entrant
calls that converge across iterations may introduce
additional calling contexts, so the sets can grow beyond
singletons.

The direct formulation requires eight semantic equations,
compared to six in Section~\ref{sec:mixin-trees}.
The additional equations reflect syntactic distinctions
of the $\lambda$-calculus that the three primitives of
inheritance-calculus unify away;
the proliferation is the cost of bypassing~$\mathcal{T}$.

\paragraph{Syntax}
Let $M$ denote a computation and $V$ a value.
\[
  \begin{array}{r@{\;::=\;}l}
    M & \mathbf{let}\; x = V_1\; V_2 \;\mathbf{in}\; M
    \mid \mathbf{let}\; x = V \;\mathbf{in}\; M
    \mid V_1\; V_2
    \mid V \\
    V & \dbi{n} \mid x \mid \lambda.\, M
  \end{array}
\]
The grammar includes application let-bindings
($\mathbf{let}\; x = V_1\; V_2$),
value let-bindings ($\mathbf{let}\; x = V$),
tail calls ($V_1\; V_2$), and tail values ($V$).
Variables use de~Bruijn indices: $\dbi{n}$ counts only
$\lambda$-abstractions between the reference and its binder,
following the standard convention.
Let-bound variables are referenced by their binding name $x$.

\paragraph{Two node types}
Each position in the B\"ohm tree is either a \emph{scope node}
or an \emph{application node}.
These correspond to the path types $\mathcal{S}$ and
$\mathcal{A}$ introduced above.

A \textbf{scope node} (path type~$\mathcal{S}$)
represents a body: a chain of let-bindings ending
in a tail expression.
Its $\mathtt{node}$ is either
$\mathrm{Abstraction}(B,\,t)$
(introduced by $\lambda.\,M$) or
$\mathrm{Body}(B,\,t)$
(a continuation); see \emph{AST} below.
Its edges are the names of \emph{application} let-bindings
$\{x_1, \ldots, x_k\}$; value let-bindings do not create
edges (see \emph{AST} below).
Navigation from a scope to a binding:
$s \snoc x_i$ gives an application path
$a \in \mathcal{A}$.

An \textbf{application node} (path type~$\mathcal{A}$)
represents a function application $V_1\; V_2$ bound by a
$\mathbf{let}$.
Its $\mathtt{node}$ is $\mathrm{App}(r_h,\, r_a)$,
where $r_h, r_a \in \mathrm{ArgRef} \cup
\mathrm{BindingRef} \cup \mathrm{InlineLambda}$.
In the evaluated B\"ohm tree, $\head$ and $\argument$
lazily resolve these references to sets of scope paths:
$\head(p) \subseteq \mathcal{S}$
and $\argument(p) \subseteq \mathcal{S}$.

The path types alternate strictly:
\[
  \mathcal{S} \xrightarrow{\;\snoc\, x_i\;}
  \mathcal{A} \xrightarrow{\;\head / \argument\;}
  \mathcal{S} \xrightarrow{\;\snoc\, y_j\;}
  \mathcal{A} \;\cdots
\]
There is no scope-to-scope or
application-to-application direct transition.

\paragraph{References}
A reference
$r \in \mathrm{ArgRef} \cup \mathrm{BindingRef}
\cup \mathrm{InlineLambda}$
encodes how to reach a target scope node.
The ADT has three variants:
\begin{itemize}
  \item $\mathrm{ArgRef}(n)$: go up $n$
    $\lambda$-scope levels, access the argument value.
  \item $\mathrm{BindingRef}(n,\, x)$: go up $n$
    $\lambda$-scope levels, access binding~$x$.
  \item $\mathrm{InlineLambda}(M)$: an inline
    $\lambda$-literal that immediately resolves to a
    scope node (no lookup needed).
\end{itemize}
Since $n$ counts only $\lambda$-abstractions (not
$\mathbf{let}$-bindings or tail calls), it matches the
standard de~Bruijn convention.
For example, in $\lambda.\,\lambda.\,\mathbf{let}\; r =
\dbi{1}\;\dbi{0}\;\mathbf{in}\;r$, the outer-$\lambda$
variable is $\mathrm{ArgRef}(1)$ and the
inner-$\lambda$ variable is $\mathrm{ArgRef}(0)$.

\paragraph{AST}
An ANF term is parsed into an AST\@.
The function $\mathtt{node}$ maps each path to a
pure-data ADT value extracted from the parsed term.
The ADT has three variants:
\begin{itemize}
  \item $\mathrm{Abstraction}(B,\, t)$:
    a $\lambda$-scope with a binder slot,
    where $B$ is the set of application let-binding names
    $\{x_1, \ldots, x_k\}$, and $t$ is the tail
    expression (defined below).
  \item $\mathrm{Body}(B,\, t)$:
    a continuation scope,
    where $B$ is the set of application let-binding names
    $\{x_1, \ldots, x_k\}$, and $t$ is the tail
    expression (defined below).
  \item $\mathrm{App}(r_h,\, r_a)$ where
    $r_h, r_a \in \mathrm{ArgRef} \cup
    \mathrm{BindingRef} \cup \mathrm{InlineLambda}$:
    a function application $V_1\; V_2$ bound by a
    $\mathbf{let}$, where $r_h$ encodes~$V_1$
    (the function being applied) and $r_a$ encodes~$V_2$
    (the argument).
\end{itemize}
For scope paths $s \in \mathcal{S}$,
$\mathtt{node}(s) \in
\mathrm{Abstraction} \cup \mathrm{Body}$;
for application paths $a \in \mathcal{A}$,
$\mathtt{node}(a) \in \mathrm{App}$.
In pattern matches where the variant tag is irrelevant,
we write $\mathtt{node}(s) = \_(B, t)$ to bind $B$
and~$t$ from either variant.

\subparagraph{Tail expression}
The tail expression
$t \in \mathrm{Ref} \cup \mathrm{TailCall}$
has two variants:
\begin{itemize}
  \item $\mathrm{Ref}(r)$ where
    $r \in \mathrm{ArgRef} \cup \mathrm{BindingRef}
    \cup \mathrm{InlineLambda}$:
    a tail value encoded as a reference.
    (Value-bound names are resolved by inlining
    and do not appear as references.)
  \item $\mathrm{TailCall}(r_1, r_2)$: a tail-position
    application, where
    $r_1, r_2 \in \mathrm{ArgRef} \cup
    \mathrm{BindingRef} \cup \mathrm{InlineLambda}$.
\end{itemize}

\subparagraph{Reference encoding}
The function $\mathrm{enc}$ maps a value $V$ to a
reference:
\[
  \mathrm{enc}(V) =
  \begin{cases}
    \mathrm{ArgRef}(m)
    & \text{if } V = \dbi{m} \\[4pt]
    \mathrm{BindingRef}(n,\, y)
    & \text{if } V = y \text{ (let-bound, $n$
    enclosing $\lambda$s between use and def.)} \\[4pt]
    \mathrm{InlineLambda}(M')
    & \text{if } V = \lambda.\, M'
  \end{cases}
\]

\subparagraph{Definition of $\mathtt{node}$}
The function $\mathtt{node}$ is defined by structural
induction on the ANF term.
The definition uses two mutually recursive procedures,
$\mathrm{scope}$ and $\mathrm{body}$, corresponding to
the two scope variants.

$\mathrm{scope}(\lambda.\,M',\; s)$ defines the
$\mathtt{node}$ at scope path~$s$ for a
$\lambda$-abstraction:
\[
  \mathrm{scope}(\lambda.\,M',\; s)
  \;:\quad
  \mathtt{node}(s) = \mathrm{Abstraction}(B,\, t)
  \quad\text{where } (B,\, t) = \mathrm{body}(M',\, s)
\]

$\mathrm{body}(M,\, s)$ processes a computation~$M$
within an existing scope~$s$, returning $(B,\, t)$
by cases on~$M$:
\begin{itemize}
  \item $M = \mathbf{let}\; x = V_1\; V_2
    \;\mathbf{in}\; M'$
    (application let-binding).\;
    Let $(B',\, t) = \mathrm{body}(M',\, s)$.
    Then $B = \{x\} \cup B'$, and
    $\mathtt{node}(s \snoc x) =
    \mathrm{App}(\mathrm{enc}(V_1),\,
    \mathrm{enc}(V_2))$.

  \item $M = \mathbf{let}\; x = V
    \;\mathbf{in}\; M'$
    (value let-binding).\;
    No path is created.
    Every reference to~$x$ in~$M'$ is replaced by
    $\mathrm{enc}(V)$, with the de~Bruijn index
    in $\mathrm{ArgRef}$ and $\mathrm{BindingRef}$
    incremented by one for each $\lambda$-abstraction
    between the substitution site and the binding site.
    Then $(B,\, t) = \mathrm{body}(M'[x := V],\, s)$.

  \item $M = V_1\; V_2$
    (tail call).\;
    $B = \varnothing$ and
    $t = \mathrm{TailCall}(\mathrm{enc}(V_1),\,
    \mathrm{enc}(V_2))$.

  \item $M = V$
    (tail value).\;
    $B = \varnothing$ and
    $t = \mathrm{Ref}(\mathrm{enc}(V))$.
\end{itemize}
When $\mathrm{enc}(V) = \mathrm{InlineLambda}(M')$,
a dedicated scope path $s_\lambda$ is introduced with
$\mathrm{scope}(\lambda.\,M',\; s_\lambda)$,
giving
$\mathtt{node}(s_\lambda) =
\mathrm{Abstraction}(B',\, t')$
where $(B',\, t') = \mathrm{body}(M',\, s_\lambda)$.
Each $\mathrm{InlineLambda}$ occurrence in the ANF
term determines a distinct~$s_\lambda$.

In every case, the recursion descends to a strict
subterm of the input; $\mathtt{node}$ is therefore
total on all paths reachable from the root.
The initial call is
$\mathrm{scope}(\lambda.\,M_0,\; \text{root})$
if the top-level term is $\lambda.\,M_0$, or
$\mathtt{node}(\text{root}) = \mathrm{Body}(B,\, t)$
with $(B,\, t) = \mathrm{body}(M_0,\, \text{root})$
otherwise.

\paragraph{Dynamic functions}
The following functions are computed via the fixpoint
equations.
Functions that directly access the AST are defined by
exhaustive pattern matching on $\mathtt{node}$;
the remaining functions are defined by set comprehension.
We write $\enclosingscope(s \snoc x) = s$.

\subparagraph{Result.}
$\result(p)$ performs one step of head reduction:
\begin{equation}\label{eq:lambda-result}
  \result(p) =
  \begin{cases}
    \resolveref(r,\; p)
    & \text{if } \mathtt{node}(p) = \_(\_,\, \mathrm{Ref}(r)) \\[4pt]
    \resolveref(r_1,\; p)
    & \text{if } \mathtt{node}(p) = \_(\_,\, \mathrm{TailCall}(r_1, r_2)) \\[4pt]
    \varnothing
    & \text{if } \mathtt{node}(p) \in \mathrm{App}
  \end{cases}
\end{equation}
The two scope cases are exhaustive by the tail-expression
ADT\@.
In the $\mathrm{Ref}$ case, $r$ is a reference;
$\resolveref$ resolves it to a set of scopes
(including the $\mathrm{InlineLambda}$ case).
In the $\mathrm{TailCall}$ case,
$r_1$ is the head reference
and $\resolveref$ resolves it to function-body scopes.
The tail call also stores the argument reference~$r_2$,
which is accessed by de~Bruijn resolution when the function
body references its argument (see \emph{Calling context} below).

%% visiblebindings is not referenced by other equations;
%% commented out to reduce clutter.
%% \subparagraph{Visible bindings.}
%% $\visiblebindings(p)$ collects all bindings along the
%% $\result$ chain:
%% \begin{equation}\label{eq:lambda-visible-bindings}
%%   \visiblebindings(p) =
%%   \begin{cases}
%%     B \;\cup\;
%%     \left\{\; b \;\left|\;
%%       \begin{aligned}
%%         &p' \in \result(p),\\
%%         &b \in \visiblebindings(p')
%%       \end{aligned}
%%     \right.\right\}
%%     & \text{if } \mathtt{node}(p) = \_(B,\, \_) \\[6pt]
%%     \varnothing
%%     & \text{if } \mathtt{node}(p) \in \mathrm{App}
%%   \end{cases}
%% \end{equation}

\subparagraph{Head.}
$\head(p)$ resolves the head reference of an application:
\begin{equation}\label{eq:lambda-head}
  \head(p) =
  \begin{cases}
    \resolveref(r_h,\; \enclosingscope(p))
    & \text{if } \mathtt{node}(p) = \mathrm{App}(r_h,\, \_) \\[4pt]
    \varnothing
    & \text{if } \mathtt{node}(p) \in
      \mathrm{Abstraction} \cup \mathrm{Body}
  \end{cases}
\end{equation}

\subparagraph{Argument.}
$\argument(p)$ resolves the argument reference:
\begin{equation}\label{eq:lambda-argument}
  \argument(p) =
  \begin{cases}
    \resolveref(r_a,\; \enclosingscope(p))
    & \text{if } \mathtt{node}(p) = \mathrm{App}(\_,\, r_a) \\[4pt]
    \varnothing
    & \text{if } \mathtt{node}(p) \in
      \mathrm{Abstraction} \cup \mathrm{Body}
  \end{cases}
\end{equation}

\paragraph{Calling context and resolution}

\subparagraph{Calling context.}
When a $\lambda$-abstraction's body is entered via
$\beta$-reduction, the resulting scope~$s$ acquires a
\emph{calling context} that records which call
introduced~$s$.
A $\lambda$-body scope~$s$ is entered in one of two ways:
via a let-binding application, or via a tail call.
Each mechanism contributes calling contexts.

\emph{Let-binding contexts.}\;
For a scope $s'$ with $\mathtt{node}(s') = \_(B,\, \_)$
and a binding name $x \in B$, the application
$s' \snoc x$ contributes:
\begin{equation}\label{eq:lambda-let-contexts}
  \letcontexts(s) =
  \left\{\; (s',\; s_a) \;\left|\;
    \begin{aligned}
      &\mathtt{node}(s') = \_(B,\, \_),\;
      x \in B,\\
      &s \in \head(s' \snoc x),\\
      &s_a \in \argument(s' \snoc x)
    \end{aligned}
  \right.\right\}
\end{equation}

\emph{Tail-call contexts.}\;
A scope $s'$ with
$\mathtt{node}(s') = \_(\_,\, \mathrm{TailCall}(r_1, r_2))$
and $s \in \resolveref(r_1,\, s')$ contributes:
\begin{equation}\label{eq:lambda-tail-contexts}
  \tailcontexts(s) =
  \left\{\; (s',\; s_a) \;\left|\;
    \begin{aligned}
      &\mathtt{node}(s') = \_(\_,\, \mathrm{TailCall}(r_1, r_2)),\\
      &s \in \resolveref(r_1,\, s'),\\
      &s_a \in \resolveref(r_2,\, s')
    \end{aligned}
  \right.\right\}
\end{equation}

The full set of calling contexts:
\begin{equation}\label{eq:lambda-contexts}
  \contexts(s) =
    \letcontexts(s) \;\cup\;
    \tailcontexts(s)
\end{equation}
The caller scope and caller argument are obtained by
projecting $(s',\; \_)$ and $(\_,\; s_a)$ from
$\contexts(s)$ in the equations that use them
($\scoperesolution$ and $\resolveref$ below).
For the root scope, $\contexts$ is empty
(the root has no caller).
At $T_P\!\uparrow\!1$, the Single-Path Lemma
(Lemma~\ref{lem:single-path}) guarantees that
$\contexts(s)$ contains at most one pair for
translated $\lambda$-terms.
At $\mathrm{lfp}(T_P)$, re-entrant calls may converge and
supply additional calling contexts to the same scope.

\subparagraph{Reference resolution.}
$\resolveref$ resolves any reference to a set of scope nodes:
\begin{equation}\label{eq:lambda-resolve-ref}
  \resolveref(r,\; s) =
  \begin{cases}
    \left\{\; s_a \;\left|\;
      \begin{aligned}
        &s' \in \scoperesolution(n,\, s),\\
        &(\_,\; s_a) \in \contexts(s')
      \end{aligned}
    \right.\right\}
    & \text{if } r = \mathrm{ArgRef}(n) \\[6pt]
    \left\{\; s_h \;\left|\;
      \begin{aligned}
        &s' \in \scoperesolution(n,\, s),\\
        &\mathtt{node}(s') = \_(B,\, \_),\;
        x \in B,\\
        &s_h \in \head(s' \snoc x)
      \end{aligned}
    \right.\right\}
    & \text{if } r = \mathrm{BindingRef}(n,\, x) \\[6pt]
    \{s_{\lambda}\}
    & \text{if } r = \mathrm{InlineLambda}(M)
  \end{cases}
\end{equation}
In the $\mathrm{InlineLambda}$ case,
$s_\lambda$ is the scope path determined by
$\mathrm{scope}(\lambda.\,M,\; s_\lambda)$
(see \emph{Definition of $\mathtt{node}$} above).
For $\mathrm{ArgRef}(n)$,
$\scoperesolution$ navigates up $n$ $\lambda$-scope levels,
yielding a set of scopes;
the caller argument is then extracted from $\contexts$
at each such scope.
For $\mathrm{BindingRef}(n,\, x)$,
$\scoperesolution$ navigates up $n$ $\lambda$-scope levels,
then $x$ is projected from each resulting scope and its head is
resolved.

\subparagraph{Scope resolution.}
$\scoperesolution$ navigates upward through the B\"ohm tree
to find the binding site of a de~Bruijn reference,
returning a set of scopes.
Only $\lambda$-abstractions count as scope levels;
application $\mathbf{let}$-bindings and tail calls do not.
\begin{equation}\label{eq:lambda-scope}
  \scoperesolution(n,\; s) =
  \begin{cases}
    \{s\} & \text{if } n = 0 \\[6pt]
    \left\{\; s'' \;\left|\;
      \begin{aligned}
        &(s',\; \_) \in \contexts(s),\\
        &s'' \in \scoperesolution(n - 1,\; s')
      \end{aligned}
    \right.\right\}
    & \text{if } n > 0
  \end{cases}
\end{equation}
Each recursive step moves from the current scope~$s$ to
each caller scope in $\contexts(s)$
and decrements~$n$.
Since $n$ is a nonnegative integer, the nesting depth of
$\scoperesolution$ is bounded by the de~Bruijn index;
well-definedness of the mutual recursion with $\contexts$
follows from the $T_P$ fixed-point construction
(Definition~\ref{def:tp-operator}), as for all dynamic
functions above.

\paragraph{Observational fixpoint semantics}
The definitions above constitute the complete semantics.
Computation arises from querying progressively deeper paths
in a lazily constructed tree: the observer drives computation
by choosing which path to inspect, and the mutually recursive
definitions unfold on demand.

Formally, the mutually recursive
equations~(\ref{eq:lambda-result})--(\ref{eq:lambda-contexts})
and~(\ref{eq:lambda-resolve-ref})--(\ref{eq:lambda-scope})
define set-valued functions recursively:
$p' \in \result(p)$ holds if and only if it can be
established by a finite chain of applications of these
equations.
Well-definedness has two parts.
The static part: $\mathtt{node}$ is total on all
reachable paths by structural induction on the ANF term,
as established above.
The dynamic part: each equation's right-hand side is
built from $\cup$, $\in$, equality, and positive set
comprehension, so the immediate consequence
operator~$T_P$ (Definition~\ref{def:tp-operator}) is
monotone on the complete lattice of query--answer maps,
and the Knaster--Tarski theorem gives the least fixed
point.
When no finite derivation exists for a given query
(as with $\Omega$), the fixed-point iteration does not
terminate.%
\footnote{
  Although $\Omega$ is a finite ANF term, the set of
  reachable paths is infinite: each head reduction step
  unfolds a new tail call, extending the $\result$ chain
  and introducing fresh application paths.
  The query set~$\mathcal{Q}$ is therefore infinite, so
  the ascending chain condition does not apply and the
  Kleene iteration $T_P\!\uparrow\!n$ need not stabilise
  at any finite~$n$.
}

\paragraph{Convergence and head normal form}
An ANF term $M$ \emph{converges}, written $M{\Downarrow}$,
if there exist scopes $s_0, s_1, \ldots, s_k$
with $s_0 = \text{root}$,
$s_{i+1} \in \result(s_i)$ for each $0 \le i < k$,
and $\mathtt{node}(s_k) \in \mathrm{Abstraction}$,
in the semantics defined by the equations above.
The term has been reduced through $k$ head reduction steps
(each tracked by one $\result$ step) to an abstraction
$\lambda.\,M'$.

The same equations support two evaluation regimes.
At $T_P\!\uparrow\!1$ (one iteration, re-entrant calls
return~$\varnothing$), the semantics coincides with the lazy
$\lambda$-calculus:
$M$ converges iff it has a head normal form
(Theorem~\ref{thm:adequacy}), and two terms are
equivalent iff they have the same
B\"ohm tree~\cite{barendregt1984-lambda-calculus}
(Theorem~\ref{thm:first-order-semantics}).
At $\mathrm{lfp}(T_P)$ (iterate to the least fixpoint),
the semantics defines the \emph{fixpoint $\lambda$-calculus}
(Remark~\ref{rem:convergence-lfp}):
re-entrant calls that returned
$\varnothing$ at $T_P\!\uparrow\!1$ may now converge and
contribute additional heads and arguments to the same scope,
so a single scope acquires multiple calling contexts:
syntactically single inheritance becomes semantically
multiple inheritance through value recursion.
A term that converges under the lazy $\lambda$-calculus
also converges under the fixpoint $\lambda$-calculus;
a term that diverges under the lazy $\lambda$-calculus
may, under the fixpoint $\lambda$-calculus, still diverge
(as $\Omega$ does), or converge to zero, one, or
multiple values.
Only the iteration depth changes; the equations themselves
are identical.

The B\"ohm tree equations above are self-contained.
For the proofs (Appendix~\ref{app:bohm-tree-proofs}),
we observe that each equation is the specialization of
the corresponding mixin tree equation
(Section~\ref{sec:mixin-trees}) to terms in the image
of the translation~$\mathcal{T}$%
\footnote{
  Concretely,
  $\result$ specializes $\properties$/$\bases$,
  $\head$/$\argument$ specialize $\resolve$,
  $\scoperesolution$ specializes $\this$,
  and $\contexts$ specializes the provenance tracking
  in $\supers$.
}
(Section~\ref{sec:forward-translation}).

\section{Encoding Datalog in Inheritance-Calculus}
\label{app:datalog-encoding}

Section~\ref{sec:emergent-phenomena} observed that
inheritance-calculus natively produces the relational semantics
of logic programming.
This appendix makes that observation precise by giving a
systematic encoding of Datalog into inheritance-calculus.
The encoding uses no extensions to the calculus; it is
expressible in the three primitives of Section~\ref{sec:syntax}.
All examples in this appendix have executable MIXINv2
implementations\selfcite{~\cite{mixin2025}}{~(supplementary
material)}.
There are three ingredients to keep separate.
First, relations are stored as tries, so tuple lookup is indexed
by prefixes rather than by scanning a flat set of facts.
Second, the control skeleton is written in the visitor-pattern
style already used in Section~\ref{sec:case-study}; from a
functional-programming viewpoint, this is the same role played by
Scott-encoded case analysis.
Third, inheritance union and tabled evaluation give the resulting
program its relational, least-fixed-point semantics.
The appendix proceeds in that order: data representation first,
then control flow, then recursion.

\subsection{Relations as Tries}

A binary relation over a finite domain $D = \{a, b, c, \ldots\}$
is represented as a \emph{trie}: a mixin tree whose paths
encode tuples.
The trie is built from three base scopes:
$\mathrm{Relation}$ (abstract base),
$\mathrm{Cons}$ (non-empty node), and $\mathrm{Nil}$ (terminator).
Each constant $d \in D$ inherits from $\mathrm{Cons}$ and carries
a $\mathrm{TailMap}$ that provides per-constant sub-tries:
\begin{align*}
  &\mathrm{Relation} \mapsto \{\},\quad
  \mathrm{Cons} \mapsto \{\mathrm{Relation},\;
  \mathrm{Tail} \mapsto \{\mathrm{Relation}\}\},\quad
  \mathrm{Nil} \mapsto \{\mathrm{Relation}\}\\[2pt]
  &d \mapsto \{\mathrm{Cons},\;
    \mathrm{Tail} \mapsto \{\},\;
  \mathrm{TailMap} \mapsto \{d \mapsto \{\mathrm{Tail}\}\}\}
\end{align*}
A fact $R(d_1, d_2)$ is encoded as a trie path of depth~2,
with a $\mathrm{Nil}$ terminator.
The per-constant sub-tries are accessed via
$\mathrm{TailMap}.d$:
\begin{align*}
  R(a, b) \;\leadsto\quad
  R \mapsto \{a,\; \mathrm{TailMap} \mapsto \{a \mapsto
  \{b,\; \mathrm{TailMap} \mapsto \{b \mapsto \{\mathrm{Nil}\}\}\}\}\}
\end{align*}
Multiple facts for the same relation are accumulated by
inheritance (trie union).
Adding $R(b, c)$:
\begin{align*}
  R \mapsto \{a,\; b,\;
    \mathrm{TailMap} \mapsto \{&
      a \mapsto \{b,\; \mathrm{TailMap} \mapsto \{b \mapsto \{\mathrm{Nil}\}\}\},\\
  &b \mapsto \{c,\; \mathrm{TailMap} \mapsto \{c \mapsto \{\mathrm{Nil}\}\}\}\}\}
\end{align*}
The point of the trie representation is not merely that it is a
tree-shaped encoding.
It also provides the indexing structure used by the subsequent
joins: once a constant $d$ has been selected, the encoding moves
directly to $\mathrm{TailMap}.d$ and continues from the matching
sub-trie.
This is why the appendix speaks about tries rather than about an
arbitrary encoding of finite sets of tuples.

\subsection{Infrastructure}

Each constant $d$ carries four groups of boilerplate
definitions used by the encoding.
These play a role analogous to derived type-class instances
in Haskell: each constant must provide a fixed set of
properties that the encoding's generic patterns rely on.
Inheritance-calculus has no built-in derivation mechanism,
so the boilerplate is written out manually; however, the
amount of boilerplate per constant is $O(1)$
and does not grow with the domain size $|D|$,
so adding a new constant
does not require modifying any existing scope, preserving
the expression-problem property.

% 我们可以通过引入类似ECMAScript的Symbol机制来消除这些重复定义，但这让语言更复杂了

\begin{description}
  \item[Acceptance]
    The visitor-pattern infrastructure from
    Section~\ref{sec:case-study}.
    Operationally, this is the object-oriented spelling of a
    case split on the current constant.
    Each constant $d$ defines a $\mathrm{VisitorMap}$
    containing a branch keyed by $d$, a $\mathrm{Visitor}$
    with a $\mathrm{Visited}$ slot for per-constant results,
    and an $\mathrm{Accepted}$ property that delegates to the
    matching branch's $\mathrm{Visited}$ value.
    This is the dispatch mechanism: given a scope that may
    contain multiple constants, $\mathrm{Acceptance}$ selects the
    branch corresponding to each constant present and
    collects the per-constant $\mathrm{Visited}$ results.

  \item[WithVisitorTail]
    Maps each constant $d$ to its sub-trie entry
    $\mathrm{TailMap}.d$, labeled as $\mathrm{VisitorTail}$.
    This provides the right-hand side of a join\footnote{
      In the context of the visitor pattern, this acts as the visitor.
    }: when two relations' $\mathrm{VisitorMap}$ entries are
    composed, $\mathrm{VisitorTail}$ carries the joined
    relation's row data for the matched constant.

  \item[WithVisiteeTail]
    Maps each constant $d$ to its sub-trie entry
    $\mathrm{TailMap}.d$, labeled as $\mathrm{VisiteeTail}$.
    This provides the left-hand side\footnote{
      This acts as the visitee.
    }: the
    relation being iterated. Together,
    $\mathrm{WithVisitorTail}$ and $\mathrm{WithVisiteeTail}$
    enable value-level joins where only constants present
    in \emph{both} relations produce output.

  \item[Replacement]
    Given a $\mathrm{NewTail}$ value, produces a
    $\mathrm{Replaced}$ scope that inherits the label $d$
    but substitutes $\mathrm{TailMap}.d$ with
    $\mathrm{NewTail}$.
    This is the output construction mechanism: it
    builds output tuples by substituting new sub-tries into
    constant wrappers.
    Crucially, $\mathrm{Replacement}$ is defined on the
    per-constant $\mathrm{TailMap}$ entry (not on the merged
    scope), ensuring that output construction operates on
    individual constants rather than their union.
\end{description}

As a running example, we use the executable two-hop encoding
$\mathrm{twoHop}(X, Z) \mathrel{{:}\!{-}} \mathrm{edge}(X, Y),\; \mathrm{edge}(Y, Z)$
from
\selfcite{\nolinkurl{RelationalTwoHopRepeatedVisitor.mixin.yaml}~\cite{mixin2025}}{\nolinkurl{RelationalTwoHopRepeatedVisitor.mixin.yaml}~(supplementary material)}.
Its top-level scopes are named after the stages of the
encoding: \texttt{\_XAcceptance} iterates the first column,
\texttt{\_YAcceptance} performs the shared-variable join,
\texttt{\_NilAcceptance0} confirms the end of the first fact,
\texttt{\_ZAcceptance} iterates the output column, and
\texttt{\_NilAcceptance1} triggers the final
\texttt{Replacement} chain.
The schematic encodings below should be read as abstractions
of that executable MIXINv2 file.
If one temporarily ignores relational accumulation and reads
each dispatch as an ordinary case split, the control flow is:
choose $X$ from the first relation, choose a matching $Y$,
confirm that the first tuple is complete, choose $Z$ from the
second relation's matching row, confirm that the second tuple
is complete, and emit $(X, Z)$.
The rest of the appendix explains how that familiar
single-valued control skeleton is realised by visitor dispatch
over tries and then lifted to relational semantics by
inheritance union.

\subsection{Encoding Rules}

A Datalog rule $H(\bar{X}) \mathrel{{:}\!{-}} B_1(\bar{Y}_1), \ldots, B_n(\bar{Y}_n)$
is encoded by nesting three mechanisms inside that control
skeleton:

\begin{enumerate}
  \item \textbf{Column iteration.}
    For each column position, an $\mathrm{Acceptance}$ dispatch
    iterates over the constants present.
    The $\mathrm{VisitorMap}$ is populated from
    $\mathrm{WithVisiteeTail}$, so each matched constant's
    $\mathrm{Visitor}$ receives $\mathrm{VisiteeTail}$: the
    per-constant sub-trie at that position.
    The result of processing each constant is placed in
    $\mathrm{Visited}$, and $\mathrm{Accepted}$ collects all
    per-constant results.
    (Details in Sections~\ref{sec:encoding-variables},
    \ref{sec:encoding-constants}, and~\ref{sec:encoding-multi}.)

  \item \textbf{Join.}
    For each shared variable between body atoms, the inner
    $\mathrm{Acceptance}$'s $\mathrm{VisitorMap}$ inherits from
    \emph{both} $\mathrm{WithVisitorTail}$ (from the joined
    relation) and $\mathrm{WithVisiteeTail}$ (from the iterated
    relation).
    A constant $d$ produces a $\mathrm{Visitor}$ entry only if
    $d$ appears in both relations, achieving a value-level
    join.
    The $\mathrm{Visitor}$ then receives both
    $\mathrm{VisiteeTail}$ and $\mathrm{VisitorTail}$,
    providing access to both relations' data for the
    matched constant.

  \item \textbf{Output construction.}
    Inside the innermost $\mathrm{Nil}$ confirmation (verifying
    that each body atom's tuple is complete), chain
    $\mathrm{Replacement}$ operations from the innermost
    output column outward, each setting $\mathrm{NewTail}$ to
    the previous $\mathrm{Replaced}$ value, terminated by
    $\mathrm{Nil}$.
    The final $\mathrm{Replaced}$ value is assigned to
    $\mathrm{Visited}$.
\end{enumerate}

\subsection{Variable Join}
\label{sec:encoding-variables}

Consider the simplest join:
$\mathrm{composed}(X, Z) \mathrel{{:}\!{-}} \alpha(X, Y),\; \beta(Y, Z)$.
The executable self-join instance of this pattern is
\selfcite{\nolinkurl{RelationalTwoHopRepeatedVisitor.mixin.yaml}~\cite{mixin2025}}{\nolinkurl{RelationalTwoHopRepeatedVisitor.mixin.yaml}~(supplementary material)},
whose scope names are shown in parentheses below.
This is the right place to see the division of labour from the
section introduction: visitor dispatch provides the control flow,
the trie provides direct access to the matching tails, and
$\supers$ accumulates the outputs produced by each successful
match.

The shared variable $Y$ requires matching constants
between $\alpha$'s second column and $\beta$'s first column.
The encoding nests five $\mathrm{Acceptance}$ dispatches:
iterate $X$ from $\alpha$, join on $Y$ between $\alpha$
and $\beta$, confirm $\mathrm{Nil}$ after $Y$ in $\alpha$,
iterate $Z$ from $\beta$'s matched row, and confirm
$\mathrm{Nil}$ after $Z$.
We present each level as a named sub-scope.

\paragraph{Level~1: iterate $X$ from $\alpha$}
(Executable scope: \texttt{\_XAcceptance}.)
{\small
  \begin{align*}
    &\mathrm{xAcc} \mapsto \{
      \alpha.\mathrm{Acceptance},\;
      \mathrm{VisitorMap} \mapsto \{\alpha.\mathrm{WithVisiteeTail}\},\\
      &\quad \mathrm{Visitor} \mapsto \{
        \_X \mapsto \{\mathrm{VisiteeTail}\},\;
        \ldots,\;
    \mathrm{Visited} \mapsto \{\mathrm{yAcc}.\mathrm{Accepted}\}\}\}
\end{align*}}

\paragraph{Level~2: join $\alpha$'s $Y$-column with $\beta$'s first column}

(Executable scope: \texttt{\_YAcceptance}.)

The $\mathrm{VisitorMap}$ inherits from \emph{both}
$\beta.\mathrm{WithVisitorTail}$ and
$\_X.\mathrm{WithVisiteeTail}$, producing an entry for
constant $d$ only if $d$ appears in both $\alpha$'s
second column (via $\_X$) and $\beta$'s first column.
{\small
  \begin{align*}
    &\mathrm{yAcc} \mapsto \{
      \_X.\mathrm{Acceptance},\\
      &\quad \mathrm{VisitorMap} \mapsto \{
        \beta.\mathrm{WithVisitorTail},\;
      \_X.\mathrm{WithVisiteeTail}\},\\
      &\quad \mathrm{Visitor} \mapsto \{
        \_Y_0 \mapsto \{\mathrm{VisiteeTail}\},\;
        \mathrm{VisitorTail} \mapsto \{\beta.\mathrm{Tail}\},\\
        &\qquad
        \ldots,\;
    \mathrm{Visited} \mapsto \{\mathrm{nilAcc}_0.\mathrm{Accepted}\}\}\}
\end{align*}}

\paragraph{Level~3: confirm $\mathrm{Nil}$ after $Y$ in $\alpha$}

(Executable scope: \texttt{\_NilAcceptance0}.)

This checks that $\alpha(X, Y)$ is a complete two-column
fact (its trie path terminates at $\mathrm{Nil}$ after $Y$).
Inside the $\mathrm{Nil}$ branch, the encoding proceeds to
iterate $\beta$'s second column via $\mathrm{VisitorTail}$:
{\small
  \begin{align*}
    &\mathrm{nilAcc}_0 \mapsto \{
      \_Y_0.\mathrm{Acceptance},\;
      \mathrm{VisitorMap} \mapsto \{
        \mathrm{Nil} \mapsto \{\ldots,\;
    \mathrm{Visited} \mapsto \{\mathrm{zAcc}.\mathrm{Accepted}\}\}\}\}
\end{align*}}

\paragraph{Level~4: iterate $Z$ from $\beta$'s matched row}
(Executable scope: \texttt{\_ZAcceptance}.)
{\small
  \begin{align*}
    &\mathrm{zAcc} \mapsto \{
      \mathrm{VisitorTail}.\mathrm{Acceptance},\;
      \mathrm{VisitorMap} \mapsto \{
      \mathrm{VisitorTail}.\mathrm{WithVisiteeTail}\},\\
      &\quad \mathrm{Visitor} \mapsto \{
        \_Z \mapsto \{\mathrm{VisiteeTail}\},\;
        \ldots,\;
    \mathrm{Visited} \mapsto \{\mathrm{nilAcc}_1.\mathrm{Accepted}\}\}\}
\end{align*}}

\paragraph{Level~5: confirm $\mathrm{Nil}$ after $Z$ and construct output}
(Executable scope: \texttt{\_NilAcceptance1}, with
\texttt{\_ZReplacement} and \texttt{\_XReplacement}.)
{\small
  \begin{align*}
    &\mathrm{nilAcc}_1 \mapsto \{
      \mathrm{VisiteeTail}.\mathrm{Acceptance},\;
      \mathrm{VisitorMap} \mapsto \{
        \mathrm{Nil} \mapsto \{\\
          &\quad
          \mathrm{zR} \mapsto \{\_Z.\mathrm{Replacement},\;
          \mathrm{NewTail} \mapsto \{\mathrm{Nil}\}\},\\
          &\quad
          \mathrm{xR} \mapsto \{\_X.\mathrm{Replacement},\;
          \mathrm{NewTail} \mapsto \{\mathrm{zR}.\mathrm{Replaced}\}\},\\
          &\quad
    \mathrm{Visited} \mapsto \{\mathrm{xR}.\mathrm{Replaced}\}\}\}\}
\end{align*}}

\paragraph{Final result}
\[
  \mathrm{composed} \mapsto \{\mathrm{xAcc}.\mathrm{Accepted}\}
\]

\paragraph{Semantics}
At Level~2, the combined $\mathrm{VisitorMap}$ ensures that a
constant $d$ produces a $\mathrm{Visitor}$ only when $d$
appears in both $\alpha$'s second column \emph{and}
$\beta$'s first column.
The $\mathrm{Visitor}$ for each matched constant $d$
independently receives $\mathrm{VisiteeTail}$ ($\alpha$'s
data for~$d$) and $\mathrm{VisitorTail}$ ($\beta$'s row
for~$d$).
Output construction (Level~5) operates inside this
per-constant $\mathrm{Visitor}$, so $\mathrm{Replacement}$
acts on the \emph{individual} constant's data, not on the
merged scope.
Each matched constant $d$ produces its own output tuples
in $\mathrm{Visited}$, and these are collected by
$\mathrm{Accepted}$ via trie union ($\supers$).
This yields standard Datalog tuple-level semantics: only
constants that actually match between both relations
contribute to the output.
For the concrete instance
$\mathrm{edge} = \{(a,b),\; (b,c),\; (c,a)\}$, as encoded in
\selfcite{\nolinkurl{RelationalTwoHopRepeatedVisitor.mixin.yaml}~\cite{mixin2025}}{\nolinkurl{RelationalTwoHopRepeatedVisitor.mixin.yaml}~(supplementary material)},
Level~1 selects $a$, Level~2 matches $b$, Level~3 confirms
that $\mathrm{edge}(a,b)$ ends in $\mathrm{Nil}$, Level~4
iterates $\mathrm{edge}.\mathrm{TailMap}.b$ and selects $c$,
and Level~5 confirms $\mathrm{Nil}$ again before the two
$\mathrm{Replacement}$ operations reconstruct $(a,c)$.

\subsection{Constant Match and Navigation}
\label{sec:encoding-constants}

Constants in Datalog atoms use two mechanisms, neither of
which is a join.
They are simpler than variable joins because no value from a
second relation needs to be matched against the currently chosen
constant:

\paragraph{Type existence match}
An atom $R(X, b)$ with a constant in the second position
requires checking that constant $b$ exists in $R$'s
second column at the matched row.
This is encoded as an $\mathrm{Acceptance}$ with a
hand-crafted $\mathrm{VisitorMap}$ containing \emph{only}
the $b$ branch (instead of composing from
  $\mathrm{WithVisitorTail}$, which would create branches for
all constants):
{\small
  \begin{align*}
    \mathrm{VisitorMap} \mapsto \{b \mapsto \{\ldots\}\}
\end{align*}}
If $b$ exists in the dispatching scope, the
$b$ branch fires; otherwise, $\mathrm{Accepted}$
receives no matching branch.

\paragraph{Direct navigation}
An atom $R(b, Z)$ with a constant in the first position
navigates directly to $b$'s sub-trie:
\[
  \mathrm{edgeB}_Z \mapsto \{R.\mathrm{TailMap}.b\}
\]
This bypasses the visitor mechanism entirely, accessing
the per-constant $\mathrm{TailMap}$ entry to restrict the
scope to only $b$'s data.

\paragraph{Example}
$\mathrm{viaB}(X, Z) \mathrel{{:}\!{-}} \mathrm{edge}(X, b),\; \mathrm{edge}(b, Z)$
combines both mechanisms.
The executable MIXINv2 encoding is
\selfcite{\nolinkurl{RelationalConstant.mixin.yaml}~\cite{mixin2025}}{\nolinkurl{RelationalConstant.mixin.yaml}~(supplementary material)}.
The encoding iterates $\mathrm{edge}$ for $X$ (Level~1),
then checks for constant $b$ in $X$'s row with a
hand-crafted $\mathrm{VisitorMap}$ (Level~2), navigates
$\mathrm{edge}.\mathrm{TailMap}.b$ for $Z$ (Level~3),
confirms $\mathrm{Nil}$ (Level~4), and constructs $(X, Z)$
via $\mathrm{Replacement}$.

\subsection{Multi-Body Rules}
\label{sec:encoding-multi}

Rules with $n > 2$ body atoms chain additional join levels.
Each shared variable introduces one combined
$\mathrm{WithVisitorTail}$/$\mathrm{WithVisiteeTail}$ join
inside the previous level's $\mathrm{Visitor}$,
and the output construction is placed inside the innermost
$\mathrm{Nil}$ confirmation.
So the binary-rule encoding is the whole story structurally:
larger bodies only add more nested copies of the same pattern.

\paragraph{Three-body rule}
$\mathrm{chain}(X, W) \mathrel{{:}\!{-}} \alpha(X, Y),\; \beta(Y, Z),\; \gamma(Z, W)$
requires two joins: on $Y$ (between $\alpha$ and $\beta$)
and on $Z$ (between $\beta$ and $\gamma$).
The structure is: iterate $X$ from $\alpha$, join on $Y$
(getting both $\alpha$'s and $\beta$'s data), confirm
$\mathrm{Nil}$ after $Y$ in $\alpha$, then inside the
$\mathrm{Nil}$ branch join $\beta$'s $Z$-column with
$\gamma$'s first column, iterate $W$ from $\gamma$'s
matched row, confirm $\mathrm{Nil}$, and construct
$(X, W)$.
The $Z$-join is nested inside the $Y$-join's
$\mathrm{Nil}$ confirmation, ensuring that $\beta$'s data
for the matched $Y$ constant (available via
$\mathrm{VisitorTail}$) is used to iterate $\beta$'s
second column.

\paragraph{Mixed constants and variables}
$\mathrm{constJoin}(X, W) \mathrel{{:}\!{-}} \mathrm{edge}(X, Y),\; \mathrm{edge}(Y, b),\; \mathrm{edge}(b, W)$
combines a variable join on $Y$, a constant match on $b$
(hand-crafted $\mathrm{VisitorMap}$ with only the $b$ entry),
and direct navigation via $\mathrm{edge}.\mathrm{TailMap}.b$
for $W$.
The corresponding executable MIXINv2 file is
\selfcite{\nolinkurl{RelationalConstantJoin.mixin.yaml}~\cite{mixin2025}}{\nolinkurl{RelationalConstantJoin.mixin.yaml}~(supplementary material)}.

\subsection{Multi-Variable Join}

When a rule shares \emph{multiple} variables between two
body atoms, each shared variable requires its own
$\mathrm{Acceptance}$ check.
The checks are chained: the outer $\mathrm{Acceptance}$
joins on one column pair using combined
$\mathrm{WithVisitorTail}$/$\mathrm{WithVisiteeTail}$, and
its $\mathrm{Visitor}$ contains an inner $\mathrm{Acceptance}$
joining on another column pair.

\paragraph{Example}
$\mathrm{multiJoin}(X, Y) \mathrel{{:}\!{-}} r(X, Y),\; s(X, Y)$
shares both $X$ and $Y$.
The executable MIXINv2 encoding is
\selfcite{\nolinkurl{RelationalMultiVariableJoin.mixin.yaml}~\cite{mixin2025}}{\nolinkurl{RelationalMultiVariableJoin.mixin.yaml}~(supplementary material)}.
The outer $\mathrm{Acceptance}$ joins $r$'s and $s$'s first
columns (variable $X$) using combined
$\mathrm{WithVisitorTail}$/$\mathrm{WithVisiteeTail}$.
Inside the outer $\mathrm{Visitor}$, an inner
$\mathrm{Acceptance}$ joins $r$'s and $s$'s second columns
(variable $Y$) similarly.
Both joins must match for the output to be produced, and
the per-constant $\mathrm{Visited}$ pattern ensures that
only constants present in both relations contribute.

\subsection{Recursive Rules}
\label{sec:recursive-rules}

Recursive Datalog rules, such as transitive closure
$\mathrm{path}(X, Y) \mathrel{{:}\!{-}} \mathrm{edge}(X, Z),\; \mathrm{path}(Z, Y)$,
are encoded identically to non-recursive rules: the
head relation $\mathrm{path}$ appears both as a defined scope
and as a reference in the body (providing
$\mathrm{WithVisitorTail}$ for the $Z$-join).
This should not be read as ordinary $\this$ of a
single-valued lazy program.
Operationally, tabled evaluation
(Section~\ref{sec:mixin-trees},
Appendix~\ref{app:well-definedness})
computes $\mathrm{lfp}(T_P)$: each query into
$\mathrm{path}$ triggers further evaluation through
the recursive body, and the trie union ($\supers$)
accumulates all reachable tuples.
This corresponds exactly to \emph{Naive
evaluation}~\cite{bancilhon1986-recursive-query-strategies} of the
$T_P$
operator~\cite{vanemden1976-predicate-logic-semantics}:
each iteration applies every rule to the current
fact set and adds newly derived facts.
Because Datalog-encoded relations have finitely many
constants per query, the per-query answer domain is
finite, and convergence is guaranteed by the ascending
chain condition on a finite lattice.
\emph{Semi-Naive} evaluation~\cite{bancilhon1986-recursive-query-strategies},
which restricts each iteration to rules that use at
least one newly derived fact, is a natural
optimisation left for future work.

\paragraph{The recursive rule as monadic pseudocode}
The recursive rule above roughly corresponds to the
following pseudocode (Haskell-like syntax with
  \texttt{RebindableSyntax} and \texttt{OverloadedLists}
  semantics, where do-notation desugars to a
\texttt{Set}-based bind and list literals denote sets):
\begin{lstlisting}[style=haskell]
{-# LANGUAGE RebindableSyntax, OverloadedLists #-}
s >>= f = unions [f x | x <- s]
return x = [x]
guard True  = [()]
guard False = []
edge = [("a","b"), ("b","c"), ("c","a")]
path = edge <> do
  (x, y0) <- edge
  (y1, z) <- path
  guard (y0 == y1)
  return (x, z)
\end{lstlisting}
\noindent
Each line of the pseudocode has a direct counterpart in
the inheritance-calculus encoding
(Section~\ref{sec:encoding-variables}):
\begin{description}
  \item[\texttt{edge}] corresponds to the base-fact trie;
  \item[\texttt{path = edge <> do \ldots}]
    corresponds to $\mathrm{path}$ inheriting $\mathrm{edge}$
    (base facts) and the recursive body (derived facts via
    $\supers$);
  \item[\texttt{(x, y0) <- edge}]
    corresponds to Level~1 ($\mathrm{xAcc}$), iterating $X$
    from $\mathrm{edge}$;
  \item[\texttt{(y1, z) <- path}]
    corresponds to Level~2 ($\mathrm{yAcc}$), joining on $Y$
    with $\mathrm{path}$, and Level~4 ($\mathrm{zAcc}$),
    iterating $Z$;
  \item[\texttt{guard (y0 == y1)}]
    corresponds to the $\mathrm{WithVisitorTail} \cap
    \mathrm{WithVisiteeTail}$ intersection in Level~2: a
    constant produces a $\mathrm{Visitor}$ entry only if it
    appears in both relations;
  \item[\texttt{return (x, z)}]
    corresponds to the $\mathrm{Replacement}$ chain in
    Level~5, constructing the output tuple $(X, Z)$.
\end{description}
The only difference is the evaluation strategy:
Haskell evaluates \texttt{path} lazily (diverges),
whereas inheritance-calculus evaluates it via tabled
$T_P$ iteration (converges to $\mathrm{lfp}$).

However, this monadic program does not converge because
\texttt{path} is defined recursively.%
\footnote{
  Enabling the \texttt{RecursiveDo} extension does not help:
  \texttt{mfix} ties a knot at the \emph{element} level via laziness,
  whereas the transitive closure equation requires a
  fixed point at the \emph{container} level, that is, finding a
  set~$S$ such that $S = F(S)$.
}
Inheritance-calculus solves this through tabled evaluation
(Section~\ref{sec:mixin-trees}), which computes the
least fixed point of the $T_P$ operator by iterating
over the lattice of mixin trees, precisely the
container-level fixed point that the pseudocode above cannot express.

\subsection{Semantic Correspondence with Standard Datalog}
\label{sec:datalog-semantics}

The encoding produces standard Datalog tuple-level semantics.
Putting the previous subsections together: tries provide indexed
tuple storage, visitor dispatch supplies the control skeleton,
and per-constant $\mathrm{Replacement}$ keeps output
construction local to each witness.
The key mechanism is the per-constant $\mathrm{Visited}$
pattern: output construction ($\mathrm{Replacement}$)
occurs inside each matched constant's $\mathrm{Visitor}$
scope, operating on the \emph{individual} constant's data
rather than on the merged scope.
The $\mathrm{Accepted}$ field then collects all per-constant
$\mathrm{Visited}$ results via trie union ($\supers$).

For example, given $\mathrm{edge} = \{(a,b),\; (b,c),\; (c,a)\}$
(a~3-cycle), the two-hop self-join
$\mathrm{twoHop}(X, Z) \mathrel{{:}\!{-}} \mathrm{edge}(X, Y),\; \mathrm{edge}(Y, Z)$
produces exactly $3$ pairs:
$\{(a,c),\; (b,a),\; (c,b)\}$, matching standard Datalog.
This is the behavior exhibited by
\selfcite{\nolinkurl{RelationalTwoHopRepeatedVisitor.mixin.yaml}~\cite{mixin2025}}{\nolinkurl{RelationalTwoHopRepeatedVisitor.mixin.yaml}~(supplementary material)}.
When the $Y$-join matches constant $b$ (present in both
  $\mathrm{edge}$'s second column from $(a,b)$ and
$\mathrm{edge}$'s first column), the $\mathrm{Visitor}$ for
$b$ receives $\mathrm{VisitorTail} = \mathrm{edge}.\mathrm{TailMap}.b$
(containing only~$c$), and $\mathrm{Replacement}$ constructs
a single output tuple $(a, c)$, rather than all of $\{a,b,c\}$.

This value-level precision arises because each constant's
$\mathrm{Replacement}$ operates on $\mathrm{TailMap}.d$
(the per-constant sub-trie) rather than on the merged
trie.
The same point is visible in
the repeated-visitor self-join regression example\selfcite{~\cite{mixin2025}}{~(supplementary material)},
where repeated visitor dispatch prevents witnesses from
different constants from being merged spuriously.
Although $\bases$ (equation~\ref{eq:bases}) still computes
cross-joins at the primitive level, the encoding routes
computation through per-constant $\mathrm{TailMap}$ entries
before any merging occurs, achieving the same effect as
standard Datalog's tuple-level evaluation.

\begin{acks}
  Claude Code, GitHub Copilot, Gemini Code Assist, and Codex were utilized to generate sections of this Work,
  including text, tables, graphs, code, formulas, citations, etc.
\end{acks}

\end{document}
